---
title: "Analysis Notebook"
toc: true
toc-depth: 3
number-sections: true
fig-width: 8
fig-height: 6

execute:
  freeze: false
  cache: false

---

# Setup
```{r}
#| label: setup
#| include: false

packages <- c("DT","MuMIn","lunar","glmmTMB","DHARMa","broom.mixed","emmeans","lmerTest","lme4","patchwork","tidyverse","writexl","readxl")

quiet_load <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    suppressWarnings(suppressMessages(install.packages(pkg, dependencies = TRUE)))
  }
  suppressPackageStartupMessages(require(pkg, character.only = TRUE, quietly = TRUE))
  invisible(TRUE)
}

options(repos = c(CRAN = "https://cloud.r-project.org"))
invisible(lapply(packages, quiet_load))

exc_file_dir  <- "data/raw/Data_Stone_piles_Rotoiti.xlsx"
raw_data_dir  <- "data/raw"
der_data_dir  <- "data/derived"
out_dir       <- "outputs"

# Read data
Site_info          <- readxl::read_excel(exc_file_dir, sheet = "Site_info")
Monitoring_data    <- readxl::read_excel(exc_file_dir, sheet = "Monitoring_data")
Chemical_data      <- readxl::read_excel(exc_file_dir, sheet = "Chemical_data")
Stone_pile_data    <- readxl::read_excel(exc_file_dir, sheet = "Stone_pile_data")  %>% select(-starts_with("..."))
Weed_data          <- readxl::read_excel(exc_file_dir, sheet = "Weed_data")  %>% select(-starts_with("..."))
Fish_data          <- readxl::read_excel(exc_file_dir, sheet = "Fish_data")  %>% select(-starts_with("..."))
Macroinvertebrates <- readxl::read_excel(exc_file_dir, sheet = "Macroinvertebrates")

# Save as csv to derived
write.csv(Site_info,          file.path(der_data_dir, "Site_info.csv"),          row.names = FALSE)
write.csv(Monitoring_data,    file.path(der_data_dir, "Monitoring_data.csv"),    row.names = FALSE)
write.csv(Chemical_data,      file.path(der_data_dir, "Chemical_data.csv"),      row.names = FALSE)
write.csv(Stone_pile_data,    file.path(der_data_dir, "Stone_pile_data.csv"),    row.names = FALSE)
write.csv(Weed_data,          file.path(der_data_dir, "Weed_data.csv"),          row.names = FALSE)
write.csv(Fish_data,          file.path(der_data_dir, "Fish_data.csv"),          row.names = FALSE)
write.csv(Macroinvertebrates, file.path(der_data_dir, "Macroinvertebrates.csv"), row.names = FALSE)

### Reproducibility note 
# The sections above require data/raw/Natural_habitat.xlsx (not tracked).
# To reproduce analyses and figures without the raw Excel file,
# run from here using the derived CSVs in data/derived/:

Site_info          <- read.csv(file.path(der_data_dir, "Site_info.csv"))
Monitoring_data    <- read.csv(file.path(der_data_dir, "Monitoring_data.csv"))
Chemical_data      <- read.csv(file.path(der_data_dir, "Chemical_data.csv"))
Stone_pile_data    <- read.csv(file.path(der_data_dir, "Stone_pile_data.csv"))
Weed_data          <- read.csv(file.path(der_data_dir, "Weed_data.csv"))
Fish_data          <- read.csv(file.path(der_data_dir, "Fish_data.csv"))
Macroinvertebrates <- read.csv(file.path(der_data_dir, "Macroinvertebrates.csv"))

# set plotting aesthetics
site_colours <- c("Stone_pile" = "#ff0000", "Control" = "#eeff00", "Reference" = "#0000ff")
site_fill <- c("Stone_pile" = "#ff0000", "Control" = "#eeff00", "Reference" = "#0000ff")
site_shapes  <- c("Stone_pile" = 21,  "Control" = 24, "Reference" = 22)

conv_colours <- c("Stone_pile"= "#ff0000","Reference" = "#0000ff")
conv_fill    <- conv_colours
conv_shapes  <- c("Stone_pile" = 21,"Reference" = 22)

site_labels <- c("Stone_pile" = "Stone pile", "Control" = "Control", "Reference" = "Reference")
conv_labels <- c("Stone_pile" = "Stone pile", "Reference" = "Reference")

monitor_labels <- c("Pre/Summer\n2025", "Autumn\n2025", "Winter\n2025", "Spring\n2025", "Summer\n2026", "Autumn\n2026", "Winter\n(upcoming)")

base_theme_bw <- theme_classic() +
  theme(text     = element_text(family = "Arial", size = 11),
    axis.title   = element_text(face = "plain"),
    axis.text    = element_text(face = "plain"),
    plot.title   = element_text(face = "plain"),
    strip.text   = element_text(face = "plain"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.3))
theme_set(base_theme_bw)

# Format of p values in text, including the correct comparison operator and
# the italic "p" label. Set label = FALSE for values chained after a leading
# "*p*" (e.g. "*p* < 0.001, 0.011, 0.004, respectively").
format_pval <- function(p, label = TRUE) {
  out <- if (p < 0.001) "< 0.001" else paste0("= ", format(round(p, 3), nsmall = 3))
  if (label) paste0("*p* ", out) else out
}

```

# Data ingestion and preparation
## Length–weight
```{r}
#| label: fig-length-weight
#| include: true
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Length–weight relationship for kōura, fitted on log₁₀–log₁₀ scale. Open points are weights predicted from the length–weight regression (used to fill missing scale measurements); filled points are directly measured weights."

# use size of koura to fill missing weights.
lw_model <- lm(log10(weight_g) ~ log10(length_mm), data = Fish_data, subset = species_name == "Freshwater_crayfish", na.action = na.exclude)

# extract the residual variance
sigma_log10 <- sigma(lw_model)
c <- 10^(0.5 * sigma_log10^2)

Fish_data <- Fish_data |>
  mutate(is_koura = species_name == "Freshwater_crayfish",
    Predicted_weight = case_when( !is.na(weight_g) ~ weight_g,
      is_koura ~ {pred_log <- predict(lw_model, newdata = pick(everything()))
        10^(pred_log) * c}, TRUE ~ NA_real_ ),
    Weight_source = case_when(is_koura & !is.na(weight_g) ~ "measured",
      is_koura & is.na(weight_g) ~ "predicted",TRUE ~ NA_character_))

count_df <- Fish_data |>
  dplyr::filter(is_koura, !is.na(Weight_source)) |>
  dplyr::count(Weight_source, name = "n")

measured_n  <- count_df$n[count_df$Weight_source == "measured"]
predicted_n <- count_df$n[count_df$Weight_source == "predicted"]

total_koura <- Fish_data |>
  dplyr::filter(is_koura) |>
  nrow()

hist(Fish_data |> dplyr::filter(is_koura) |> dplyr::pull(length_mm), xlab = "Length (mm)")

# Extract model coefficients
a <- coef(lw_model)[1]
b <- coef(lw_model)[2]

# Create formula text for the plot
formula_text <- paste0( "log10(Weight[g]) = ", round(a, 3), " + ", round(b, 3), " * log10(OCL Length[mm])")

# Plot with formula added
length_weight_plot <- ggplot(
Fish_data |> dplyr::filter(is_koura),
aes(x = length_mm, y = Predicted_weight, shape = Weight_source)) +
geom_point(size = 2.2, colour = "black", stroke = 0.6) +
scale_shape_manual(values = c("measured" = 16, "predicted" = 1)) +
labs(x = "OCL length (mm)", y = "Weight (g)", shape = "Weight source") +
annotate("text", x = Inf, y = Inf, label = formula_text, hjust = 1.1, vjust = 1.3, size = 3) 

ggsave(file.path(out_dir, "fig-length-weight.png"),length_weight_plot,width = 8, height = 5, dpi = 300)

length_weight_plot

```

## Build CPUE and BPUE
```{r}
#| label: cpue-bpue
#| include: false

CPUE_BPUE_legacy <- Fish_data %>%
  filter(!is.na(species)) %>%
  group_by(monitoring_id, species, net_type) %>%
  dplyr::reframe(
    Total_Individuals = sum(amount, na.rm = TRUE),
    Total_Weight      = sum(weight_g, na.rm = TRUE),
    Total_Effort      = dplyr::first(amount_nets),
    CPUE              = Total_Individuals / Total_Effort,
    BPUE              = Total_Weight      / Total_Effort,
    Mean_Length       = mean(length_mm, na.rm = TRUE),
    Min_Length        = ifelse(all(is.na(length_mm)), NA, min(length_mm, na.rm = TRUE)),
    Max_Length        = ifelse(all(is.na(length_mm)), NA, max(length_mm, na.rm = TRUE)),
    Mean_Weight       = mean(weight_g, na.rm = TRUE),
    Min_Weight        = ifelse(all(is.na(weight_g)), NA, min(weight_g, na.rm = TRUE)),
    Max_Weight        = ifelse(all(is.na(weight_g)), NA, max(weight_g, na.rm = TRUE))
  )

CPUE_BPUE_weighted <- CPUE_BPUE_legacy %>%
  group_by(monitoring_id, species) %>%
  summarise(
    Total_Individuals       = sum(Total_Individuals, na.rm = TRUE),
    Total_Weight            = sum(Total_Weight,      na.rm = TRUE),
    Weighted_CPUE_numerator = sum(CPUE * Total_Effort,  na.rm = TRUE),
    Weighted_BPUE_numerator = sum(BPUE * Total_Effort,  na.rm = TRUE),
    Total_Effort_sum        = sum(Total_Effort,         na.rm = TRUE),
    Mean_Length             = mean(Mean_Length, na.rm = TRUE),
    Min_Length              = ifelse(all(is.na(Min_Length)), NA, min(Min_Length, na.rm = TRUE)),
    Max_Length              = ifelse(all(is.na(Max_Length)), NA, max(Max_Length, na.rm = TRUE)),
    Mean_Weight             = mean(Mean_Weight, na.rm = TRUE),
    Min_Weight              = ifelse(all(is.na(Min_Weight)), NA, min(Min_Weight, na.rm = TRUE)),
    Max_Weight              = ifelse(all(is.na(Max_Weight)), NA, max(Max_Weight, na.rm = TRUE)),
    .groups = "drop") %>%
  ungroup() %>%
  mutate(Total_Effort_sum = ifelse(monitoring_id %in% c("16_0", "10_1", "3_2", "16_2"), 3, 4),
    Weighted_CPUE    = Weighted_CPUE_numerator / Total_Effort_sum,
    Weighted_BPUE    = Weighted_BPUE_numerator / Total_Effort_sum)

species_presence_absence <- Fish_data %>%
  filter(!is.na(species)) %>%
  distinct(monitoring_id, species) %>%
  mutate(presence = 1) %>%
  pivot_wider(
    names_from  = species,
    values_from = presence,
    values_fill = list(presence = 0),
    names_prefix = "Presence_") %>%
  mutate(Predator_Fish_Presence = pmax(Presence_Trout, Presence_Eel, Presence_Catfish))

CPUE_BPUE_weighted_summary <- CPUE_BPUE_weighted %>%
  pivot_wider(
    names_from  = species,
    values_from = c(
      Total_Individuals, Weighted_CPUE, Weighted_BPUE, Total_Weight,
      Mean_Length, Mean_Weight, Weighted_CPUE_numerator, Weighted_BPUE_numerator,
      Total_Effort_sum, Min_Length, Max_Length, Min_Weight, Max_Weight),
    names_sep   = "_", values_fill = list(Total_Individuals = 0, Weighted_CPUE = 0, Weighted_BPUE = 0)) %>%
  mutate(
    Richness  = rowSums(dplyr::select(., starts_with("Total_Individuals_")) > 0),
    Abundance = rowSums(dplyr::select(., starts_with("Total_Individuals_") & !ends_with(c("_Bullies", "_Common_smelt")))))
```


## Combined dataset construction
```{r}
#| label: combine-data
#| include: true

unit_metadata <- dplyr::bind_rows(Monitoring_data %>%
    dplyr::select(parameter, unit), Chemical_data %>%
    dplyr::select(parameter, unit)) %>%
  dplyr::distinct() %>%
  dplyr::rename(Variable = parameter, Unit = unit)

Monitoring_summary <- Monitoring_data %>%
  dplyr::select(-group, -notes, -unit, -monitoring_id) %>%
  tidyr::pivot_wider(names_from  = parameter,
    values_from = value,
    values_fill = list(value = NA)) %>%
  dplyr::mutate(
    dplyr::across(c(Riparian_vegetation, Vegetation_nearby, Overhanging_trees, Erosion, Structure, Bedrock, Boulders, Cobble, Gravel, Sand, Mud, Organic_matter, Wood_cover ), ~ as.numeric(.)))

Chemical_summary <- Chemical_data %>%
  dplyr::select(-group, -notes, -unit, -site_id) %>%
  tidyr::pivot_wider(names_from  = parameter,
    values_from = value,
    values_fill = list(value = NA)) %>%
  dplyr::mutate(
    dplyr::across(c(Temperature, DO_mgl, DO_percent, Conductivity, Specific_conductivity, pH), ~ as.numeric(.)))

Weed_summary <- Weed_data %>%
  group_by(monitoring_id, weed_type, native_status) %>%
  summarise(Total_Cover = sum(percentage_cover, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from  = c(weed_type, native_status), values_from = Total_Cover, values_fill = 0) %>% 
  dplyr::select(monitoring_id, Emergent_Native, Emergent_Non_Native, Submerged_Non_Native, Turf_Native)

Stone_pile_summary <- Stone_pile_data %>%
  select(-unit, -notes, -site_id)%>%
  pivot_wider(names_from = c(parameter, stone_pile_id), values_from = value, names_sep = ".")

Macroinvertebrates_sum <- Macroinvertebrates %>%
  group_by(monitoring_id, species) %>%
  summarise(Total_amount = sum(amount, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = c(species), values_from = Total_amount, values_fill = 0) %>%
  mutate(Invertebrates_Richness  = rowSums(dplyr::select(., -monitoring_id) > 0),
    Invertebrates_Abundance = rowSums(dplyr::select(., -monitoring_id)))

Combined_data <- Site_info %>%
  left_join(Monitoring_summary, by = "site_id") %>%
  left_join(Chemical_summary, by = c("monitoring_id")) %>%
  left_join(Weed_summary, by = c("monitoring_id")) %>%
  left_join(Stone_pile_summary, by = c("monitoring_id")) %>%
  left_join(CPUE_BPUE_weighted_summary, by = "monitoring_id") %>%
  left_join(species_presence_absence, by = "monitoring_id") %>%
  left_join(Macroinvertebrates_sum, by = "monitoring_id")

```

## Derived variables
```{r}
#| label: derived-variables
#| include: true

Combined_data <- Combined_data %>%
  dplyr::mutate(date_time = as.POSIXct(date_time, tz = "Pacific/Auckland"),
    Presence_rocks = dplyr::if_else(Cobble > 1 | Boulders > 1, 1, 0),
    Period = factor(dplyr::if_else(monitoring == 0, "Before", "After"),levels = c("Before", "After")),
    Date  = as.Date(date_time),
    Time  = format(date_time, "%H:%M:%S"),
    Year  = lubridate::year(date_time),
    Month = lubridate::month(date_time, label = TRUE),
    Day   = lubridate::day(date_time),
    Season = dplyr::case_when(
      Month %in% c("Dec", "Jan", "Feb") ~ "Summer",
      Month %in% c("Mar", "Apr", "May") ~ "Autumn",
      Month %in% c("Jun", "Jul", "Aug") ~ "Winter",
      Month %in% c("Sep", "Oct", "Nov") ~ "Spring",
      TRUE ~ NA_character_),
    Date_Time_Numeric = as.numeric(date_time),
    Monitoring_label = dplyr::case_when(
      monitoring == 0 ~ "0 Pre/Summer",
      monitoring == 1 ~ "1 Autumn",
      monitoring == 2 ~ "2 Winter",
      monitoring == 3 ~ "3 Spring",
      monitoring == 4 ~ "4 Summer",
      monitoring == 5 ~ "5 Autumn",
      monitoring == 6 ~ "6 Winter",
      TRUE ~ as.character(monitoring)
    ) %>% as.factor(), site_id = as.factor(site_id)) %>%
  dplyr::mutate(
  Depth_5m    = suppressWarnings(as.numeric(Distance_5m)),
  Depth_20m   = suppressWarnings(as.numeric(Distance_20m)),
  Slope_0_5m  = 5  / Depth_5m,    # 5m rise / horizontal distance in m
  Slope_0_20m = 20 / Depth_20m)   # 20m rise / horizontal distance in m
        

habitat_classification <- Combined_data %>%
  filter(monitoring == 0) %>%
  dplyr::select(site_id, DHT, lake,
                Bedrock, Boulders, Cobble, Gravel, Sand, Mud, Organic_matter,
                Emergent_Native, Emergent_Non_Native, Submerged_Non_Native, Wood_cover) %>%
  pivot_longer(cols = c(Bedrock, Boulders, Cobble, Gravel, Sand, Mud, Organic_matter,
                  Emergent_Native, Emergent_Non_Native, Submerged_Non_Native, Wood_cover),
    names_to  = "Type", values_to = "Percentage") %>%
  group_by(site_id) %>%
  summarise(Rocky_Percentage  = sum(Percentage[Type %in% c("Bedrock", "Boulders", "Cobble")], na.rm = TRUE),
    Sand_Percentage     = sum(Percentage[Type == "Sand"], na.rm = TRUE),
    Mud_Percentage      = sum(Percentage[Type %in% c("Mud", "Organic_matter")], na.rm = TRUE),
    Emergent_Percentage = sum(Percentage[Type %in% c("Emergent_Native")], na.rm = TRUE),
    Substrate_index_before = sum(
      0.08 * Percentage[Type == "Bedrock"] +
      0.07 * Percentage[Type == "Boulders"] +
      0.06 * Percentage[Type == "Cobble"] +
      0.04 * Percentage[Type == "Gravel"] +
      0.03 * Percentage[Type == "Sand"] +
      0.02 * Percentage[Type == "Organic_matter"] +
      0.01 * Percentage[Type == "Mud"],
      na.rm = TRUE),.groups = "drop") %>%
  mutate(Habitat_Type = case_when(
      Rocky_Percentage    > 25 ~ "Rocky",
      Emergent_Percentage > 25 ~ "Emergent Macrophyte",
      Sand_Percentage   >= Mud_Percentage ~ "Sandy",
      TRUE ~ "Muddy")) %>%
  dplyr::select(site_id, Habitat_Type, Substrate_index_before)

Combined_data <- Combined_data %>%
  left_join(habitat_classification, by = c("site_id"))%>%
  dplyr::mutate(monitoring_int = as.integer(monitoring),
    pair_id = factor(ceiling(as.numeric(as.character(site_id)) / 2)),
    koura_present = as.integer(`Total_Individuals_Kōura` > 0),
    Site_type_detail = factor(dplyr::case_when(
      site == "Stone_pile" ~ "Stone_pile",
      site == "Control"    ~ "Control",
      site == "Reference" & as.numeric(as.character(site_id)) %% 2 == 1 ~ "Reference_1",
      site == "Reference" & as.numeric(as.character(site_id)) %% 2 == 0 ~ "Reference_2",
      TRUE ~ site),
      levels = c("Stone_pile", "Control", "Reference_1", "Reference_2")))

# Substrate_index_before above is fixed at the baseline (pre-construction) substrate
# survey and never updated, even though the stone piles physically added new
# rocky structure at Stone_pile sites. Rather than assume a flat added-cover
# percentage for every site, use each site's own measured pile footprint:
# total footprint area of the ~3 piles at that site, as a percentage of the
# 10 m x 15 m (150 m^2) site survey area, classified 80% Cobble / 20%
# Boulders (most stones were below boulder size) and displacing an equal
# area of Sand (the pre-existing substrate at all 5 stone-pile sites).
# Substrate_index_after equals the original Substrate_index_before
# everywhere except Stone_pile sites at monitoring_int >= 1 (post-
# construction), where it reflects this site-specific addition.
site_area_m2 <- 10 * 15

pile_footprint_pct <- Stone_pile_data %>%
  dplyr::filter(stringr::str_detect(monitoring_id, "_0"), parameter == "Area") %>%
  dplyr::mutate(value_num = suppressWarnings(as.numeric(value)),
                site_id   = as.character(site_id)) %>%
  dplyr::group_by(site_id) %>%
  dplyr::summarise(footprint_pct = sum(value_num, na.rm = TRUE) / site_area_m2 * 100, .groups = "drop")

substrate_updated <- Combined_data %>%
  dplyr::filter(monitoring_int == 0) %>%
  dplyr::distinct(site_id, Bedrock, Boulders, Cobble, Gravel, Sand, Mud, Organic_matter) %>%
  dplyr::mutate(site_id = as.character(site_id)) %>%
  dplyr::left_join(pile_footprint_pct, by = "site_id") %>%
  tidyr::replace_na(list(footprint_pct = 0)) %>%
  dplyr::mutate(
    Boulders_new = Boulders + footprint_pct * 0.20,
    Cobble_new   = Cobble   + footprint_pct * 0.80,
    Sand_new     = Sand     - footprint_pct,
    Substrate_index_after = 0.08 * Bedrock + 0.07 * Boulders_new + 0.06 * Cobble_new +
      0.04 * Gravel + 0.03 * Sand_new + 0.02 * Organic_matter + 0.01 * Mud) %>%
  dplyr::select(site_id, Substrate_index_after)

Combined_data <- Combined_data %>%
  dplyr::mutate(site_id = as.character(site_id)) %>%
  dplyr::left_join(substrate_updated, by = "site_id") %>%
  dplyr::mutate(
    # as.factor() alone would sort "1","10","11",...,"2","20",...
    # alphabetically — sort numerically first so site_id reads 1-20.
    site_id = factor(site_id, levels = as.character(sort(as.numeric(unique(site_id))))),
    Substrate_index_after = dplyr::if_else(
      site == "Stone_pile" & monitoring_int >= 1,
      Substrate_index_after,
      Substrate_index_before))

Combined_data <- Combined_data %>%
  mutate(lunar_phase = lunar.phase(Date, name = FALSE),  # 0-2π radians
    lunar_illumination = (1 - cos(lunar.phase(Date))) / 2,  # 0=new, 1=full
    lunar_name = lunar.phase(Date, name = TRUE))     # "New", "Waxing", "Full", "Waning"

# Monitoring quality flag — based on known compromised events
compromised_ids <- c("6_1", "13_1", "14_1", "1_2", "2_2", "19_2", "20_2", "1_3", "2_3", "6_5")

Combined_data <- Combined_data %>%
  mutate(monitoring_quality = case_when(monitoring_id %in% compromised_ids ~ "compromised", is.na(`Total_Individuals_Kōura`) ~ "missing", TRUE  ~ "complete"))

writexl::write_xlsx(Combined_data, file.path(der_data_dir, "Combined_data.xlsx"))
write.csv(Combined_data, file.path(der_data_dir, "Combined_data.csv"), row.names = FALSE)
write.csv(habitat_classification, file.path(der_data_dir, "habitat_classification.csv"), row.names = FALSE)

all_data <- Combined_data

```

## Missing data summary
```{r}
#| label: tbl-missing-data-summary
#| include: true
#| tbl-cap: "Monitoring events excluded due to environmental disturbance."

# Missing data summary
missing_table <- all_data %>%
  filter(monitoring_quality != "complete") %>%
  select(site_id, site, Monitoring_label, monitoring_id, monitoring_quality) %>%
  arrange(site_id, Monitoring_label)

knitr::kable(missing_table,
             col.names = c("Site ID", "Site", "Monitoring label", "Monitoring ID", "Quality"))

```


# Site overview
## Site overview table
```{r}
#| label: tbl-site-overview
#| include: true
#| tbl-cap: "Site overview: baseline physical characteristics by site. Stone pile structural dimensions are reported in Methods and the Stone pile structure figure instead, since they have no Control/Reference equivalent to display alongside."

# Site-level info from all_data — baseline monitoring only
site_info_summary <- all_data %>%
  filter(monitoring == 0) %>%
  distinct(site_id, site, lat, lon, DHT, 
           Habitat_Type, Substrate_index_before,
           Slope_0_20m) %>%
  mutate(site_id = as.factor(site_id))

# Stone pile summary — mean across 3 piles per modified site at baseline
stone_pile_site_summary <- Stone_pile_data %>%
  filter(str_detect(monitoring_id, "_0")) %>%
  filter(parameter %in% c("Weight", "Volume", "Height_above", 
                           "Distance_to_shore", "Depth_on_R", 
                           "Depth_on_L", "Depth_after", "Area")) %>%
  mutate(value_num = suppressWarnings(as.numeric(value))) %>%
  group_by(site_id, parameter) %>%
  summarise(
    value_num = mean(value_num, na.rm = TRUE),
    n_piles   = n_distinct(stone_pile_id),
    .groups   = "drop") %>%
  pivot_wider(names_from = parameter, values_from = value_num) %>%
  mutate(
    # Cast to character (not factor) so the join below inherits
    # site_info_summary's (already correctly numeric-ordered) factor
    # instead of re-sorting this subset alphabetically on its own —
    # left_join() coerces factor-vs-character but errors on factor-vs-integer.
    site_id          = as.character(site_id),
    Depth_m          = round((Depth_on_R + Depth_on_L) / 2 / 1000, 2),
    Depth_after_m    = round(Depth_after       / 1000, 2),
    Height_above_m   = round(Height_above       / 1000, 2),
    Distance_shore_m = round(Distance_to_shore / 1000, 2),
    Area_m2          = round(Area,   2),
    Weight_kg        = round(Weight, 0),
    Volume_m3        = round(Volume, 2)) %>%
  select(site_id, Distance_shore_m, Depth_m, Height_above_m, Area_m2, Volume_m3)

# Join and finalise
site_data <- site_info_summary %>%
  left_join(stone_pile_site_summary, by = "site_id") %>%
  arrange(as.numeric(as.character(site_id))) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2))) %>%
  select(site_id, site, DHT, Substrate_index_before, Slope_0_20m, Distance_shore_m, Depth_m, Height_above_m, Area_m2, Volume_m3)

# Display table drops the stone-pile build dimensions (Distance_shore_m,
# Depth_m, Height_above_m, Area_m2, Volume_m3) — they only exist for
# Stone_pile sites, so they don't fit a per-site comparison table. site_data
# itself keeps them, since stone_pile_data2/mean_dist_shore/mean_depth/
# mean_area below still depend on them.
site_overview_display <- site_data %>%
  dplyr::select(site_id, site, DHT, Substrate_index_before, Slope_0_20m)

write.csv(site_overview_display, file = file.path(out_dir, "tbl-site-overview.csv"), row.names = FALSE)

knitr::kable(site_overview_display, digits = 2)

```

# Descriptive results
## Descriptive context plots
Background/methods-supporting plots — combined into a single figure below
rather than shown individually, since none are primary results.

### Habitat characteristics
```{r}
#| label: fig-habitat-plot
#| include: false
#| fig-cap: "Baseline habitat characteristics (slope, substrate index, riparian vegetation, overhanging tree cover, wood cover), mean ± SE by site type."

Habitat_data <- all_data %>%
  dplyr::select(site_id, site, monitoring_int, Habitat_Type,Slope_0_5m, Slope_0_20m, Substrate_index_before, Substrate_index_after, Riparian_vegetation, Overhanging_trees, Wood_cover) %>%
  tidyr::pivot_longer(cols = c(Slope_0_5m, Slope_0_20m, Substrate_index_before, Substrate_index_after, Riparian_vegetation, Overhanging_trees, Wood_cover), names_to  = "Variable", values_to = "Value") %>%
  dplyr::group_by(site_id, site, Variable) %>%
  # Most of these are static baseline-only measurements (constant across all
  # rounds), so first/last makes no difference. Substrate_index_after is
  # the exception — it only changes from round 1 onward for Stone_pile sites
  # — so take the latest round to show the current (post-construction) state
  # rather than the frozen baseline.
  dplyr::slice_max(monitoring_int, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(site = factor(site, levels = c("Stone_pile", "Control", "Reference")))

# Mean ± SE by site type rather than one bar per site_id — individual-site
# detail is already in the site overview table; at combined-figure scale,
# 20 site labels per facet aren't legible and aren't the point here.
Habitat_summary <- Habitat_data %>%
  dplyr::group_by(site, Variable) %>%
  dplyr::summarise(
    mean_v = mean(Value, na.rm = TRUE),
    se     = sd(Value, na.rm = TRUE) / sqrt(sum(!is.na(Value))),
    .groups = "drop")

habitat_var_labels <- c(
  Slope_0_5m             = "Slope (0-5m)",
  Slope_0_20m            = "Slope (0-20m)",
  Substrate_index_before = "Substrate index\n(before)",
  Substrate_index_after  = "Substrate index\n(after)",
  Riparian_vegetation    = "Riparian\nvegetation",
  Overhanging_trees      = "Overhanging\ntrees",
  Wood_cover             = "Wood cover")

Habitat_plot <- ggplot(Habitat_summary, aes(site, mean_v, fill = site)) +
  geom_col(colour = "black") +
  geom_errorbar(aes(ymin = mean_v - se, ymax = mean_v + se), width = 0.25) +
  facet_wrap(~ Variable, scales = "free_y", nrow = 2, labeller = as_labeller(habitat_var_labels)) +
  scale_fill_manual(values = site_fill, labels = site_labels) +
  labs(y = "Mean value (± SE)", x = NULL) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), legend.position = "none")

ggsave(file.path(out_dir, "fig-habitat-plot.png"), Habitat_plot, width = 10, height = 6, dpi = 300)
```

### Stone pile structure
```{r}
#| label: fig-stone-pile-structure
#| include: false
#| fig-cap: "Stone pile structure parameters at each site."

stone_pile_data2 <- site_data %>%
  dplyr::select(site_id, site, Distance_shore_m, Depth_m, Height_above_m, Area_m2, Volume_m3) %>%
  tidyr::pivot_longer(
    cols      = c(Distance_shore_m, Depth_m, Height_above_m, Area_m2, Volume_m3),
    names_to  = "Parameter",
    values_to = "Value") %>%
  dplyr::mutate(site = factor(site, levels = c("Stone_pile", "Control", "Reference")))

parameter_labels <- c(
  Distance_shore_m = "Distance to shore (m)",
  Depth_m          = "Depth (m)",
  Height_above_m   = "Height above lakebed (m)",
  Area_m2          = "Area (m²)",
  Volume_m3        = "Volume (m³)")

# Only Stone_pile sites have real values here (Control/Reference are NA
# from the left_join in site_data) — filter the NAs out before plotting so
# the unused factor levels actually drop, instead of leaving 15 empty tick
# positions squeezed in next to the 5 real bars. No fill/legend needed
# either, since every remaining bar is the same site type.
stone_pile_structure_plot <- stone_pile_data2 %>%
  dplyr::filter(!is.na(Value)) %>%
  ggplot(aes(site_id, Value)) +
  geom_col(colour = "black", fill = site_fill[["Stone_pile"]]) +
  facet_wrap(~ Parameter, scales = "free_y", nrow = 2, labeller = as_labeller(parameter_labels)) +
  labs(y = "Value", x = "Stone pile site")

ggsave(file.path(out_dir, "fig-stone-pile-structure.png"), stone_pile_structure_plot, width = 8, height = 6, dpi = 300)



# Overall means across sites, used inline in the Methods text
mean_dist_shore <- mean(stone_pile_data2$Value[stone_pile_data2$Parameter == "Distance_shore_m"], na.rm = TRUE)
mean_depth      <- mean(stone_pile_data2$Value[stone_pile_data2$Parameter == "Depth_m"], na.rm = TRUE)
mean_area       <- mean(stone_pile_data2$Value[stone_pile_data2$Parameter == "Area_m2"], na.rm = TRUE)
```

### Chemical conditions over time
```{r}
#| label: fig-chemical-plot
#| include: false
#| fig-cap: "Chemical conditions (temperature, dissolved oxygen, conductivity, pH) over the monitoring period, coloured by site type."

Chemical_data2 <- all_data %>%
  dplyr::select(monitoring_id, monitoring_int, site_id, site, Monitoring_label, Habitat_Type, DO_mgl, DO_percent, Conductivity, Specific_conductivity, pH, Temperature) %>%
  dplyr::mutate(site = factor(site, levels = c("Stone_pile", "Control", "Reference"))) %>%
  tidyr::pivot_longer(cols = c(Temperature, DO_mgl, DO_percent, Conductivity, Specific_conductivity, pH), names_to = "Variable", values_to = "Values")

chemical_var_labels <- c(
  Temperature           = "Temperature (°C)",
  DO_mgl                = "Dissolved oxygen\n(mg/L)",
  DO_percent            = "Dissolved oxygen\n(%)",
  Conductivity          = "Conductivity\n(µS/cm)",
  Specific_conductivity = "Specific conductivity\n(µS/cm)",
  pH                    = "pH")

# Coloured by site type (not individual site_id) for consistency with every
# other panel — a 20-line/colour legend was competing with the data itself.
# x-axis uses monitoring_int + monitor_labels, same as the species-trends
# panel, so both panels read the same round labels.
Chemical_plot <- ggplot(Chemical_data2, aes(monitoring_int, Values, colour = site)) +
  geom_point(alpha = 0.5, position = position_jitter(width = 0.1)) +
  geom_smooth(aes(group = site), se = FALSE) +
  facet_wrap(~ Variable, scales = "free", nrow = 2, labeller = as_labeller(chemical_var_labels)) +
  scale_colour_manual(values = site_colours, labels = site_labels) +
  scale_x_continuous(breaks = 0:6, labels = monitor_labels) +
  labs(y = "Value", x = "Monitoring", colour = "Site type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

ggsave(file.path(out_dir, "fig-chemical-plot.png"), Chemical_plot, width = 8, height = 6,  dpi = 300)
```

### All species trends
```{r}
#| label: fig-species-trends
#| include: false
#| fig-cap: "Mean catch per monitoring round (± 95% CI) by site type for all monitored species."

# Species to plot and their display names
species_vec_trends <- c("Catfish", "Kōaro", "Goldfish")

species_labels <- c("Kōura" = "Kōura", "Catfish" = "Catfish", "Kōaro" = "Kōaro", "Goldfish" = "Goldfish")

# Function to build one trend plot per species
make_trend_plot <- function(sp) {

  col_name <- paste0("Weighted_CPUE_", sp)
  if (!col_name %in% names(all_data)) return(NULL)

  summary_df <- all_data %>%
    group_by(monitoring_int, site) %>%
    summarise(
      mean_n = mean(.data[[col_name]], na.rm = TRUE),
      se     = sd(.data[[col_name]], na.rm = TRUE) /
               sqrt(sum(!is.na(.data[[col_name]]))),
      n      = sum(!is.na(.data[[col_name]])),
      ci95   = qt(0.975, df = pmax(n - 1, 1)) * se,
      .groups = "drop") %>%
    mutate(
      site   = factor(site, levels = c("Stone_pile", "Control", "Reference")),
      ci_low  = pmax(mean_n - ci95, 0),
      ci_high = mean_n + ci95)

  ggplot(summary_df,
      aes(monitoring_int, mean_n, colour = site, shape = site, fill = site)) +
    geom_errorbar(aes(ymin = ci_low, ymax = ci_high), position = position_dodge(width = 0.4), width = 0.25, linewidth = 0.4, colour = "grey40") +
    geom_line(aes(group = site), position = position_dodge(width = 0.4), linewidth = 0.6) +
    geom_point(colour = "black", size = 2.5, stroke = 0.4, position = position_dodge(width = 0.4)) +
    scale_colour_manual(values = site_colours, labels = site_labels) +
    scale_fill_manual(values = site_fill, labels = site_labels) +
    scale_shape_manual(values = site_shapes, labels = site_labels) +
    scale_x_continuous(breaks = 0:6, labels = monitor_labels, limits = c(-0.3, 6.3)) +
    scale_y_continuous(trans = "log1p") +
    labs(title  = species_labels[[sp]], x = NULL, y = "Mean CPUE (± 95% CI, log scale)", colour = "Site type", shape = "Site type", fill = "Site type") +
    theme(axis.text.x     = element_text(size = 7,
        colour = c(rep("black", 6), "grey50")),
      axis.text.y     = element_text(size = 8),
      axis.title.y    = element_text(size = 9),
      plot.title      = element_text(size = 10, face = "bold"),
      legend.position = "none")   # collect legend from patchwork
}

# Generate all plots, drop NULLs
trend_plots <- purrr::map(species_vec_trends, make_trend_plot) %>%
  purrr::set_names(species_vec_trends) %>%
  purrr::compact()

combined_trends <- patchwork::wrap_plots(trend_plots, ncol = 1) +
  patchwork::plot_layout(guides = "collect") +
  patchwork::plot_annotation(theme= theme(legend.position = "right")) & theme(legend.position = "right")

# Save combined
ggsave(file.path(out_dir, "fig-all-species-trends.png"), combined_trends, width = 6, height = 6, dpi = 300)

# Also save individual plots if needed
#purrr::iwalk(trend_plots, ~ ggsave(filename = file.path(out_dir, paste0("fig-trend-", .y, ".png")), plot= .x, width= 8, height = 4, units = "in", dpi = 300))

```

### Combined figure
```{r}
#| label: fig-descriptive-context
#| fig-width: 14
#| fig-height: 12
#| dpi: 300
#| fig-cap: "Descriptive context plots, combined. Left column (top to bottom): baseline habitat characteristics (mean ± SE by site type), stone pile structure dimensions (5 stone pile sites), chemical conditions over time (mean ± smoothed trend by site type). Right column: mean catch per monitoring round by site type, for catfish, kōaro, and goldfish. Site-type colours are shared across all panels — see the legend on the right. Background/methods-supporting figures, not primary results — see Results for the BACI, trend, and convergence models."

# Left column stacks the 3 site-characteristic panels; right column is the
# species-trends panel (already a 3-row stack of its own), so both sides
# end up with matching internal proportions instead of a forced 2x2 grid
# that would otherwise squash the taller species-trends panel.
left_column <- Habitat_plot / stone_pile_structure_plot / Chemical_plot

descriptive_context_plot <- (left_column | combined_trends) +
  patchwork::plot_layout(widths = c(1.3, 1)) +
  patchwork::plot_annotation(tag_levels = "A")

ggsave(file.path(out_dir, "fig-descriptive-context.png"), descriptive_context_plot, width = 14, height = 12, dpi = 300)

descriptive_context_plot
```

### Kōura trend plot
```{r}
#| label: fig-koura-trends
#| fig-width: 5
#| fig-height: 3
#| dpi: 300
#| fig-cap: "Mean catch per monitoring round (± 95% CI) by site type for kōura."

# kōura trend plot — more detail than combined figure
koura_trend <- make_trend_plot("Kōura") +
  labs(title = NULL) +   # remove title since it's obvious
  theme(
    axis.text.x     = element_text(size = 9),
    axis.text.y     = element_text(size = 9),
    axis.title.y    = element_text(size = 10),
    legend.position = "right")

ggsave(file.path(out_dir, "fig-koura-trends.png"), koura_trend, width = 6, height = 3, dpi = 300)

koura_trend
```

### Kōura CPUE plot per site
```{r}
#| label: fig-koura-cpue
#| include: true
#| fig-width: 8
#| fig-height: 7
#| fig-cap: "Kōura CPUE (log scale) over time at each individual site, faceted by site ID. Point shape and smoother colour indicate site type."

site_id_order <- c(1, 2, 11, 12, 3, 4, 13, 14, 5, 6, 15, 16, 7, 8, 17, 18, 9, 10, 19, 20)

koura_cpue_plot <- all_data %>%
  dplyr::mutate(
    site    = factor(site, levels = c("Stone_pile", "Control", "Reference")),
    site_id = factor(as.numeric(as.character(site_id)), levels = site_id_order)) %>%
  ggplot(aes(monitoring_int, `Weighted_CPUE_Kōura`, shape = site, fill = site)) +
  geom_smooth(se = FALSE, aes(group = site, col = site)) +
  geom_point(size = 2, col = "black") +
  facet_wrap(~site_id, ncol = 4) +
  scale_colour_manual(values = site_colours, labels = site_labels) +
  scale_fill_manual(  values = site_fill,    labels = site_labels) +
  scale_shape_manual( values = site_shapes,  labels = site_labels) +
  scale_x_continuous(breaks = 0:6, labels = monitor_labels, limits = c(-0.3, 6.3)) +
  scale_y_continuous(trans = "log1p") +
  labs(y = "Kōura CPUE (log scale)", x = "Monitoring",
       col = "Site type", shape = "Site type", fill = "Site type") +
  theme(axis.text.x     = element_text(angle = 45, hjust = 1, size = 7),
        legend.position = "right")

ggsave(file.path(out_dir, "fig-koura-cpue.png"), koura_cpue_plot, width = 10, height = 7, dpi = 300)

koura_cpue_plot
```

## Combined environmental and fish summary table
```{r}
#| label: tbl-env-fish-summary
#| tbl-cap: Distribution of environmental and biotic variables measured at sampling sites. Summaries are grouped by site (ID) across repeated sampling events; an overall ‘All sites’ summary is also provided

group_var <- "site_id" 
var_order_exact <- c("Substrate index","Riparian vegetation","Overhanging trees","Wood cover","Emergent vegetation","Submerged vegetation", "Temperature","Dissolved oxygen","Specific conductivity","pH","Presence Kōura", "Presence Catfish","Presence Eel","Presence Goldfish","Presence Common smelt", "Presence Kōaro","Presence Trout")

ci95 <- function(x) {
  x <- x[!is.na(x)]
  n <- length(x)
  if (n < 2) return(c(NA_real_, NA_real_))
  se <- stats::sd(x) / sqrt(n)
  tcrit <- stats::qt(0.975, df = n - 1)
  m <- mean(x)
  c(m - tcrit * se, m + tcrit * se)
}

site_lookup <- all_data %>%
  dplyr::select(site_id, site) %>%
  dplyr::distinct() %>%
  dplyr::mutate(site_id   = as.character(site_id), site = as.character(site))

unit_map <- dplyr::tibble(Variable = c("Substrate_index_before","Riparian_vegetation","Overhanging_trees","Wood_cover", "Temperature","DO_mgl","Specific_conductivity","pH"))

Env_data <- all_data %>%
  dplyr::mutate(
    Emergent_vegetation  = Emergent_Native + Emergent_Non_Native,
    Submerged_vegetation = Submerged_Non_Native) %>%
  dplyr::select(
    dplyr::all_of(group_var),
    Substrate_index_before, Riparian_vegetation, Overhanging_trees, Wood_cover,
    Emergent_vegetation, Submerged_vegetation,
    Temperature, DO_mgl, Specific_conductivity, pH ) %>%
  tidyr::pivot_longer(
    cols = -dplyr::all_of(group_var),
    names_to = "Variable",
    values_to = "Value")

Env_summary_by_group <- Env_data %>%
  dplyr::group_by(dplyr::across(dplyr::all_of(group_var)), Variable) %>%
  dplyr::summarise(
    n       = sum(!is.na(Value)),
    Mean    = mean(Value, na.rm = TRUE),
    Median  = median(Value, na.rm = TRUE),
    Min     = min(Value, na.rm = TRUE),
    Max     = max(Value, na.rm = TRUE),
    CI_low  = ci95(Value)[1],
    CI_high = ci95(Value)[2],
    .groups = "drop") %>%
  dplyr::rename(Group = !!group_var) %>%
  dplyr::mutate(Group = as.character(Group))

Env_summary_all <- Env_data %>%
  dplyr::group_by(Variable) %>%
  dplyr::summarise(
    Group   = "All sites",
    n       = sum(!is.na(Value)),
    Mean    = mean(Value, na.rm = TRUE),
    Median  = median(Value, na.rm = TRUE),
    Min     = min(Value, na.rm = TRUE),
    Max     = max(Value, na.rm = TRUE),
    CI_low  = ci95(Value)[1],
    CI_high = ci95(Value)[2],
    .groups = "drop") %>%
  dplyr::mutate(Group = as.character(Group))

Env_summary_table <- dplyr::bind_rows(Env_summary_by_group, Env_summary_all)

# ---- Fish presence data (long) ----
Fish_data_long <- all_data %>%
  dplyr::select(
    dplyr::all_of(group_var),
    Presence_Kōura, Presence_Catfish, Presence_Eel, Presence_Goldfish, Presence_Common_smelt, Presence_Kōaro, Presence_Trout ) %>%
  tidyr::pivot_longer(
    cols = -dplyr::all_of(group_var),
    names_to = "Variable",
    values_to = "Presence")

Fish_summary_by_group <- Fish_data_long %>%
  dplyr::group_by(dplyr::across(dplyr::all_of(group_var)), Variable) %>%
  dplyr::summarise(
    n       = sum(!is.na(Presence)),
    Mean    = mean(Presence, na.rm = TRUE),
    Median  = median(Presence, na.rm = TRUE),
    Min     = min(Presence, na.rm = TRUE),
    Max     = max(Presence, na.rm = TRUE),
    CI_low  = ci95(Presence)[1],
    CI_high = ci95(Presence)[2],
    .groups = "drop") %>%
  dplyr::rename(Group = !!group_var) %>%
  dplyr::mutate(Group = as.character(Group))

Fish_summary_all <- Fish_data_long %>%
  dplyr::group_by(Variable) %>%
  dplyr::summarise(
    Group   = "All sites",
    n       = sum(!is.na(Presence)),
    Mean    = mean(Presence, na.rm = TRUE),
    Median  = median(Presence, na.rm = TRUE),
    Min     = min(Presence, na.rm = TRUE),
    Max     = max(Presence, na.rm = TRUE),
    CI_low  = ci95(Presence)[1],
    CI_high = ci95(Presence)[2],
    .groups = "drop") %>%
  dplyr::mutate(Group = as.character(Group))

Fish_summary_table <- dplyr::bind_rows(Fish_summary_by_group, Fish_summary_all)

# ---- Combine + attach units + pretty variable names + exact ordering ----
EnvBio_summary_table <- dplyr::bind_rows(Env_summary_table, Fish_summary_table) %>%
  dplyr::mutate(Group = as.character(Group)) %>%
  dplyr::left_join(
    site_lookup,
    by = c("Group" = "site_id")) %>%
  dplyr::left_join(unit_metadata, by = "Variable") %>%
  dplyr::mutate(
    Variable = dplyr::recode(
      Variable,
      Substrate_index_before       = "Substrate index",
      Riparian_vegetation   = "Riparian vegetation",
      Overhanging_trees     = "Overhanging trees",
      Wood_cover            = "Wood cover",
      Emergent_vegetation   = "Emergent vegetation",
      Submerged_vegetation  = "Submerged vegetation",
      Temperature           = "Temperature",
      DO_mgl                = "Dissolved oxygen",
      Specific_conductivity = "Specific conductivity",
      pH                    = "pH",
      Presence_Kōura        = "Presence Kōura",
      Presence_Catfish      = "Presence Catfish",
      Presence_Eel          = "Presence Eel",
      Presence_Goldfish     = "Presence Goldfish",
      Presence_Common_smelt = "Presence Common smelt",
      Presence_Kōaro        = "Presence Kōaro",
      Presence_Trout        = "Presence Trout" ),
    Unit = dplyr::coalesce(Unit, "")) %>%
  dplyr::mutate(
    Variable = factor(as.character(Variable), levels = var_order_exact),
    Group    = as.character(Group)) %>%
  dplyr::select(Group, Variable, Unit, n, Mean, Median, Min, Max, CI_low, CI_high) %>%
  dplyr::arrange(Variable, dplyr::desc(Group == "All sites"), Group)

# Optional: if you want Group sorted numerically when Group is ID
if (group_var == "site_id") {
  EnvBio_summary_table <- EnvBio_summary_table %>%
    dplyr::mutate(
      Group_num = suppressWarnings(as.numeric(Group))
    ) %>%
    dplyr::arrange(
      Variable,
      dplyr::desc(Group == "All sites"),
      Group_num,
      Group
    ) %>%
    dplyr::select(-Group_num)
}

write.csv(EnvBio_summary_table, file = file.path(out_dir, "EnvBio_summary_table.csv"), row.names = FALSE)

DT::datatable(
  EnvBio_summary_table,
  filter = "top",
  options = list(
    pageLength = 20,
    dom = 'Bfrtip',
    buttons = c('csv', 'excel'),
    columnDefs = list(list(className = 'dt-center', targets = 3:9))
  ),
  extensions = 'Buttons',
  rownames = FALSE,
  colnames = c("Site", "Variable", "Unit", "n", "Mean", "Median", "Min", "Max", "CI low", "CI high")
) |>
  DT::formatRound(columns = c("Mean","Median","Min","Max","CI_low","CI_high"), digits = 2)

```


## Baseline equivalence — Were sites comparable before construction?
```{r}
#| label: tbl-baseline-combined
#| tbl-cap: "Baseline (pre-construction) comparability of stone pile, control, and reference sites. Values are mean ± SD; p-values test Stone pile vs Control (paired Wilcoxon signed-rank test, pairs matched by pair_id, supporting the BACI design) and Stone pile vs Reference (unpaired Wilcoxon rank-sum test, documenting the starting gap relative to natural conditions, motivating the convergence analysis). With only 5 paired sites, these are low-power checks — a non-significant result indicates no obvious gross imbalance, not formal proof of equivalence."

# Baseline-only data (monitoring round 0), all three site types
baseline_data <- all_data %>%
  filter(monitoring_int == 0) %>%
  mutate(site = factor(site, levels = c("Stone_pile", "Control", "Reference")))

# Display order and nicer names for the paper 
var_labels <- c(
  Weighted_CPUE_Kōura    = "Kōura CPUE",
  Weighted_CPUE_Catfish  = "Catfish CPUE",
  Weighted_BPUE_Kōura    = "Kōura BPUE",
  Weighted_BPUE_Catfish  = "Catfish BPUE",
  Substrate_index_before = "Substrate index (before)",
  Substrate_index_after  = "Substrate index (after)",
  Slope_0_20m            = "Slope (0–20 m)",
  Temperature            = "Temperature (°C)",
  DO_mgl                 = "Dissolved oxygen (mg/L)",
  Specific_conductivity  = "Specific conductivity (µS/cm)",
  pH                     = "pH")

baseline_vars <- setdiff(names(var_labels), "Substrate_index_after")

# Stone pile vs Control — paired by pair_id (BACI design check).
paired_test <- function(var, data = baseline_data) {
  wide <- data %>%
    dplyr::filter(site %in% c("Stone_pile", "Control")) %>%
    dplyr::select(pair_id, site, value = dplyr::all_of(var)) %>%
    tidyr::pivot_wider(names_from = site, values_from = value)

  test <- suppressWarnings(wilcox.test(wide$Stone_pile, wide$Control, paired = TRUE, exact = FALSE))
  tibble::tibble(Comparison = "Stone pile vs Control (paired)", Variable = var,
                 n_pairs = nrow(wide), statistic = unname(test$statistic), p.value = test$p.value)
}

# Stone pile vs Reference — unpaired (documents the baseline gap for RQ2)
unpaired_test <- function(var, data = baseline_data) {
  d <- data %>%
    dplyr::filter(site %in% c("Stone_pile", "Reference")) %>%
    dplyr::select(site, value = dplyr::all_of(var)) %>%
    dplyr::filter(!is.na(value)) %>%
    dplyr::mutate(site = factor(site, levels = c("Stone_pile", "Reference")))

  test <- suppressWarnings(wilcox.test(value ~ site, data = d, exact = FALSE))
  tibble::tibble(Comparison = "Stone pile vs Reference (unpaired)", Variable = var,
                 n_pairs = NA_integer_, statistic = unname(test$statistic), p.value = test$p.value)
}

baseline_tests <- dplyr::bind_rows(
  purrr::map_dfr(baseline_vars, paired_test),
  purrr::map_dfr(baseline_vars, unpaired_test)) %>%
  dplyr::mutate(across(c(statistic, p.value), ~ round(.x, 3)))

#write.csv(baseline_tests, file.path(out_dir, "tbl-baseline-tests.csv"), row.names = FALSE)

# Combined view
baseline_group_summary <- baseline_data %>%
  dplyr::select(site, dplyr::all_of(baseline_vars)) %>%
  tidyr::pivot_longer(-site, names_to = "Variable", values_to = "Value") %>%
  dplyr::group_by(Variable, site) %>%
  dplyr::summarise(
    mean_sd = paste0(round(mean(Value, na.rm = TRUE), 2), " ± ", round(sd(Value, na.rm = TRUE), 2)),
    .groups = "drop") %>%
  tidyr::pivot_wider(names_from = site, values_from = mean_sd)

baseline_tests_wide <- baseline_tests %>%
  dplyr::mutate(Comparison = dplyr::if_else(
    stringr::str_detect(Comparison, "Control"), "p_vs_Control", "p_vs_Reference")) %>%
  dplyr::select(Variable, Comparison, p.value) %>%
  tidyr::pivot_wider(names_from = Comparison, values_from = p.value)

baseline_combined <- baseline_group_summary %>%
  dplyr::left_join(baseline_tests_wide, by = "Variable") %>%
  dplyr::select(Variable, Stone_pile, Control, Reference, p_vs_Control, p_vs_Reference)

after_data <- all_data %>%
  filter(monitoring_int >= 1) %>%
  distinct(site_id, pair_id, site, Substrate_index_after) %>%
  mutate(site = factor(site, levels = c("Stone_pile", "Control", "Reference")))

after_tests <- dplyr::bind_rows(
  paired_test("Substrate_index_after", data = after_data),
  unpaired_test("Substrate_index_after", data = after_data)) %>%
  dplyr::mutate(across(c(statistic, p.value), ~ round(.x, 3)))

after_group_summary <- after_data %>%
  dplyr::select(site, Substrate_index_after) %>%
  tidyr::pivot_longer(-site, names_to = "Variable", values_to = "Value") %>%
  dplyr::group_by(Variable, site) %>%
  dplyr::summarise(
    mean_sd = paste0(round(mean(Value, na.rm = TRUE), 2), " ± ", round(sd(Value, na.rm = TRUE), 2)),
    .groups = "drop") %>%
  tidyr::pivot_wider(names_from = site, values_from = mean_sd)

after_tests_wide <- after_tests %>%
  dplyr::mutate(Comparison = dplyr::if_else(
    stringr::str_detect(Comparison, "Control"), "p_vs_Control", "p_vs_Reference")) %>%
  dplyr::select(Variable, Comparison, p.value) %>%
  tidyr::pivot_wider(names_from = Comparison, values_from = p.value)

after_combined <- after_group_summary %>%
  dplyr::left_join(after_tests_wide, by = "Variable") %>%
  dplyr::select(Variable, Stone_pile, Control, Reference, p_vs_Control, p_vs_Reference)

baseline_combined <- dplyr::bind_rows(baseline_combined, after_combined) %>%
  dplyr::mutate(Variable = factor(Variable, levels = names(var_labels))) %>%
  dplyr::arrange(Variable) %>%
  dplyr::mutate(Variable = var_labels[as.character(Variable)])

write.csv(baseline_combined, file.path(out_dir, "tbl-baseline-combined.csv"), row.names = FALSE)

knitr::kable(baseline_combined, digits = 3)
```

# Statistical analysis
*BACI model*
RQ1 - Did stone pile construction increase species occupancy/abundance/biomass relative to control sites?
  M1ok - Occupancy, Koura,   Stone_pile vs Control, full timeline, Binomial
  M1oc - Occupancy, Catfish, Stone_pile vs Control, full timeline, Binomial
  M1ak - Abundance, Koura,   Stone_pile vs Control, full timeline, Tweedie
  M1ac - Abundance, Catfish, Stone_pile vs Control, full timeline, Tweedie
  M1bk - Biomass,   Koura,   Stone_pile vs Control, full timeline, Tweedie
  M1bc - Biomass,   Catfish, Stone_pile vs Control, full timeline, Tweedie

*Post-construction trend*
RQ2 - Did stone pile construction increase species abundance/biomass relative to control sites after construction?
  M2ak - Abundance, Koura,   Stone_pile vs Control, round 1-5, Tweedie
  M2ac - Abundance, Catfish, Stone_pile vs Control, round 1-5, Tweedie
  M2bk - Biomass,   Koura,   Stone_pile vs Control, round 1-5, Tweedie
  M2bc - Biomass,   Catfish, Stone_pile vs Control, round 1-5, Tweedie

*Convergence to reference*
RQ3 - Did species abundance/biomass at stone pile sites converge toward reference sites conditions over time?
  M3ak - Abundance, Koura,   Stone_pile vs Reference, full timeline, Tweedie
  M3ac - Abundance, Catfish, Stone_pile vs Reference, full timeline, Tweedie
  M3bk - Biomass,   Koura,   Stone_pile vs Reference, full timeline, Tweedie
  M3bc - Biomass,   Catfish, Stone_pile vs Reference, full timeline, Tweedie

*Baseline and equivalence*
RQ4 — Did species abundance/biomass differ between stone pile, control, and reference sites before construction, and did these differences change after construction?
  M4TBak - Treatment, Before, Abundance, Koura,   Stone_pile vs Control, round 0, Tweedie
  M4TBbk - Treatment, Before, Biomass,   Koura,   Stone_pile vs Control, round 0, Tweedie
  M4TBac - Treatment, Before, Abundance, Catfish, Stone_pile vs Control, round 0, Tweedie
  M4TBbc - Treatment, Before, Biomass,   Catfish, Stone_pile vs Control, round 0, Tweedie

  M4RBak - Reference, Before, Abundance, Koura,   Reference 1 vs Reference 2, round 0, Tweedie
  M4RBbk - Reference, Before, Biomass,   Koura,   Reference 1 vs Reference 2, round 0, Tweedie
  M4RBac - Reference, Before, Abundance, Catfish, Reference 1 vs Reference 2, round 0, Tweedie
  M4RBbc - Reference, Before, Biomass,   Catfish, Reference 1 vs Reference 2, round 0, Tweedie

  M4TAak - Treatment, After, Abundance, Koura,   Stone_pile vs Control, round 1-5, Tweedie
  M4TAbk - Treatment, After, Biomass,   Koura,   Stone_pile vs Control, round 1-5, Tweedie
  M4TAac - Treatment, After, Abundance, Catfish, Stone_pile vs Control, round 1-5, Tweedie
  M4TAbc - Treatment, After, Biomass,   Catfish, Stone_pile vs Control, round 1-5, Tweedie

  M4RAak - Reference, After, Abundance, Koura,   Reference 1 vs Reference 2, round 1-5, Tweedie
  M4RAbk - Reference, After, Biomass,   Koura,   Reference 1 vs Reference 2, round 1-5, Tweedie
  M4RAac - Reference, After, Abundance, Catfish, Reference 1 vs Reference 2, round 1-5, Tweedie
  M4RAbc - Reference, After, Biomass,   Catfish, Reference 1 vs Reference 2, round 1-5, Tweedie

## Model data subsets
```{r}
#| label: model-data-subsets

# RQ1 Dataset
baci_data <- all_data %>%
  filter(site %in% c("Stone_pile", "Control")) %>%
  mutate(
    Period  = factor(Period, levels = c("Before", "After")),
    site    = factor(site,   levels = c("Control", "Stone_pile")),
    site_id = factor(site_id))

# RQ2 Dataset
baci_data_after <- baci_data %>%
  dplyr::filter(monitoring_int >= 1)

# RQ3 Dataset
convergence_data <- all_data %>%
  filter(site %in% c("Stone_pile", "Reference")) %>%
  mutate(
    site    = factor(site, levels = c("Reference", "Stone_pile")),
    site_id = factor(site_id),
    log_cpue_catfish = log1p(Weighted_CPUE_Catfish),
    log_cpue_eel     = log1p(Weighted_CPUE_Eel),
    log_cpue_trout   = log1p(Weighted_CPUE_Trout))

all_species <- c("Kōura", "Catfish")

species_letter_map  <- c(Kōura = "k", Catfish = "c")

response_letter_map <- c(occupancy = "o", abundance = "a", biomass = "b")

repoint_call <- function(m, form, data_name, family_expr) {
  if (is.null(m)) return(NULL)
  m$call$formula <- as.formula(form)
  m$call$data    <- as.name(data_name)
  m$call$family  <- family_expr
  m
}

# Pulls one species/response-type's fitted model out of a results tibble
pull_model <- function(results, sp, resp) {
  results %>%
    dplyr::filter(species == sp, response_type == resp) %>%
    dplyr::pull(model) %>%
    purrr::pluck(1)
}
```

## RQ1 — Effect of construction (step change: occupancy, abundance, biomass)
```{r}
#| label: tbl-rq1-full
#| tbl-cap: "RQ1 (step change) — full fixed-effects estimates, kōura and catfish, occupancy/abundance/biomass."

fit_rq1 <- function(species, response_type) {
  response_col <- dplyr::case_when(
    response_type == "occupancy" ~ paste0("Presence_", species),
    response_type == "abundance" ~ paste0("Weighted_CPUE_", species),
    response_type == "biomass"   ~ paste0("Weighted_BPUE_", species))
  if (!response_col %in% names(baci_data)) return(NULL)

  form    <- paste0("`", response_col, "` ~ Period * site + monitoring_int + Temperature + (1 | pair_id / site_id)")
  fam_obj <- if (response_type == "occupancy") binomial() else tweedie()
  zif     <- if (species == "Catfish" && response_type == "abundance") ~1 else ~0
  m <- tryCatch(glmmTMB::glmmTMB(as.formula(form), data = baci_data, REML = TRUE, family = fam_obj,
                  ziformula = zif),
                error = function(e) NULL)
  if (!is.null(m) && m$fit$convergence != 0)
    m <- tryCatch(glmmTMB::glmmTMB(as.formula(form), data = baci_data, REML = TRUE, family = fam_obj,
                    ziformula = zif,
                    control = glmmTMBControl(optCtrl = list(iter.max = 1000, eval.max = 2000))),
                  error = function(e) NULL)
  repoint_call(m, form, "baci_data",
               if (response_type == "occupancy") quote(binomial()) else quote(tweedie()))
}

rq1_results <- tidyr::crossing(species = all_species, response_type = c("occupancy", "abundance", "biomass")) %>%
  dplyr::mutate(
    RQ                = "RQ1",
    comparison        = "vs_control",
    time_frame        = "full",
    term_of_interest  = "PeriodAfter:siteStone_pile",
    model_id          = paste0("M1", response_letter_map[response_type], species_letter_map[species]),
    model             = purrr::map2(species, response_type, fit_rq1)) %>%
  dplyr::arrange(match(species, names(species_letter_map)), match(response_type, names(response_letter_map))) %>%
  dplyr::mutate(
    p_value = purrr::map2_dbl(model, term_of_interest, function(m, trm) {
      if (is.null(m)) return(NA_real_)
      broom.mixed::tidy(m, effects = "fixed") %>%
        dplyr::filter(term == trm) %>%
        dplyr::pull(p.value)
    }),
    p_adj_BH = p.adjust(p_value, method = "BH"))

rq1_full_tidy <- rq1_results %>%
  dplyr::mutate(tidy = purrr::map(model, ~ if (is.null(.x)) NULL else broom.mixed::tidy(.x, effects = "fixed", conf.int = TRUE))) %>%
  dplyr::filter(!purrr::map_lgl(tidy, is.null)) %>%
  tidyr::unnest(tidy) %>%
  dplyr::mutate(p_adj_BH = dplyr::if_else(term == term_of_interest, p_adj_BH, NA_real_)) %>%
  dplyr::select(model_id, species, response_type, term, estimate, std.error, statistic, p.value, p_adj_BH, conf.low, conf.high) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 3)))

DT::datatable(
  rq1_full_tidy,
  filter = "top",
  options = list(pageLength = 38, dom = 'Bfrtip', buttons = c('csv', 'excel')),
  extensions = 'Buttons',
  rownames = FALSE
) |>
  DT::formatRound(columns = c("estimate","std.error","statistic","p.value","p_adj_BH","conf.low","conf.high"), digits = 3)
```

## RQ2 — Post-construction trend (abundance and biomass, Stone_pile vs Control)
```{r}
#| label: tbl-rq2-full
#| tbl-cap: "RQ2 (post-construction trend) — full fixed-effects estimates, kōura and catfish, abundance/biomass."

fit_rq2 <- function(species, response_type) {
  response_col <- if (response_type == "abundance") paste0("Weighted_CPUE_", species) else paste0("Weighted_BPUE_", species)
  if (!response_col %in% names(baci_data_after)) return(NULL)

  form <- paste0("`", response_col, "` ~ site * monitoring_int + Temperature + Substrate_index_before + (1 | site_id)")
  zif  <- if (species == "Catfish" && response_type == "abundance") ~1 else ~0
  m <- tryCatch(glmmTMB::glmmTMB(as.formula(form), data = baci_data_after, REML = TRUE, family = tweedie(),
                  ziformula = zif),
                error = function(e) NULL)
  if (!is.null(m) && m$fit$convergence != 0)
    m <- tryCatch(glmmTMB::glmmTMB(as.formula(form), data = baci_data_after, REML = TRUE, family = tweedie(),
                    ziformula = zif,
                    control = glmmTMBControl(optCtrl = list(iter.max = 1000, eval.max = 2000))),
                  error = function(e) NULL)
  repoint_call(m, form, "baci_data_after", quote(tweedie()))
}

rq2_results <- tidyr::crossing(species = all_species, response_type = c("abundance", "biomass")) %>%
  dplyr::mutate(
    RQ                = "RQ2",
    comparison        = "vs_control",
    time_frame        = "post",
    term_of_interest  = "siteStone_pile:monitoring_int",
    model_id          = paste0("M2", response_letter_map[response_type], species_letter_map[species]),
    model             = purrr::map2(species, response_type, fit_rq2)) %>%
  dplyr::arrange(match(species, names(species_letter_map)), match(response_type, names(response_letter_map))) %>%
  dplyr::mutate(
    p_value = purrr::map2_dbl(model, term_of_interest, function(m, trm) {
      if (is.null(m)) return(NA_real_)
      broom.mixed::tidy(m, effects = "fixed") %>%
        dplyr::filter(term == trm) %>%
        dplyr::pull(p.value)
    }),
    p_adj_BH = p.adjust(p_value, method = "BH"))

rq2_full_tidy <- rq2_results %>%
  dplyr::mutate(tidy = purrr::map(model, ~ if (is.null(.x)) NULL else broom.mixed::tidy(.x, effects = "fixed", conf.int = TRUE))) %>%
  dplyr::filter(!purrr::map_lgl(tidy, is.null)) %>%
  tidyr::unnest(tidy) %>%
  dplyr::mutate(p_adj_BH = dplyr::if_else(term == term_of_interest, p_adj_BH, NA_real_)) %>%
  dplyr::select(model_id, species, response_type, term, estimate, std.error, statistic, p.value, p_adj_BH, conf.low, conf.high) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 3)))

DT::datatable(
  rq2_full_tidy,
  filter = "top",
  options = list(pageLength = 26, dom = 'Bfrtip', buttons = c('csv', 'excel')),
  extensions = 'Buttons',
  rownames = FALSE
) |>
  DT::formatRound(columns = c("estimate","std.error","statistic","p.value","p_adj_BH","conf.low","conf.high"), digits = 3)
```

## RQ3 — Convergence to reference conditions (abundance and biomass, Stone_pile vs Reference)
```{r}
#| label: tbl-rq3-full
#| tbl-cap: "RQ3 (convergence to Reference) — full fixed-effects estimates, kōura and catfish, abundance/biomass."

fit_rq3 <- function(species, response_type) {
  response_col <- if (response_type == "abundance") paste0("Weighted_CPUE_", species) else paste0("Weighted_BPUE_", species)
  if (!response_col %in% names(convergence_data)) return(NULL)

  form <- paste0("`", response_col, "` ~ site * monitoring_int + Season + Substrate_index_after + (1 | site_id)")
  zif  <- if (species == "Catfish" && response_type == "abundance") ~1 else ~0
  m <- tryCatch(glmmTMB::glmmTMB(as.formula(form), data = convergence_data, REML = TRUE, family = tweedie(),
                  ziformula = zif),
                error = function(e) NULL)
  repoint_call(m, form, "convergence_data", quote(tweedie()))
}

rq3_results <- tidyr::crossing(species = all_species, response_type = c("abundance", "biomass")) %>%
  dplyr::mutate(
    RQ                = "RQ3",
    comparison        = "vs_reference",
    time_frame        = "full",
    term_of_interest  = "siteStone_pile:monitoring_int",
    model_id          = paste0("M3", response_letter_map[response_type], species_letter_map[species]),
    model             = purrr::map2(species, response_type, fit_rq3)) %>%
  dplyr::arrange(match(species, names(species_letter_map)), match(response_type, names(response_letter_map))) %>%
  dplyr::mutate(
    p_value = purrr::map2_dbl(model, term_of_interest, function(m, trm) {
      if (is.null(m)) return(NA_real_)
      broom.mixed::tidy(m, effects = "fixed") %>%
        dplyr::filter(term == trm) %>%
        dplyr::pull(p.value)
    }),
    p_adj_BH = p.adjust(p_value, method = "BH"))

rq3_full_tidy <- rq3_results %>%
  dplyr::mutate(tidy = purrr::map(model, ~ if (is.null(.x)) NULL else broom.mixed::tidy(.x, effects = "fixed", conf.int = TRUE))) %>%
  dplyr::filter(!purrr::map_lgl(tidy, is.null)) %>%
  tidyr::unnest(tidy) %>%
  dplyr::mutate(p_adj_BH = dplyr::if_else(term == term_of_interest, p_adj_BH, NA_real_)) %>%
  dplyr::select(model_id, species, response_type, term, estimate, std.error, statistic, p.value, p_adj_BH, conf.low, conf.high) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 3)))

DT::datatable(
  rq3_full_tidy,
  filter = "top",
  options = list(pageLength = 34, dom = 'Bfrtip', buttons = c('csv', 'excel')),
  extensions = 'Buttons',
  rownames = FALSE
) |>
  DT::formatRound(columns = c("estimate","std.error","statistic","p.value","p_adj_BH","conf.low","conf.high"), digits = 3)
```

### Predicted trajectories (kōura abundance convergence)
```{r}
#| label: fig-convergence-plot
#| fig-width: 8
#| fig-height: 4
#| dpi: 300
#| fig-cap: "Predicted kōura CPUE trajectories over monitoring rounds at stone pile and reference sites from the RQ3 abundance convergence model. Shaded ribbons represent 95% confidence intervals. The annotation shows the key interaction term (β ± p-value) and model R² values for fixed effects (R²m) and conditional (R²c)."


M3ak <- pull_model(rq3_results, "Kōura", "abundance")

# Predicted trajectories at each monitoring round.
season_lookup <- c(`0` = "Summer", `1` = "Autumn", `2` = "Winter", `3` = "Spring",  `4` = "Summer", `5` = "Autumn", `6` = "Winter")

emmeans::emm_options(msg.interaction = FALSE)

site_substrate_lookup <- convergence_data %>%
  dplyr::distinct(site, monitoring_int, Substrate_index_after) %>%
  dplyr::mutate(period = dplyr::if_else(monitoring_int == 0, "baseline", "post")) %>%
  dplyr::group_by(site, period) %>%
  dplyr::summarise(Substrate_index_after = mean(Substrate_index_after, na.rm = TRUE), .groups = "drop")

emm_m3ak <- purrr::map_dfr(0:6, function(i) {
  period_i <- if (i == 0) "baseline" else "post"
  purrr::map_dfr(c("Stone_pile", "Reference"), function(s) {
    sub_idx <- site_substrate_lookup$Substrate_index_after[
      site_substrate_lookup$site == s & site_substrate_lookup$period == period_i]
    emmeans(M3ak, ~ site,
      at   = list(monitoring_int = i, Season = season_lookup[[as.character(i)]],
                   Substrate_index_after = sub_idx),
      type = "response") %>%
      as.data.frame() %>%
      dplyr::filter(site == s) %>%
      dplyr::mutate(monitoring_int = i)
  })
}) %>%
  dplyr::rename(lower.CL = dplyr::any_of("asymp.LCL"),
                upper.CL = dplyr::any_of("asymp.UCL")) %>%
  mutate(site = factor(site, levels = c("Stone_pile", "Reference")))

# Clean numeric values for annotation
beta_val <- round(glmmTMB::fixef(M3ak)$cond["siteStone_pile:monitoring_int"], 3)
p_val    <- round(summary(M3ak)$coefficients$cond["siteStone_pile:monitoring_int", "Pr(>|z|)"], 3)
r2_vals  <- MuMIn::r.squaredGLMM(M3ak)
r2m_val  <- round(r2_vals[1, "R2m"], 2)
r2c_val  <- round(r2_vals[1, "R2c"], 2)

# Use plotmath-safe expression
model_label <- paste0("italic(beta)[Stone_pile%*%time]==", beta_val, "~(italic(p)==", p_val, ")", "~~italic(R)[m]^2==", r2m_val, "~~italic(R)[c]^2==", r2c_val)

convergence_data_plot <- convergence_data %>%
  filter(!is.na(`Weighted_CPUE_Kōura`)) %>%
  mutate(site = factor(site, levels = c("Stone_pile", "Reference")))

convergence_plot <- ggplot() +
  geom_point(data = convergence_data_plot, aes(monitoring_int, `Weighted_CPUE_Kōura`,  fill = site, shape = site),  colour = "black",  alpha = 0.4, size = 1.8, stroke = 0.5, position = position_jitter(width = 0.08, seed = 42),show.legend = FALSE) +
  geom_ribbon(data = emm_m3ak, aes(monitoring_int, response,  ymin = lower.CL, ymax = upper.CL, fill = site),  alpha = 0.15, colour = NA,  show.legend = FALSE) +
  geom_line(data = emm_m3ak,  aes(monitoring_int, response,  colour = site, group = site), linewidth = 1.1) +
  geom_point(data = emm_m3ak, aes(monitoring_int, response,  fill = site, shape = site), colour = "black", size = 3, stroke = 0.4) +
  annotate("text", x = 0, y = Inf, label = model_label, parse = TRUE,  hjust = 0, vjust = 1.4, size = 3.2, colour = "grey30") +
  scale_colour_manual(values = conv_colours, labels = conv_labels) +
  scale_fill_manual(  values = conv_fill,    labels = conv_labels) +
  scale_shape_manual( values = conv_shapes,  labels = conv_labels) +
  scale_x_continuous(breaks = 0:6, labels = monitor_labels, limits = c(-0.3, 6.3)) +
  scale_y_continuous(trans = "log1p") +
  labs(x=NULL ,y= "Kōura CPUE (log scale)",colour = "Site type",fill= "Site type",shape  = "Site type") +
  theme(axis.text.x = element_text(size = 9), axis.text.y = element_text(size = 9), axis.title.y = element_text(size = 10), legend.position = "right")+
  theme(axis.text.x = element_text(colour = c(rep("black", 6), "grey50")))

ggsave(file.path(out_dir, "fig-convergence-plot.png"), convergence_plot, width = 6, height = 3, dpi = 300)

convergence_plot
```

## RQ4 — Baseline and equivalence (abundance and biomass, Stone_pile vs Control / Reference 1 vs Reference 2)
```{r}
#| label: tbl-rq4-full
#| tbl-cap: "RQ4 — abundance and biomass before and after construction: Stone pile vs Control (treatment comparison) and Reference 1 vs Reference 2 (reference equivalence). BH-adjusted p-values are applied within each sub-question group."

# RQ4 Data subsets 
rq4_trt_before <- all_data %>%
  dplyr::filter(site %in% c("Stone_pile", "Control"), monitoring == 0) %>%
  dplyr::mutate(
    Site_type_detail = factor(Site_type_detail, levels = c("Control", "Stone_pile")),
    pair_id = factor(pair_id), site_id = factor(site_id))

rq4_trt_after <- all_data %>%
  dplyr::filter(site %in% c("Stone_pile", "Control"), monitoring_int >= 1) %>%
  dplyr::mutate(
    Site_type_detail = factor(Site_type_detail, levels = c("Control", "Stone_pile")),
    pair_id = factor(pair_id), site_id = factor(site_id))

rq4_ref_before <- all_data %>%
  dplyr::filter(site == "Reference", monitoring == 0) %>%
  dplyr::mutate(
    Site_type_detail = factor(Site_type_detail, levels = c("Reference_1", "Reference_2")),
    pair_id = factor(pair_id), site_id = factor(site_id))

rq4_ref_after <- all_data %>%
  dplyr::filter(site == "Reference", monitoring_int >= 1) %>%
  dplyr::mutate(
    Site_type_detail = factor(Site_type_detail, levels = c("Reference_1", "Reference_2")),
    pair_id = factor(pair_id), site_id = factor(site_id))

# Fit functions 
# Before
fit_rq4_before <- function(data, species, response_type) {
  response_col <- if (response_type == "abundance") paste0("Weighted_CPUE_", species) else paste0("Weighted_BPUE_", species)
  if (!response_col %in% names(data)) return(NULL)
  form <- paste0("`", response_col, "` ~ Site_type_detail")
  tryCatch(
    glmmTMB::glmmTMB(as.formula(form), data = data, family = tweedie()),
    error = function(e) NULL)
}

# After: longitudinal — interaction with monitoring time.
# Temperature inclusion per model-selection (tbl-rq4-temperature-selection):
#   - Biomass (both species, both comparisons): drop — ΔAICc > 2 without it
#   - Catfish abundance treatment: drop — Temperature causes non-PD Hessian
#   - Catfish abundance reference: keep — dropping causes non-PD Hessian
#   - Kōura abundance (both): keep — marginal (|ΔAICc| < 2)
# Catfish abundance gets zero-inflation (excess zeros in count data).
fit_rq4_after <- function(data, species, response_type, drop_temp = FALSE) {
  response_col <- if (response_type == "abundance") paste0("Weighted_CPUE_", species) else paste0("Weighted_BPUE_", species)
  if (!response_col %in% names(data)) return(NULL)
  zif  <- if (species == "Catfish" && response_type == "abundance") ~1 else ~0
  form <- if (drop_temp)
    paste0("`", response_col, "` ~ Site_type_detail * monitoring_int + (1 | pair_id / site_id)")
  else
    paste0("`", response_col, "` ~ Site_type_detail * monitoring_int + Temperature + (1 | pair_id / site_id)")

  conv_failed <- FALSE
  m <- withCallingHandlers(
    tryCatch(
      glmmTMB::glmmTMB(as.formula(form), data = data, REML = TRUE, family = tweedie(),
        ziformula = zif),
      error = function(e) NULL),
    warning = function(w) {
      if (grepl("convergence|false conv", conditionMessage(w), ignore.case = TRUE))
        conv_failed <<- TRUE
      invokeRestart("muffleWarning")
    })

  if (conv_failed || is.null(m)) {
    m <- tryCatch(
      glmmTMB::glmmTMB(as.formula(form), data = data, REML = TRUE, family = tweedie(),
        ziformula = zif,
        control = glmmTMBControl(optCtrl = list(iter.max = 1000, eval.max = 2000))),
      error = function(e) NULL)
  }
  m
}

# Treatment after: drop Temperature for biomass (both species) + catfish abundance
fit_rq4_trt_after <- function(data, species, response_type)
  fit_rq4_after(data, species, response_type,
    drop_temp = response_type == "biomass" || (species == "Catfish" && response_type == "abundance"))

# Reference after: drop Temperature for biomass only
fit_rq4_ref_after <- function(data, species, response_type)
  fit_rq4_after(data, species, response_type,
    drop_temp = response_type == "biomass")

# Helper to build a results tibble for one sub-question 
make_rq4_results <- function(data, fit_fn, comparison, time_frame, term_of_interest, model_prefix) {
  tidyr::crossing(species = all_species, response_type = c("abundance", "biomass")) %>%
    dplyr::mutate(
      RQ               = "RQ4",
      comparison       = comparison,
      time_frame       = time_frame,
      term_of_interest = term_of_interest,
      model_id         = paste0(model_prefix, response_letter_map[response_type], species_letter_map[species]),
      model            = purrr::map2(species, response_type, ~ fit_fn(data, .x, .y))) %>%
    dplyr::arrange(match(species, names(species_letter_map)), match(response_type, names(response_letter_map))) %>%
    dplyr::mutate(
      p_value  = purrr::map2_dbl(model, term_of_interest, function(m, trm) {
        if (is.null(m)) return(NA_real_)
        broom.mixed::tidy(m, effects = "fixed", component = "cond") %>%
          dplyr::filter(term == trm) %>%
          dplyr::pull(p.value)
      }),
      p_adj_BH = p.adjust(p_value, method = "BH"))
}

# Fit all four sub-questions 
# Model ID scheme: M4 + T/R (treatment/reference) + B/A (before/after) + a/b + k/c
rq4_trt_before_results <- make_rq4_results(
  rq4_trt_before, fit_rq4_before,
  comparison = "treatment", time_frame = "before",
  term_of_interest = "Site_type_detailStone_pile",
  model_prefix = "M4TB")

rq4_trt_after_results <- make_rq4_results(
  rq4_trt_after, fit_rq4_trt_after,
  comparison = "treatment", time_frame = "after",
  term_of_interest = "Site_type_detailStone_pile:monitoring_int",
  model_prefix = "M4TA")

rq4_ref_before_results <- make_rq4_results(
  rq4_ref_before, fit_rq4_before,
  comparison = "reference", time_frame = "before",
  term_of_interest = "Site_type_detailReference_2",
  model_prefix = "M4RB")

rq4_ref_after_results <- make_rq4_results(
  rq4_ref_after, fit_rq4_ref_after,
  comparison = "reference", time_frame = "after",
  term_of_interest = "Site_type_detailReference_2:monitoring_int",
  model_prefix = "M4RA")

rq4_results <- dplyr::bind_rows(
  rq4_trt_before_results,
  rq4_ref_before_results,
  rq4_trt_after_results,
  rq4_ref_after_results)

# Full tidy table
rq4_full_tidy <- rq4_results %>%
  dplyr::mutate(
    time_frame  = factor(time_frame,  levels = c("before", "after")),
    comparison  = factor(comparison,  levels = c("treatment", "reference")),
    species     = factor(species,     levels = names(species_letter_map)),
    response_type = factor(response_type, levels = names(response_letter_map))) %>%
  dplyr::arrange(time_frame, comparison, species, response_type) %>%
  dplyr::mutate(tidy = purrr::map(model, ~ if (is.null(.x)) NULL else
    broom.mixed::tidy(.x, effects = "fixed", component = "cond", conf.int = TRUE))) %>%
  dplyr::filter(!purrr::map_lgl(tidy, is.null)) %>%
  tidyr::unnest(tidy) %>%
  dplyr::mutate(p_adj_BH = dplyr::if_else(term == term_of_interest, p_adj_BH, NA_real_)) %>%
  dplyr::select(model_id, comparison, time_frame, species, response_type,
                term, estimate, std.error, statistic, p.value, p_adj_BH, conf.low, conf.high) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 3)))

print(rq4_full_tidy, n=53)

DT::datatable(
  rq4_full_tidy,
  filter = "top",
  options = list(pageLength = 51, dom = 'Bfrtip', buttons = c('csv', 'excel')),
  extensions = 'Buttons',
  rownames = FALSE
) |>
  DT::formatRound(columns = c("estimate","std.error","statistic","p.value","p_adj_BH","conf.low","conf.high"), digits = 3)



```

### RQ4 Model figure
```{r}
#| label: fig-RQ4-models
#| fig-width: 8
#| fig-height: 10
#| fig-cap: "Observed distributions of kōura and catfish CPUE and BPUE by site type and monitoring period (before/after stone pile installation). Boxes show interquartile range, line = median, whiskers extend to 1.5×IQR."

rq4_combined <- dplyr::bind_rows(
  dplyr::mutate(rq4_trt_before, period = "Before"),
  dplyr::mutate(rq4_trt_after,  period = "After"),
  dplyr::mutate(rq4_ref_before, period = "Before"),
  dplyr::mutate(rq4_ref_after,  period = "After")) |>
  dplyr::mutate(
    period          = factor(period, levels = c("Before", "After")),
    Site_type_detail = factor(Site_type_detail,
      levels = c("Stone_pile", "Control", "Reference_1", "Reference_2")))

response_cols <- intersect(
  c(paste0("Weighted_CPUE_", all_species), paste0("Weighted_BPUE_", all_species)),
  names(rq4_combined))

rq4_long <- rq4_combined |>
  tidyr::pivot_longer(cols = dplyr::all_of(response_cols),
                      names_to = "response_var", values_to = "value") |>
  dplyr::mutate(
    species  = stringr::str_extract(response_var,
                 paste(all_species, collapse = "|")),
    response = dplyr::if_else(stringr::str_detect(response_var, "CPUE"),
                              "Abundance (CPUE)", "Biomass (BPUE)"),
    panel    = factor(paste0(species, " — ", response),
                levels = c(paste0(all_species[1], " — Abundance (CPUE)"),
                           paste0(all_species[1], " — Biomass (BPUE)"),
                           paste0(all_species[2], " — Abundance (CPUE)"),
                           paste0(all_species[2], " — Biomass (BPUE)"))))

ggplot2::ggplot(rq4_long,
    ggplot2::aes(x = Site_type_detail, y = value, fill = period)) +
  ggplot2::geom_boxplot(outlier.size = 0.6,
                        position = ggplot2::position_dodge(0.8)) +
  ggplot2::facet_wrap(~ panel, scales = "free_y", ncol = 2) +
  ggplot2::scale_fill_manual(values = c("Before" = "#c9a96e", "After" = "#3d7a5e")) +
  ggplot2::scale_x_discrete(labels = c(
    "Stone_pile"  = "Stone pile",
    "Control"     = "Control",
    "Reference_1" = "Reference 1",
    "Reference_2" = "Reference 2")) +
  ggplot2::scale_y_continuous(trans = "log1p") +
  ggplot2::labs(x = NULL, y = "log(value + 1)", fill = "Period") +
  ggplot2::theme_bw() +
  ggplot2::theme(
    strip.text      = ggplot2::element_text(face = "bold"),
    axis.text.x     = ggplot2::element_text(angle = 35, hjust = 1),
    legend.position = "bottom")
```



## Model diagnostics 
```{r}
#| label: fig-residual-diagnostics
#| fig-width: 8
#| fig-height: 10
#| fig-cap: "DHARMa scaled-residual diagnostics for each model: QQ plot (left) and residual vs. predicted (right). Red flags indicate a significant departure from the expected uniform distribution, dispersion, or outlier rate."

model_registry <- dplyr::bind_rows(rq1_results, rq2_results, rq3_results, rq4_results) %>%
  dplyr::select(model_id, RQ, species, response_type, comparison, time_frame, term_of_interest, p_adj_BH, model)


is_dharma_safe <- function(m) {
  if (is.null(m) || m$fit$convergence != 0) return(FALSE)
  tryCatch({
    v <- diag(as.matrix(vcov(m, full = TRUE)))
    all(is.finite(v) & v > 0)
  }, error = function(e) FALSE, warning = function(w) FALSE)
}

diagnostic_models <- model_registry %>%
  dplyr::filter(!(RQ == "RQ4" & time_frame == "before")) %>%
  dplyr::filter(purrr::map_lgl(model, is_dharma_safe)) %>%
  { purrr::set_names(.$model, .$model_id) } %>%
  purrr::compact()

simulate_check <- function(model, name, seed = 42) {
  sim <- DHARMa::simulateResiduals(model, n = 1000, seed = seed)
  plot(sim, main = name)
  sim
}

dharma_sims <- purrr::imap(diagnostic_models, simulate_check)
```

### Residual diagnostics summary table
```{r}
#| label: tbl-residual-diagnostics
#| tbl-cap: "DHARMa goodness-of-fit tests per model. p < 0.05 indicates a significant departure from model assumptions (uniformity of residuals, over/under-dispersion, or excess outliers) and warrants closer inspection of the corresponding diagnostic plot above."

dharma_tests <- purrr::imap_dfr(dharma_sims, function(sim, name) {
  tibble(
    Model           = name,
    Uniformity_p    = DHARMa::testUniformity(sim, plot = FALSE)$p.value,
    Dispersion_p    = DHARMa::testDispersion(sim, plot = FALSE)$p.value,
    Outlier_p       = DHARMa::testOutliers(sim, plot = FALSE)$p.value)
}) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

knitr::kable(dharma_tests)

write.csv(dharma_tests, file.path(out_dir, "tbl-residual-diagnostics.csv"), row.names = FALSE)
```

# Model results
## Question structure reference table
```{r}
#| label: tbl-question-structure
#| tbl-cap: "Model structure for each of the 3 questions tested per species and response type (Table X). Each question uses one consistent formula across species/response types, fit via glmmTMB (Tweedie for abundance/biomass, binomial for occupancy)."

question_structure <- model_registry %>%
  dplyr::distinct(RQ, response_type, comparison, time_frame, term_of_interest) %>%
  dplyr::mutate(
    family = dplyr::if_else(response_type == "occupancy", "binomial", "tweedie"),
    data = dplyr::case_when(
      comparison == "vs_control" & time_frame == "full" ~ "baci_data",
      comparison == "vs_control" & time_frame == "post" ~ "baci_data_after",
      comparison == "vs_reference"                      ~ "convergence_data",
      comparison == "treatment"  & time_frame == "before" ~ "rq4_trt_before",
      comparison == "treatment"  & time_frame == "after"  ~ "rq4_trt_after",
      comparison == "reference"  & time_frame == "before" ~ "rq4_ref_before",
      comparison == "reference"  & time_frame == "after"  ~ "rq4_ref_after"),
    formula = dplyr::case_when(
      comparison == "vs_control" & time_frame == "full" ~
        "<response> ~ Period * site + monitoring_int + Temperature + (1 | pair_id / site_id)",
      comparison == "vs_control" & time_frame == "post" ~
        "<response> ~ site * monitoring_int + Temperature + Substrate_index_before + (1 | site_id)",
      comparison == "vs_reference" ~
        "<response> ~ site * monitoring_int + Season + Substrate_index_after + (1 | site_id)",
      comparison %in% c("treatment", "reference") & time_frame == "before" ~
        "<response> ~ Site_type_detail + (1 | pair_id / site_id)",
      comparison == "treatment" & time_frame == "after" ~
        "<response> ~ Site_type_detail * monitoring_int + Temperature† + (1 | pair_id / site_id)",
      comparison == "reference" & time_frame == "after" ~
        "<response> ~ Site_type_detail * monitoring_int + Temperature† + (1 | pair_id / site_id)")) %>%
  dplyr::mutate(
    time_frame = factor(time_frame, levels = c("full", "post", "before", "after")),
    comparison = factor(comparison, levels = c("vs_control", "vs_reference", "treatment", "reference"))) %>%
  dplyr::arrange(RQ, time_frame, comparison, response_type) %>%
  dplyr::select(RQ, response_type, comparison, time_frame, family, data, formula, term_of_interest)

write.csv(question_structure, file.path(out_dir, "tbl-question-structure.csv"), row.names = FALSE)

knitr::kable(question_structure)

```


## Models summary table
```{r}
#| label: tbl-models-summary
#| include: true
#| tbl-cap: "Key interaction-term estimate (β, SE, 95% CI, p-value) for every model in model_registry (RQ1 step-change, RQ2 trend, RQ3 convergence; occupancy/abundance/biomass; kōura and catfish)."

tidy_all <- model_registry %>%
  dplyr::mutate(tidy = purrr::map2(model, term_of_interest, function(m, trm) {
    if (is.null(m)) return(NULL)
    broom.mixed::tidy(m, effects = "fixed", conf.int = TRUE) %>%
      dplyr::filter(term == trm)
  })) %>%
  dplyr::filter(!purrr::map_lgl(tidy, is.null)) %>%
  tidyr::unnest(tidy) %>%
  dplyr::select(model_id, RQ, species, response_type, comparison, time_frame,
                term, estimate, std.error, statistic, p.value, conf.low, conf.high) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 3)))

DT::datatable(
  tidy_all,
  filter = "top",
  options = list(pageLength = 30, dom = 'Bfrtip', buttons = c('csv', 'excel')),
  extensions = 'Buttons',
  rownames = FALSE
) |>
  DT::formatRound(columns = c("estimate","std.error","statistic","p.value","conf.low","conf.high"), digits = 3)

```

## Full fixed-effects table
```{r}
#| label: tbl-models-summary-full
#| include: true
#| tbl-cap: "Full fixed-effect estimates (β, SE, 95% CI, p-value) for every term in every model currently in model_registry. Unlike the interaction-only summary above, this surfaces covariate effects (e.g. Substrate_index_after, Season) directly rather than requiring a one-off check per model."

tidy_all_full <- model_registry %>%
  dplyr::mutate(tidy = purrr::map(model, function(m) {
    if (is.null(m)) return(NULL)
    broom.mixed::tidy(m, effects = "fixed", conf.int = TRUE)
  })) %>%
  dplyr::filter(!purrr::map_lgl(tidy, is.null)) %>%
  tidyr::unnest(tidy) %>%
  # p_adj_BH only applies to each model's own term of interest — NA elsewhere.
  dplyr::mutate(p_adj_BH = dplyr::if_else(term == term_of_interest, p_adj_BH, NA_real_)) %>%
  dplyr::select(model_id, RQ, species, response_type, comparison, time_frame,
                term, estimate, std.error, statistic, p.value, p_adj_BH, conf.low, conf.high) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 3)))

write.csv(tidy_all_full, file.path(out_dir, "tbl-models-summary-full.csv"), row.names = FALSE)

DT::datatable(
  tidy_all_full,
  filter = "top",
  options = list(pageLength = 51, dom = 'Bfrtip', buttons = c('csv', 'excel')),
  extensions = 'Buttons',
  rownames = FALSE
) |>
  DT::formatRound(columns = c("estimate","std.error","statistic","p.value","p_adj_BH","conf.low","conf.high"), digits = 3)
```

## All models summary plot
```{r}
#| label: fig-forest-plot
#| fig-width: 3
#| fig-height: 12
#| dpi: 300
#| fig-cap: "Effect sizes (β ± 95% CI) for the key term of interest of every model in model_registry (kōura and catfish, RQ1–RQ4). Filled points indicate significant effects (p < 0.05). Each panel has its own x-axis scale — β magnitudes are not directly comparable across panels because each RQ tests a different contrast (step-change, per-round slope, or simple group difference) and RQ1 occupancy is on the log-odds scale while all others use the Tweedie log link. Models with degenerate confidence intervals (non-positive-definite Hessian) are excluded."

# Built directly from tidy_all — drop models with degenerate CIs (non-PD Hessian)
forest_data <- tidy_all %>%
  dplyr::filter(is.finite(conf.low) & is.finite(conf.high),
                abs(conf.high - conf.low) < 200) %>%
  dplyr::mutate(
    comparison_label = dplyr::case_when(
      comparison == "vs_control"   ~ "Stone pile vs Control",
      comparison == "vs_reference" ~ "Stone pile vs Reference",
      comparison == "treatment"    ~ "Stone pile vs Control",
      comparison == "reference"    ~ "Ref 1 vs Ref 2",
      TRUE ~ comparison),
    model_name = paste0(model_id, " ", species, " (", response_type, ")\n(", comparison_label, ")"),
    group = dplyr::case_when(
      RQ == "RQ3"                              ~ "Convergence",
      RQ == "RQ2"                              ~ "Post-construction trend",
      RQ == "RQ4" & time_frame == "before"     ~ "Baseline",
      RQ == "RQ4" & time_frame == "after"      ~ "Equivalence",
      TRUE                                     ~ "Construction-effect"),
    significant = p.value < 0.05,
    model_name  = factor(model_name, levels = rev(unique(model_name))),
    group       = factor(group, levels = c(
      "Construction-effect",
      "Post-construction trend",
      "Convergence",
      "Baseline",
      "Equivalence")))

species_colours <- c("Kōura" = "#3d7a5e", "Catfish" = "#c9603a")
species_shapes  <- c("Kōura" = 21,       "Catfish" = 24)

forest_plot <- ggplot(forest_data,
    aes(estimate, model_name)) +
  geom_vline(xintercept = 0, linetype = "dashed",
             colour = "grey40", linewidth = 0.5) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high,
                     colour = species),
                 height = 0.25, linewidth = 0.6) +
  geom_point(aes(fill = species, shape = species),
             colour = "black", size = 3.5, stroke = 0.5) +
  geom_point(data = forest_data %>% filter(!significant),
             aes(fill = species, shape = species),
             colour = "black", size = 3.5, stroke = 0.5, alpha = 0.2) +
  scale_fill_manual(  values = species_colours) +
  scale_colour_manual(values = species_colours) +
  scale_shape_manual( values = species_shapes) +
  facet_wrap(~ group, ncol = 1, scales = "free") +
  labs(
    x      = "Effect size (β ± 95% CI)",
    y      = NULL,
    fill   = "Species",
    colour = "Species",
    shape  = "Species") +
  theme(
    strip.text         = element_text(size = 9, face = "bold"),
    axis.text.y        = element_text(size = 8),
    legend.position    = "bottom",
    panel.grid.major.x = element_line(colour = "grey92", linewidth = 0.3))

ggsave(file.path(out_dir, "fig-forest-plot.png"), forest_plot, width = 4, height = 12, dpi = 300)

forest_plot
```

# Additional and sensitivity models
## Check various additional models
```{r}
#| label: additional-models

# test the effect of koura length.
m_size <- lmerTest::lmer(
  Mean_Length_Kōura ~  site * monitoring_int + (1 | site_id),
  data = all_data %>% filter(!is.na(Mean_Length_Kōura)))

# Data-quality sensitivity check
quality_sensitivity_check <- function(results, data_obj) {
  purrr::pmap_dfr(results, function(model_id, term_of_interest, model, ...) {
    if (is.null(model)) return(NULL)
    dat_sens <- data_obj %>% dplyr::filter(monitoring_quality == "complete")
    m_sens <- tryCatch(update(model, data = dat_sens), error = function(e) NULL)
    orig <- broom.mixed::tidy(model, effects = "fixed") %>%
      dplyr::filter(term == term_of_interest) %>%
      dplyr::mutate(model_id = model_id, version = "All data")
    if (is.null(m_sens)) return(orig)
    sens <- broom.mixed::tidy(m_sens, effects = "fixed") %>%
      dplyr::filter(term == term_of_interest)
    if (nrow(sens) == 0 || is.na(sens$std.error) || sens$std.error <= 0) return(orig)
    sens <- sens %>% dplyr::mutate(model_id = model_id, version = "Complete quality only")
    dplyr::bind_rows(orig, sens)
  })
}

quality_sensitivity_all <- dplyr::bind_rows(
  quality_sensitivity_check(rq1_results, baci_data),
  quality_sensitivity_check(rq2_results, baci_data_after),
  quality_sensitivity_check(rq3_results, convergence_data)) %>%
  dplyr::select(model_id, version, term, estimate, std.error, statistic, p.value) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 3)))

knitr::kable(quality_sensitivity_all)

# Matched-season check, generalized to all 6 RQ1 models: baseline
# (monitoring 0, Pre/Summer 2025) vs. only the following Summer 2026 round
# (monitoring 4), instead of pooling all 5 post-construction rounds into
# "After". Only meaningful for RQ1 (the full-timeline step-change design) —
# RQ2 already excludes round 0, and RQ3 compares against Reference, not
# Control. Simplified fixed-effects-only formula: with only 2 time points
# (n = 20), the full RQ1 formula (random effects + Temperature) fails to
# converge (non-positive-definite Hessian / singular fit).
baci_data_summer <- baci_data %>%
  dplyr::filter(monitoring_int %in% c(0, 4)) %>%
  dplyr::mutate(Period = factor(Period, levels = c("Before", "After")))

matched_season_check <- function(species, response_type) {
  response_col <- dplyr::case_when(
    response_type == "occupancy" ~ paste0("Presence_", species),
    response_type == "abundance" ~ paste0("Weighted_CPUE_", species),
    response_type == "biomass"   ~ paste0("Weighted_BPUE_", species))
  if (!response_col %in% names(baci_data_summer)) return(NULL)
  fam_obj <- if (response_type == "occupancy") binomial() else tweedie()
  form <- paste0("`", response_col, "` ~ Period * site")
  m <- tryCatch(glmmTMB::glmmTMB(as.formula(form), data = baci_data_summer, family = fam_obj),
                error = function(e) NULL)
  if (is.null(m)) return(NULL)
  broom.mixed::tidy(m, effects = "fixed") %>%
    dplyr::filter(term == "PeriodAfter:siteStone_pile") %>%
    dplyr::mutate(species = species, response_type = response_type)
}

matched_season_all <- purrr::map2_dfr(rq1_results$species, rq1_results$response_type, matched_season_check) %>%
  dplyr::select(species, response_type, term, estimate, std.error, statistic, p.value) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 3)))

knitr::kable(matched_season_all)

# Substrate necessity check, generalized to every RQ2/RQ3 model: what
# happens to the term of interest if the substrate covariate is dropped
# entirely? RQ1 has no substrate term, so it's excluded here.
substrate_drop_check <- function(results, drop_term) {
  purrr::pmap_dfr(results, function(model_id, term_of_interest, model, ...) {
    if (is.null(model)) return(NULL)
    m_drop <- tryCatch(update(model, as.formula(paste(". ~ . -", drop_term))), error = function(e) NULL)
    orig <- broom.mixed::tidy(model, effects = "fixed") %>%
      dplyr::filter(term %in% c(term_of_interest, drop_term)) %>%
      dplyr::mutate(model_id = model_id, version = paste("With", drop_term))
    if (is.null(m_drop)) return(orig)
    dropped <- broom.mixed::tidy(m_drop, effects = "fixed") %>%
      dplyr::filter(term == term_of_interest) %>%
      dplyr::mutate(model_id = model_id, version = paste("Without", drop_term))
    dplyr::bind_rows(orig, dropped)
  })
}

substrate_necessity_all <- dplyr::bind_rows(
  substrate_drop_check(rq2_results, "Substrate_index_before"),
  substrate_drop_check(rq3_results, "Substrate_index_after")) %>%
  dplyr::select(model_id, version, term, estimate, std.error, statistic, p.value) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 3)))

knitr::kable(substrate_necessity_all)

# Validate Season vs Temperature, generalized to all 4 RQ3 models:
# Temperature (a single linear predictor) previously failed DHARMa's
# residual-quantile test because kōura/catfish abundance cycles
# non-monotonically over the year — Season (categorical) can capture
# that, Temperature can't. Direct AICc comparison, swapping one for the
# other in the same base model. Only relevant to RQ3 — RQ1/RQ2 already
# use Temperature, not Season.
season_vs_temperature <- purrr::pmap_dfr(rq3_results, function(model_id, model, ...) {
  if (is.null(model)) return(NULL)
  m_temp <- tryCatch(update(model, . ~ . - Season + Temperature), error = function(e) NULL)
  tibble::tibble(
    model_id         = model_id,
    Season_AICc      = MuMIn::AICc(model),
    Temperature_AICc = if (is.null(m_temp)) NA_real_ else MuMIn::AICc(m_temp))
}) %>%
  dplyr::mutate(delta_AICc = Temperature_AICc - Season_AICc)

knitr::kable(season_vs_temperature, digits = 1)

```

### RQ4 after-period covariate selection (Temperature vs Season vs neither)
```{r}
#| label: tbl-rq4-covariate-selection
#| include: true
#| tbl-cap: "Covariate selection for all 8 RQ4 after-period models: Temperature (1 parameter), Season (3 parameters, categorical), or neither. REML = FALSE for AICc comparison. ΔAICc is relative to the best candidate per model (lowest AICc = 0). Models with non-positive-definite Hessian are flagged."

hessian_ok <- function(m) {
  if (is.null(m) || m$fit$convergence != 0) return(FALSE)
  tryCatch({
    v <- diag(as.matrix(vcov(m, full = TRUE)))
    all(is.finite(v) & v > 0)
  }, error = function(e) FALSE, warning = function(w) FALSE)
}

rq4_after_sets <- list(treatment = rq4_trt_after, reference = rq4_ref_after)

covariate_compare_one <- function(comparison, species, response_type) {
  data         <- rq4_after_sets[[comparison]]
  response_col <- if (response_type == "abundance") paste0("Weighted_CPUE_", species)
                  else paste0("Weighted_BPUE_", species)
  if (!response_col %in% names(data)) return(NULL)

  prefix   <- if (comparison == "treatment") "M4TA" else "M4RA"
  model_id <- paste0(prefix, response_letter_map[response_type], species_letter_map[species])
  zif      <- if (species == "Catfish" && response_type == "abundance") ~1 else ~0
  base     <- paste0("`", response_col, "` ~ Site_type_detail * monitoring_int")

  candidates <- list(
    Temperature = paste0(base, " + Temperature + (1 | pair_id / site_id)"),
    Season      = paste0(base, " + Season      + (1 | pair_id / site_id)"),
    Neither     = paste0(base,                  " + (1 | pair_id / site_id)"))

  fit_m <- function(form) tryCatch(
    glmmTMB::glmmTMB(as.formula(form), data = data, REML = FALSE, family = tweedie(),
      ziformula = zif, control = glmmTMBControl(optCtrl = list(iter.max = 1000, eval.max = 2000))),
    error = function(e) NULL)

  models <- purrr::map(candidates, fit_m)
  aiccs  <- purrr::map_dbl(models, ~ if (is.null(.x)) NA_real_ else MuMIn::AICc(.x))
  best   <- min(aiccs, na.rm = TRUE)

  tibble::tibble(
    model_id   = model_id,
    species    = species,
    response   = response_type,
    comparison = comparison,
    covariate  = names(candidates),
    conv       = purrr::map_int(models, ~ if (is.null(.x)) NA_integer_ else .x$fit$convergence),
    hess_ok    = purrr::map_lgl(models, hessian_ok),
    K          = purrr::map_dbl(models, ~ if (is.null(.x)) NA_real_ else attr(logLik(.x), "df")),
    AICc       = round(aiccs, 1),
    delta_AICc = round(aiccs - best, 2))
}

rq4_covariate_selection <- purrr::pmap_dfr(
  tidyr::crossing(
    tibble::tibble(comparison = c("treatment", "reference")),
    tidyr::crossing(species = all_species, response_type = c("abundance", "biomass"))),
  ~ covariate_compare_one(..1, ..2, ..3))

knitr::kable(rq4_covariate_selection)
```

### Colonisation drivers — what explains the colonisation pattern?
```{r}
#| label: tbl-aic-m3ak-selection
#| include: true
#| tbl-cap: "Model selection results for candidate drivers of kōura colonisation. Models are compared to a base model with only site, monitoring_int, season, and substrate index as predictors. ΔAICc values are relative to the best model (lowest AICc); models with ΔAICc < 2 are considered competitive. The quadratic trajectory model tests whether kōura colonisation follows a non-linear pattern over time."

M3ak_complete_vars <- c("Weighted_CPUE_Kōura",
 "Temperature", "DO_mgl", "pH", "Specific_conductivity",
 "Substrate_index_after", "lunar_illumination",  "Season",
 "log_cpue_catfish")

convergence_data_complete <- convergence_data %>%
  filter(dplyr::if_all(dplyr::all_of(M3ak_complete_vars), ~ !is.na(.)))

# Refit base model with ML for comparison
M3ak_base <- update(M3ak, data = convergence_data_complete, REML = FALSE)

# Linear vs quadratic trajectory
M3ak_quad <- update(M3ak_base, . ~ . - monitoring_int - site:monitoring_int + poly(monitoring_int, 2) * site)
anova(M3ak_base, M3ak_quad)


# Candidate predictors to test against the base model
M3ak_candidates <- list(
  Temperature = "Temperature",
  DO       = "DO_mgl",
  pH       = "pH",
  Conductivity = "Specific_conductivity",
  Lunar    = "lunar_illumination",
  Catfish  = "log_cpue_catfish",
  Global   = c("lunar_illumination", "log_cpue_catfish"))

M3ak_models <- purrr::map(M3ak_candidates, function(terms) {
  update(M3ak_base, as.formula(paste(". ~ . +", paste(terms, collapse = " + "))))
})
M3ak_models <- c(list(Base = M3ak_base), M3ak_models)

# AICc comparison
M3ak_aicc_table <- tibble(
  Model = names(M3ak_models),
  K     = purrr::map_dbl(M3ak_models, ~ attr(logLik(.x), "df")),
  AICc  = purrr::map_dbl(M3ak_models, MuMIn::AICc)) %>%
  mutate(delta_AICc = AICc - min(AICc)) %>%
  arrange(delta_AICc)

# Readable label for each entry in M3ak_candidates
M3ak_model_labels <- c(
  Base     = "Base (site × time + Season + Substrate index)",
  Temperature = "Base + Temperature",
  DO       = "Base + Dissolved oxygen",
  pH       = "Base + pH",
  Conductivity = "Base + Specific conductivity",
  Lunar    = "Base + Lunar illumination",
  Catfish  = "Base + Catfish CPUE",
  Season   = "Base + Season",
  Global   = "Base + Lunar + Catfish CPUE")

aicc_table_final <- M3ak_aicc_table %>%
  mutate(
    Model      = dplyr::recode(Model, !!!M3ak_model_labels),
    AICc       = round(AICc, 1),
    delta_AICc = round(delta_AICc, 2)) %>%
  select(Model, K, AICc, delta_AICc) %>%
  rename(`ΔAICc` = delta_AICc)

knitr::kable(aicc_table_final)

#write.csv(aicc_table_final,  file.path(out_dir, "tbl-aic-comparison.csv"), row.names = FALSE)
```

# Session info
```{r}
#| label: session-info
#| include: true
#| code-fold: true
sessionInfo()
```