---
title: "Kōura two-method occupancy analysis — Lake Rotoiti stone-pile BACI"
format: html
execute:
  message: false
  warning: false
---

# Setup
```{r}
#| label: setup
#| cache: false
#knitr::opts_chunk$set(fig.path = "occupancy_analysis_files/figure-html/")
library(brms)
library(cmdstanr)
library(tidyverse)
library(tidybayes)
library(posterior)
library(bayesplot)
library(patchwork)


options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

base_theme <- theme_classic() +
  theme(
    text         = element_text(family = "Arial", size = 11),
    axis.title   = element_text(face = "plain"),
    axis.text    = element_text(face = "plain"),
    plot.title   = element_text(face = "plain"),
    strip.text   = element_text(face = "plain"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.3)
  )
theme_set(base_theme)

der_dir <- file.path("data", "derived")
dir.create(der_dir, showWarnings = FALSE, recursive = TRUE)

location_levels <- c("Stone_pile", "Control", "Reference1", "Reference2")
period_levels   <- c("Before", "After")
season_levels   <- c("Summer", "Autumn", "Winter", "Spring")
block_levels    <- c("Impact", "Reference")

round_lookup <- tibble(
  round  = 0:6,
  period = c("Before", rep("After", 6)),
  season = c("Summer", "Autumn", "Winter", "Spring", "Summer", "Autumn", "Winter"),
  year   = c(1, 1, 1, 1, 2, 2, 2)
) |>
  mutate(
    period = factor(period, levels = period_levels),
    season = factor(season, levels = season_levels)
  )

print(round_lookup)
```

# Custom two-method occupancy family

Two detection methods deployed at every site every round:

- **Method A** — Box traps (~32 % detection probability)
- **Method B** — Fyke nets (~63 % detection probability)

The custom brms/Stan family marginalises the latent occupancy state *z* over both
methods jointly, and tolerates partial missingness (one trap lost to storms) via
four `vint()` helper columns per occasion: `yA_obs`, `yB_obs` (trap retrieved?),
`yA`, `yB` (kōura detected?).

```{r}
#| label: occupancy-family
stan_funs_occ <- "
  real occ2_lpmf(int y, real psi, real pA, real pB,
                  int yA_obs, int yB_obs, int yA, int yB) {
    real ll_given_occ = 0;
    int any_detect = 0;

    if (yA_obs == 1) {
      ll_given_occ += yA * log(pA) + (1 - yA) * log1m(pA);
      if (yA == 1) any_detect = 1;
    }
    if (yB_obs == 1) {
      ll_given_occ += yB * log(pB) + (1 - yB) * log1m(pB);
      if (yB == 1) any_detect = 1;
    }

    if (any_detect == 1) {
      return log(psi) + ll_given_occ;
    } else {
      return log_sum_exp(log(psi) + ll_given_occ, log1m(psi));
    }
  }
"

stanvars_occ <- brms::stanvar(scode = stan_funs_occ, block = "functions")

# brms custom family — dpars must start with "mu" (brms >= 2.21)
# mu maps to psi (true occupancy); pA and pB are detection submodels
# vars must use brms internal vint names matching vint() order in formula
occ2 <- custom_family(
  name  = "occ2",
  dpars = c("mu", "pA", "pB"),
  links = c("logit", "logit", "logit"),
  lb    = c(0, 0, 0),
  ub    = c(1, 1, 1),
  type  = "int",
  vars  = c("vint1[n]", "vint2[n]", "vint3[n]", "vint4[n]")
)

log_lik_occ2 <- function(i, prep) {
  psi    <- brms::get_dpar(prep, "mu", i = i)
  pA     <- brms::get_dpar(prep, "pA", i = i)
  pB     <- brms::get_dpar(prep, "pB", i = i)
  yA_obs <- prep$data$vint1[i]; yB_obs <- prep$data$vint2[i]
  yA     <- prep$data$vint3[i]; yB     <- prep$data$vint4[i]
  ll_given_occ <- 0; any_detect <- 0
  if (yA_obs == 1) {
    ll_given_occ <- ll_given_occ + yA * log(pA) + (1 - yA) * log1p(-pA)
    if (yA == 1) any_detect <- 1
  }
  if (yB_obs == 1) {
    ll_given_occ <- ll_given_occ + yB * log(pB) + (1 - yB) * log1p(-pB)
    if (yB == 1) any_detect <- 1
  }
  if (any_detect == 1) log(psi) + ll_given_occ
  else log(psi * exp(ll_given_occ) + (1 - psi))
}

posterior_predict_occ2 <- function(i, prep, ...) {
  psi    <- brms::get_dpar(prep, "mu", i = i)
  pA     <- brms::get_dpar(prep, "pA", i = i)
  pB     <- brms::get_dpar(prep, "pB", i = i)
  yA_obs <- prep$data$vint1[i]; yB_obs <- prep$data$vint2[i]
  z      <- rbinom(length(psi), 1, psi)
  yA_sim <- ifelse(yA_obs == 1, rbinom(length(psi), 1, z * pA), NA_integer_)
  yB_sim <- ifelse(yB_obs == 1, rbinom(length(psi), 1, z * pB), NA_integer_)
  rowSums(cbind(yA_sim, yB_sim), na.rm = TRUE)
}

# posterior_epred returns psi (true occupancy), not detection-weighted catch
posterior_epred_occ2 <- function(prep) {
  brms::get_dpar(prep, "mu")
}

cat("Custom family 'occ2' defined.\n")
```

# Import and prepare data
```{r}
#| label: import-data
model_data_path <- file.path(der_dir, "model_data.rds")

if (!file.exists(model_data_path)) {

  combined <- readr::read_csv(file.path("data", "derived", "Combined_data.csv"),
                               show_col_types = FALSE)
  fish     <- readr::read_csv(file.path("data", "derived", "Fish_data.csv"),
                               show_col_types = FALSE)

  backbone <- combined |>
    dplyr::transmute(
      site_id            = as.integer(site_id),
      round              = as.integer(monitoring),
      monitoring_id      = monitoring_id,
      location_type      = dplyr::case_when(
        Site_type_detail == "Stone_pile"  ~ "Stone_pile",
        Site_type_detail == "Control"     ~ "Control",
        Site_type_detail == "Reference_1" ~ "Reference1",
        Site_type_detail == "Reference_2" ~ "Reference2",
        TRUE                              ~ NA_character_),
      block              = dplyr::case_when(
        Site_type_detail %in% c("Stone_pile", "Control")      ~ "Impact",
        Site_type_detail %in% c("Reference_1", "Reference_2") ~ "Reference",
        TRUE                                                    ~ NA_character_),
      temperature        = as.numeric(Temperature),
      monitoring_quality = as.character(monitoring_quality),
      pair_id            = pair_id
    ) |>
    dplyr::filter(!is.na(location_type))

  # One row per kōura individual → collapse to presence per site×round×method
  koura_by_method <- fish |>
    dplyr::filter(species == "Kōura") |>
    dplyr::group_by(site_id, monitoring_id, net_type) |>
    dplyr::summarise(detected = 1L, .groups = "drop") |>
    dplyr::mutate(site_id = as.integer(site_id))

  # Cross all valid site×round combinations with both methods; fill 0 for no-catch
  all_combos <- backbone |>
    dplyr::select(site_id, monitoring_id) |>
    tidyr::crossing(net_type = c("Box", "Fyke"))

  detection_wide <- all_combos |>
    dplyr::left_join(
      koura_by_method |> dplyr::select(site_id, monitoring_id, net_type, detected),
      by = c("site_id", "monitoring_id", "net_type")) |>
    dplyr::mutate(detected = tidyr::replace_na(detected, 0L)) |>
    tidyr::pivot_wider(names_from = net_type, values_from = detected) |>
    dplyr::rename(yA = Box, yB = Fyke)

  real_data <- backbone |>
    dplyr::left_join(detection_wide, by = c("site_id", "monitoring_id")) |>
    dplyr::mutate(
      # monitoring_quality == "complete" → both traps usable
      usable = monitoring_quality == "complete",
      yA_obs = as.integer(usable),
      yB_obs = as.integer(usable),
      yA     = dplyr::if_else(usable, yA, NA_integer_),
      yB     = dplyr::if_else(usable, yB, NA_integer_)
    ) |>
    dplyr::filter(!(yA_obs == 0 & yB_obs == 0)) |>   # drop zero-info occasions
    dplyr::left_join(round_lookup, by = "round") |>
    dplyr::mutate(
      location_type = factor(location_type, levels = location_levels),
      block         = factor(block, levels = block_levels),
      location_id   = paste(site_id, location_type, sep = "_"),
      temp_z        = as.numeric(scale(temperature))  # global standardisation
    ) |>
    dplyr::select(site_id, block, location_type, location_id, round,
                  period, season, year, temperature, temp_z,
                  yA_obs, yB_obs, yA, yB, pair_id, monitoring_quality)

  # Stan requires no NAs: replace NA detection with 0 (yA_obs/yB_obs flags
  # tell Stan to ignore the value when the trap was lost)
  model_data <- real_data |>
    dplyr::mutate(
      y_dummy = 0L,
      yA = dplyr::if_else(is.na(yA), 0L, yA),
      yB = dplyr::if_else(is.na(yB), 0L, yB)
    )

  # Validation
  stopifnot(
    all(model_data$yA %in% c(0L, 1L)),
    all(model_data$yB %in% c(0L, 1L)),
    all(model_data$round %in% 0:6),
    all(model_data$yA_obs == 1 | model_data$yB_obs == 1)
  )

  saveRDS(model_data, model_data_path)
  cat("Imported and saved:", nrow(model_data), "occasions\n")

} else {
  model_data <- readRDS(model_data_path)
  cat("Loaded cached model data:", nrow(model_data), "occasions\n")
}

cat("\nOccasions by location × period × season:\n")
print(dplyr::count(model_data, location_type, period, season))

cat("\nKōura detection rates by method:\n")
cat("  Box trap (yA):", round(mean(model_data$yA), 3), "\n")
cat("  Fyke net (yB):", round(mean(model_data$yB), 3), "\n")
```

# Priors

Weakly informative normal(0, 1.5) priors on log-odds scale for fixed effects
(allows ±3 units ≈ occupancy/detection probability ranging ~0.05–0.95).
Normal(0, 1) on random-effect SDs.

```{r}
#| label: priors
priors_occ <- c(
  prior(normal(0, 1.5), class = "b"),
  prior(normal(0, 1.5), class = "b", dpar = "pA"),
  prior(normal(0, 1.5), class = "b", dpar = "pB"),
  prior(normal(0, 1),   class = "sd"),
  prior(normal(0, 1),   class = "sd", dpar = "pA"),
  prior(normal(0, 1),   class = "sd", dpar = "pB")
)
```

# Model 1: Full model (all seasons)

Primary model using all 128 occasions. Tests `location_type × period × season`
interaction. Provides whole-study headline occupancy estimates and uses the full
data to estimate variance components (site random effects, detection–temperature
slope) most precisely.

```{r}
#| label: fit-full
full_path <- file.path(der_dir, "fit_full.rds")

if (!file.exists(full_path)) {

  bf_full <- bf(
    y_dummy | vint(yA_obs, yB_obs, yA, yB) ~
      location_type * period * season + (1 | site_id),
    pA ~ 1 + temp_z + (1 | site_id),
    pB ~ 1 + temp_z + (1 | site_id),
    family = occ2
  )

  fit_full <- brm(
    bf_full,
    data     = model_data,
    prior    = priors_occ,
    stanvars = stanvars_occ,
    chains   = 4, iter = 4000, warmup = 1500,
    seed     = 2026,
    control  = list(adapt_delta = 0.97, max_treedepth = 13)
  )

  saveRDS(fit_full, full_path)
  cat("Fitted and saved full model.\n")

} else {
  fit_full <- readRDS(full_path)
  cat("Loaded cached full model.\n")
}

summary(fit_full)
cat("\nMax Rhat:   ", max(brms::rhat(fit_full), na.rm = TRUE), "\n")
cat("Min ESS:    ", min(posterior::summarise_draws(fit_full)$ess_bulk, na.rm = TRUE), "\n")
cat("Divergences:", sum(brms::nuts_params(fit_full, pars = "divergent__")$Value), "\n")
```

```{r}
#| label: traceplots-full
#| include: false
#| cache: false
bayesplot::mcmc_trace(fit_full, regex_pars = "^b_") +
  labs(title = "Full model — fixed-effect trace plots")
```

# Model 2: Summer-only model

Confirmatory model restricted to Summer rounds (R0 Before, R4 After). Targets
the a-priori hypothesis that refuge value is greatest in summer when catfish
pressure and thermal stratification are highest. Treated as a robustness check
on the summer cells of the full model, not a replacement for it.

```{r}
#| label: fit-summer
summer_path <- file.path(der_dir, "fit_summer.rds")
summer_data <- model_data |> dplyr::filter(season == "Summer")

cat("Summer subset:", nrow(summer_data), "occasions\n")
print(dplyr::count(summer_data, location_type, period))

if (!file.exists(summer_path)) {

  bf_summer <- bf(
    y_dummy | vint(yA_obs, yB_obs, yA, yB) ~
      location_type * period + (1 | site_id),
    pA ~ 1 + temp_z + (1 | site_id),
    pB ~ 1 + temp_z + (1 | site_id),
    family = occ2
  )

  fit_summer <- brm(
    bf_summer,
    data     = summer_data,
    prior    = priors_occ,
    stanvars = stanvars_occ,
    chains   = 4, iter = 4000, warmup = 1500,
    seed     = 2026,
    control  = list(adapt_delta = 0.97, max_treedepth = 13)
  )

  saveRDS(fit_summer, summer_path)
  cat("Fitted and saved summer model.\n")

} else {
  fit_summer <- readRDS(summer_path)
  cat("Loaded cached summer model.\n")
}

summary(fit_summer)
cat("\nMax Rhat:   ", max(brms::rhat(fit_summer), na.rm = TRUE), "\n")
cat("Min ESS:    ", min(posterior::summarise_draws(fit_summer)$ess_bulk, na.rm = TRUE), "\n")
cat("Divergences:", sum(brms::nuts_params(fit_summer, pars = "divergent__")$Value), "\n")
```

```{r}
#| label: traceplots-summer
#| include: false
#| cache: false
bayesplot::mcmc_trace(fit_summer, regex_pars = "^b_") +
  labs(title = "Summer model — fixed-effect trace plots")
```

# Model 3: Warm-season model (Summer + Autumn)

Ecological rationale: thermal stratification in Lake Rotoiti peaks in summer and
persists through autumn, compressing kōura into the oxygenated nearshore zone
where stone piles sit. Catfish predation pressure is also highest in these two
seasons. This model tests the restoration signal in the combined warm season
(76 occasions) with more power than summer-only (40 occasions).

Season is included as an additive covariate — not a full three-way interaction —
because the data support estimating seasonal baseline differences in occupancy
but not the full `location_type × period × season` interaction in this subset.

```{r}
#| label: fit-warm
warm_path <- file.path(der_dir, "fit_warm.rds")
warm_data <- model_data |>
  dplyr::filter(season %in% c("Summer", "Autumn")) |>
  dplyr::mutate(season = droplevels(season))

cat("Warm subset:", nrow(warm_data), "occasions\n")
print(dplyr::count(warm_data, location_type, period, season))

if (!file.exists(warm_path)) {

  bf_warm <- bf(
    y_dummy | vint(yA_obs, yB_obs, yA, yB) ~
      location_type * period + season + (1 | site_id),
    pA ~ 1 + temp_z + (1 | site_id),
    pB ~ 1 + temp_z + (1 | site_id),
    family = occ2
  )

  fit_warm <- brm(
    bf_warm,
    data     = warm_data,
    prior    = priors_occ,
    stanvars = stanvars_occ,
    chains   = 4, iter = 4000, warmup = 1500,
    seed     = 2026,
    control  = list(adapt_delta = 0.99, max_treedepth = 13)
  )

  saveRDS(fit_warm, warm_path)
  cat("Fitted and saved warm model.\n")

} else {
  fit_warm <- readRDS(warm_path)
  cat("Loaded cached warm model.\n")
}

summary(fit_warm)
cat("\nMax Rhat:   ", max(brms::rhat(fit_warm), na.rm = TRUE), "\n")
cat("Min ESS:    ", min(posterior::summarise_draws(fit_warm)$ess_bulk, na.rm = TRUE), "\n")
cat("Divergences:", sum(brms::nuts_params(fit_warm, pars = "divergent__")$Value), "\n")
```

```{r}
#| label: traceplots-warm
#| include: false
#| cache: false
bayesplot::mcmc_trace(fit_warm, regex_pars = "^b_") +
  labs(title = "Warm model — fixed-effect trace plots")
```

# Extract posterior contrasts plots

Four contrasts of ecological interest, evaluated in the After period:

1. **Stone_pile − Control** — did rocks help, relative to the paired control?
2. **Reference1 − Reference2** — spatial noise floor (how similar are reference sites?)
3. **Control − Reference (mean)** — how far below reference quality is the control?
4. **Stone_pile − Reference (mean)** — how close to optimal did stone piles get?

Plus the **BACI double-difference**: `(Stone_pile After−Before) − (Control After−Before)` —
the most defensible causal estimate, as it removes any pre-existing gap between site types.

```{r}
#| label: contrast-functions
summarise_draws_vec <- function(x, label) {
  tibble(
    contrast     = label,
    mean         = mean(x),
    median       = median(x),
    lower95      = quantile(x, 0.025),
    upper95      = quantile(x, 0.975),
    p_direction  = mean(x > 0)
  )
}

get_psi_draws <- function(fit, model_data, include_season = TRUE) {
  grid_vars <- if (include_season) {
    model_data |> dplyr::distinct(location_type, period, season)
  } else {
    model_data |> dplyr::distinct(location_type, period)
  }
  newdata <- grid_vars |>
    dplyr::mutate(
      site_id  = model_data$site_id[1],
      temp_z   = 0, yA_obs = 1, yB_obs = 1, yA = 0, yB = 0, y_dummy = 0
    )
  draws <- posterior_epred(fit, newdata = newdata, re_formula = NA, dpar = "mu")
  colnames(draws) <- if (include_season) {
    paste(newdata$location_type, newdata$period, newdata$season, sep = "|")
  } else {
    paste(newdata$location_type, newdata$period, sep = "|")
  }
  list(draws = draws, grid = newdata)
}

build_contrasts <- function(draws, grid, period_val, season_val = NULL) {
  key <- function(loc) {
    if (is.null(season_val)) paste(loc, period_val, sep = "|")
    else paste(loc, period_val, season_val, sep = "|")
  }
  d_add      <- draws[, key("Stone_pile")]
  d_ctrl     <- draws[, key("Control")]
  d_ref1     <- draws[, key("Reference1")]
  d_ref2     <- draws[, key("Reference2")]
  d_ref_mean <- (d_ref1 + d_ref2) / 2
  bind_rows(
    summarise_draws_vec(d_add  - d_ctrl,      "Stone_pile - Control"),
    summarise_draws_vec(d_ref1 - d_ref2,      "Reference1 - Reference2"),
    summarise_draws_vec(d_ctrl - d_ref_mean,  "Control - Reference (mean)"),
    summarise_draws_vec(d_add  - d_ref_mean,  "Stone_pile - Reference (mean)")
  ) |>
    dplyr::mutate(period = period_val,
                  season = if (is.null(season_val)) NA_character_ else season_val)
}

build_baci <- function(draws, include_season, season_val = NULL) {
  key <- function(loc, period) {
    if (include_season) paste(loc, period, season_val, sep = "|")
    else paste(loc, period, sep = "|")
  }
  d <- (draws[, key("Stone_pile", "After")] - draws[, key("Stone_pile", "Before")]) -
       (draws[, key("Control",    "After")] - draws[, key("Control",    "Before")])
  summarise_draws_vec(d, "BACI: (Stone_pile After-Before) - (Control After-Before)") |>
    dplyr::mutate(period = "Before-vs-After",
                  season = if (include_season) season_val else NA_character_)
}

run_contrasts <- function(fit, model_data, model_label, include_season = TRUE) {
  psi_out <- get_psi_draws(fit, model_data, include_season)
  draws   <- psi_out$draws
  grid    <- psi_out$grid

  cell_summary <- tibble(
    location_type = grid$location_type,
    period        = grid$period,
    season        = if (include_season) grid$season else NA
  ) |>
    dplyr::mutate(
      psi_mean = apply(draws, 2, mean),
      psi_lwr  = apply(draws, 2, quantile, 0.025),
      psi_upr  = apply(draws, 2, quantile, 0.975)
    )

  if (include_season) {
    seasons_present      <- unique(grid$season)
    contrasts_after      <- purrr::map_dfr(seasons_present,
                              ~ build_contrasts(draws, grid, "After", .x))
    seasons_with_before  <- grid |>
      dplyr::filter(period == "Before") |> dplyr::pull(season) |> unique()
    baci <- purrr::map_dfr(seasons_with_before,
               ~ build_baci(draws, include_season = TRUE, season_val = .x))
  } else {
    contrasts_after <- build_contrasts(draws, grid, "After")
    baci            <- build_baci(draws, include_season = FALSE)
  }

  contrasts_all <- dplyr::bind_rows(contrasts_after, baci) |>
    dplyr::mutate(model = model_label)

  readr::write_csv(cell_summary,
    file.path(der_dir, paste0(model_label, "_psi_estimates.csv")))
  readr::write_csv(contrasts_all,
    file.path(der_dir, paste0(model_label, "_contrasts.csv")))

  # Shared display helpers — colours and shapes match analysis.qmd
  site_labels  <- c(Stone_pile = "Stone pile", Control = "Control",
                    Reference1 = "Reference 1", Reference2 = "Reference 2")
  site_colours <- c(Stone_pile = "#D55E00", Control = "#E69F00",
                    Reference1 = "#0072B2", Reference2 = "#56B4E9")
  site_shapes  <- c(Stone_pile = 21, Control = 24,
                    Reference1 = 22, Reference2 = 23)
  season_colours <- c(Summer = "#D55E00", Autumn = "#E69F00",
                      Winter = "#56B4E9", Spring = "#009E73")
  season_shapes  <- c(Summer = 21, Autumn = 24, Winter = 22, Spring = 23)

  # Tidy contrast labels for display
  contrasts_plot <- contrasts_all |>
    dplyr::mutate(
      contrast = dplyr::recode(contrast,
        "Stone_pile - Control"                                         = "Stone pile vs Control",
        "Reference1 - Reference2"                                      = "Reference 1 vs Reference 2",
        "Control - Reference (mean)"                                   = "Control vs Reference",
        "Stone_pile - Reference (mean)"                                = "Stone pile vs Reference",
        "BACI: (Stone_pile After-Before) - (Control After-Before)"     = "BACI (Stone pile − Control)",
        "BACI interaction: (Stone_pile After-Before) - (Control After-Before)" = "BACI (Stone pile − Control)"
      ),
      period = dplyr::recode(period,
        "After"          = "After treatment",
        "Before-vs-After"= "BACI"
      )
    )

  # ---- Plot 1: occupancy probability ----------------------------------------
  # Only draw connecting lines for location × season groups that have
  # both Before and After estimates (avoids geom_line warning in Autumn panel)
  lines_data <- cell_summary |>
    dplyr::group_by(location_type, if (include_season) season else NULL) |>
    dplyr::filter(dplyr::n_distinct(period) > 1) |>
    dplyr::ungroup()

  p1 <- ggplot(cell_summary,
               aes(x = period, y = psi_mean,
                   colour = location_type, fill = location_type,
                   shape = location_type, group = location_type)) +
    geom_line(data = lines_data,
              position = position_dodge(0.35), linewidth = 0.7, alpha = 0.6) +
    geom_linerange(aes(ymin = psi_lwr, ymax = psi_upr),
                   position = position_dodge(0.35), linewidth = 0.8) +
    geom_point(colour = "black", size = 3, stroke = 0.4,
               position = position_dodge(0.35)) +
    {if (include_season) facet_wrap(~season, nrow = 1)} +
    scale_colour_manual(values = site_colours, labels = site_labels,
                        name = "Site type") +
    scale_fill_manual(values = site_colours, labels = site_labels,
                      name = "Site type") +
    scale_shape_manual(values = site_shapes, labels = site_labels,
                       name = "Site type") +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25),
                       labels = scales::percent_format(accuracy = 1)) +
    scale_x_discrete(labels = c(Before = "Before", After = "After")) +
    labs(y = "Occupancy probability (ψ)", x = NULL) +
    theme_classic(base_size = 11) +
    theme(
      text             = element_text(family = "Arial"),
      panel.border     = element_rect(colour = "black", fill = NA, linewidth = 0.3),
      panel.grid.minor = element_blank(),
      strip.background = element_rect(fill = "white", colour = "black", linewidth = 0.5),
      strip.text       = element_text(face = "plain"),
      legend.position  = "right"
    )
  ggsave(file.path("outputs", paste0("fig-occ-", model_label, "-psi.png")),
         p1, width = 10, height = 5, dpi = 300)

  # ---- Plot 2: contrast forest plot -----------------------------------------
  x_lim <- max(abs(c(contrasts_plot$lower95, contrasts_plot$upper95)),
               na.rm = TRUE) * 1.05

  shared_theme <- list(
    theme_classic(base_size = 11),
    theme(
      text               = element_text(family = "Arial"),
      panel.border       = element_rect(colour = "black", fill = NA, linewidth = 0.3),
      panel.grid.minor   = element_blank(),
      panel.grid.major.y = element_blank(),
      strip.background   = element_rect(fill = "white", colour = "black", linewidth = 0.5),
      strip.text         = element_text(face = "plain"),
      axis.text.y        = element_text(size = 10)
    )
  )
  fill_scale <- if (include_season)
    scale_fill_manual(values = season_colours, name = "Season")
  else
    scale_fill_manual(values = c("After treatment" = "grey30"), guide = "none")
  shape_scale <- if (include_season)
    scale_shape_manual(values = season_shapes, name = "Season")
  else
    scale_shape_manual(values = c("After treatment" = 21), guide = "none")
  x_scale <- scale_x_continuous(limits = c(-x_lim, x_lim),
                                 labels = scales::label_number(accuracy = 0.1))

  contrast_levels <- c(
    "BACI (Stone pile − Control)",
    "Reference 1 vs Reference 2",
    "Control vs Reference",
    "Stone pile vs Reference",
    "Stone pile vs Control"
  )

  make_forest <- function(data) {
    data <- data |>
      dplyr::mutate(contrast = factor(contrast, levels = contrast_levels))
    base_plot <- ggplot(data, aes(x = mean, y = contrast)) +
      geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50", linewidth = 0.6)
    if (include_season) {
      base_plot <- base_plot +
        geom_linerange(aes(xmin = lower95, xmax = upper95, colour = season),
                       position = position_dodge(0.6), linewidth = 0.8) +
        geom_point(aes(fill = season, shape = season),
                   position = position_dodge(0.6),
                   colour = "black", size = 3, stroke = 0.8) +
        scale_colour_manual(values = season_colours, guide = "none") +
        fill_scale + shape_scale
    } else {
      base_plot <- base_plot +
        geom_linerange(aes(xmin = lower95, xmax = upper95),
                       position = position_dodge(0.6), linewidth = 0.8,
                       colour = "grey30") +
        geom_point(position = position_dodge(0.6), shape = 21,
                   fill = "grey30", colour = "black", size = 4.5, stroke = 0.8)
    }
    base_plot + x_scale + shared_theme +
      labs(x = "Posterior contrast (probability scale)", y = NULL)
  }

  baci_data  <- contrasts_plot |> dplyr::filter(period == "BACI")
  after_data <- contrasts_plot |> dplyr::filter(period == "After treatment")

  n_baci  <- dplyr::n_distinct(baci_data$contrast)
  n_after <- dplyr::n_distinct(after_data$contrast)

  strip_theme <- theme(
    strip.text       = element_text(size = 9, face = "bold"),
    strip.background = element_rect(fill = "white", colour = "black", linewidth = 0.5))

  p_baci  <- make_forest(baci_data)  +
    facet_wrap(~ "BACI") +
    strip_theme +
    theme(legend.position = "none") +
    guides(colour = "none", fill = "none", shape = "none")
  p_after <- make_forest(after_data) +
    facet_wrap(~ "After treatment") +
    strip_theme +
    theme(legend.position = "none")

  p2 <- p_baci / p_after +
    patchwork::plot_layout(heights = c(n_baci, n_after), guides = "collect") &
    theme(legend.position  = "top",
          legend.direction  = "horizontal",
          legend.title = element_blank(),
          legend.text       = element_text(size = 9),
          legend.key.size   = unit(0.8, "lines"))

  ggsave(file.path("outputs", paste0("fig-occ-", model_label, "-contrasts.png")),
         p2, width = 6, height = 5.5, dpi = 300)

  cat("\n====", model_label, "====\n")
  print(contrasts_all |>
    dplyr::select(model, contrast, period, season, mean, lower95, upper95, p_direction))

  list(cell_summary = cell_summary, contrasts = contrasts_all,
       psi_plot = p1, contrast_plot = p2)
}
```

```{r}
#| label: run-contrasts-full
#| include: false
results_full <- run_contrasts(fit_full, model_data,
                               model_label    = "full",
                               include_season = TRUE)
```

```{r}
#| label: fig-contrasts-full
#| include: false
#| cache: false
#| fig-cap: "Full model (all seasons): occupancy probability ψ by site type and period (left) and posterior contrasts with 95% credible intervals (right)."
#| fig-width: 12
#| fig-height: 7
results_full$psi_plot + results_full$contrast_plot
```

```{r}
#| label: run-contrasts-summer
#| include: false
results_summer <- run_contrasts(fit_summer, summer_data,
                                 model_label    = "summer",
                                 include_season = FALSE)
```

```{r}
#| label: fig-contrasts-summer
#| include: false
#| cache: false
#| fig-cap: "Summer-only model: occupancy probability ψ by site type and period (left) and posterior contrasts with 95% credible intervals (right)."
#| fig-width: 12
#| fig-height: 5
results_summer$psi_plot + results_summer$contrast_plot
```

```{r}
#| label: run-contrasts-warm
#| include: false
results_warm <- run_contrasts(fit_warm, warm_data,
                               model_label    = "warm",
                               include_season = TRUE)
```

```{r}
#| label: fig-contrasts-warm
#| include: false
#| cache: false
#| fig-cap: "Warm-season model (Summer and Autumn): occupancy probability ψ by site type and period (left) and posterior contrasts with 95% credible intervals by season (right). This is the primary inferential model."
#| fig-width: 12
#| fig-height: 5
results_warm$psi_plot + results_warm$contrast_plot
```

## table psi-warm
```{r}
#| label: tbl-psi-warm
#| tbl-cap: "Posterior kōura occupancy probability ($\\psi$) from the warm-season occupancy model (Summer and Autumn rounds). Values are population-level estimates averaged over random site effects, with 95% credible intervals. Autumn has no Before period as baseline monitoring was conducted in summer only."
psi_warm_tbl <- results_warm$cell_summary |>
  dplyr::mutate(
    season      = factor(season, levels = c("Summer", "Autumn")),
    period      = factor(period, levels = c("Before", "After")),
    location_type = factor(
                      dplyr::recode(as.character(location_type),
                        Stone_pile = "Stone pile",
                        Reference1 = "Reference 1",
                        Reference2 = "Reference 2"),
                      levels = c("Stone pile", "Control", "Reference 1", "Reference 2"))) |>
  dplyr::arrange(season, period, location_type) |>
  dplyr::transmute(
    Season      = as.character(season),
    Period      = as.character(period),
    `Site type` = location_type,
    `ψ (mean [95% CrI])` = sprintf("%.2f [%.2f–%.2f]", psi_mean, psi_lwr, psi_upr)
  )

readr::write_csv(psi_warm_tbl, file.path("outputs", "tbl-psi-warm.csv"))

knitr::kable(psi_warm_tbl, align = c("l", "l", "l", "c"),
             col.names = c("Season", "Period", "Site type", "$\\psi$ (mean [95% CrI])"),
             escape = FALSE)
```

# Combined contrast summary
```{r}
#| label: tbl-combined-summary
#| tbl-cap: "Posterior contrasts across all three occupancy models (95% credible intervals)."
all_contrasts <- dplyr::bind_rows(
  results_full$contrasts,
  results_summer$contrasts,
  results_warm$contrasts
)

readr::write_csv(all_contrasts,
  file.path(der_dir, "all_models_contrasts.csv"))

all_contrasts |>
  dplyr::select(model, contrast, period, season, mean, lower95, upper95, p_direction) |>
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  knitr::kable()
```

# Additional and sensitivity models — wave exposure

Two sensitivity models test whether wave exposure confounds or modifies the
primary results from the warm-season model:

- **Model S1 (fetch-ψ)**: adds max effective fetch (site-level, static) as a
  covariate on occupancy (ψ). Tests whether more exposed sites have systematically
  different baseline occupancy independent of treatment type.

- **Model S2 (wave-p)**: adds the round-specific wave energy proxy as an
  additional detection covariate alongside temperature. Tests whether high-energy
  rounds suppress detection probability.

LOO-IC comparison with the base warm model indicates whether either covariate
improves predictive fit.

```{r}
#| label: fetch-covariates
site_fetch    <- readr::read_csv(file.path(der_dir, "site_fetch.csv"),
                                  show_col_types = FALSE)
wave_by_round <- readr::read_csv(file.path(der_dir, "wave_exposure_by_round.csv"),
                                  show_col_types = FALSE)

warm_data_fetch <- warm_data |>
  dplyr::left_join(
    site_fetch |> dplyr::select(site_id, max_eff_fetch_km),
    by = "site_id"
  ) |>
  dplyr::left_join(
    wave_by_round |> dplyr::select(site_id, round, wave_energy_proxy),
    by = c("site_id", "round")
  ) |>
  dplyr::mutate(
    fetch_z = as.numeric(scale(max_eff_fetch_km)),
    wave_z  = as.numeric(scale(wave_energy_proxy))
  )

cat("Fetch covariate range (km):", round(range(warm_data_fetch$max_eff_fetch_km), 2), "\n")
cat("Wave energy proxy range:   ", round(range(warm_data_fetch$wave_energy_proxy, na.rm = TRUE), 0), "\n")
cat("Missing wave_z:            ", sum(is.na(warm_data_fetch$wave_z)), "rows\n")
```

```{r}
#| label: fit-fetch-psi
fetch_psi_path <- file.path(der_dir, "fit_fetch_psi.rds")

if (!file.exists(fetch_psi_path)) {

  bf_fetch_psi <- bf(
    y_dummy | vint(yA_obs, yB_obs, yA, yB) ~
      location_type * period + season + fetch_z + (1 | site_id),
    pA ~ 1 + temp_z + (1 | site_id),
    pB ~ 1 + temp_z + (1 | site_id),
    family = occ2
  )

  fit_fetch_psi <- brm(
    bf_fetch_psi,
    data     = warm_data_fetch,
    prior    = priors_occ,
    stanvars = stanvars_occ,
    chains   = 4, iter = 4000, warmup = 1500,
    seed     = 2026,
    control  = list(adapt_delta = 0.99, max_treedepth = 13)
  )

  saveRDS(fit_fetch_psi, fetch_psi_path)
  cat("Fitted and saved fetch-ψ model.\n")

} else {
  fit_fetch_psi <- readRDS(fetch_psi_path)
  cat("Loaded cached fetch-ψ model.\n")
}

summary(fit_fetch_psi)
cat("\nMax Rhat:   ", max(brms::rhat(fit_fetch_psi), na.rm = TRUE), "\n")
cat("Min ESS:    ", min(posterior::summarise_draws(fit_fetch_psi)$ess_bulk, na.rm = TRUE), "\n")
cat("Divergences:", sum(brms::nuts_params(fit_fetch_psi, pars = "divergent__")$Value), "\n")
```

```{r}
#| label: fit-wave-p
wave_p_path <- file.path(der_dir, "fit_wave_p.rds")

if (!file.exists(wave_p_path)) {

  bf_wave_p <- bf(
    y_dummy | vint(yA_obs, yB_obs, yA, yB) ~
      location_type * period + season + (1 | site_id),
    pA ~ 1 + temp_z + wave_z + (1 | site_id),
    pB ~ 1 + temp_z + wave_z + (1 | site_id),
    family = occ2
  )

  fit_wave_p <- brm(
    bf_wave_p,
    data     = warm_data_fetch,
    prior    = priors_occ,
    stanvars = stanvars_occ,
    chains   = 4, iter = 4000, warmup = 1500,
    seed     = 2026,
    control  = list(adapt_delta = 0.99, max_treedepth = 13)
  )

  saveRDS(fit_wave_p, wave_p_path)
  cat("Fitted and saved wave-detection model.\n")

} else {
  fit_wave_p <- readRDS(wave_p_path)
  cat("Loaded cached wave-detection model.\n")
}

summary(fit_wave_p)
cat("\nMax Rhat:   ", max(brms::rhat(fit_wave_p), na.rm = TRUE), "\n")
cat("Min ESS:    ", min(posterior::summarise_draws(fit_wave_p)$ess_bulk, na.rm = TRUE), "\n")
cat("Divergences:", sum(brms::nuts_params(fit_wave_p, pars = "divergent__")$Value), "\n")
```

```{r}
#| label: loo-comparison
loo_warm      <- loo(fit_warm,      reloo = TRUE)
loo_fetch_psi <- loo(fit_fetch_psi, reloo = TRUE)
loo_wave_p    <- loo(fit_wave_p,    reloo = TRUE)

loo_comp <- loo_compare(loo_warm, loo_fetch_psi, loo_wave_p)
print(loo_comp, digits = 2)

readr::write_csv(
  as.data.frame(loo_comp) |> tibble::rownames_to_column("fit_name"),
  file.path(der_dir, "loo_sensitivity_comparison.csv")
)
```

```{r}
#| label: fetch-effect-plot
#| fig-cap: "Sensitivity model posterior estimates. Left: fetch-ψ model — posterior for the fetch_z coefficient on occupancy (positive = more exposed sites have higher occupancy). Right: wave-p model — posterior for the wave_z coefficient on detection probability for both trap methods (negative = high wave energy suppresses detection)."
#| fig-width: 10
#| fig-height: 4

fetch_coef <- posterior::as_draws_df(fit_fetch_psi) |>
  dplyr::select(b_fetch_z) |>
  tidyr::pivot_longer(everything(), names_to = "param", values_to = "value") |>
  dplyr::mutate(param = "fetch_z → ψ", model = "S1: fetch on ψ")

wave_coefs <- posterior::as_draws_df(fit_wave_p) |>
  dplyr::select(b_pA_wave_z, b_pB_wave_z) |>
  tidyr::pivot_longer(everything(), names_to = "param", values_to = "value") |>
  dplyr::mutate(
    param = dplyr::recode(param,
      b_pA_wave_z = "wave_z → pA (box trap)",
      b_pB_wave_z = "wave_z → pB (fyke net)"),
    model = "S2: wave on detection"
  )

all_coefs <- dplyr::bind_rows(fetch_coef, wave_coefs)

ggplot(all_coefs, aes(x = value, y = param, fill = model)) +
  ggdist::stat_halfeye(.width = c(0.5, 0.95), point_interval = "median_qi",
                       alpha = 0.85, normalize = "groups") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = c("S1: fetch on ψ"       = "#0072B2",
                               "S2: wave on detection" = "#D55E00"),
                    name = NULL) +
  labs(x = "Posterior coefficient (log-odds scale)", y = NULL) +
  theme(legend.position = "bottom")
```

# Power analysis — sites and rounds needed for statistical detection

The current study shows a consistent positive direction for the stone pile effect
but wide credible intervals due to limited site replication. This simulation-based
power analysis asks: **how many site pairs, or how many additional monitoring rounds,
would be needed to reliably detect the observed effect size?**

**Method:** Parameters (occupancy probabilities, detection rates, site-level variance)
are drawn from the posterior means of the warm-season model. For each combination
of site pairs and After rounds, 300 datasets are simulated and a logistic GLMM
is fitted as a fast surrogate. Power = proportion of simulations where the 95 % CI
of the BACI term (Stone pile × Period interaction) excludes zero.

```{r}
#| label: power-extract-params
# Extract posterior means from the warm model
fe         <- brms::fixef(fit_warm)[, "Estimate"]
draws_warm <- posterior::as_draws_df(fit_warm)

logit_psi_ctrl_before <- fe["Intercept"]
b_period_after        <- fe["periodAfter"]
b_sp                  <- fe["location_typeStone_pile"]
b_sp_after            <- fe["location_typeStone_pile:periodAfter"]
logit_pA              <- fe["pA_Intercept"]
logit_pB              <- fe["pB_Intercept"]

# Site-level random effect SD (occupancy submodel)
sigma_site <- posterior::summarise_draws(fit_warm) |>
  dplyr::filter(variable == "sd_site_id__Intercept") |>
  dplyr::pull(mean)

# Report the key parameters being used
cat("Parameters extracted from warm model posterior means:\n")
cat("  ψ Control  Before:", round(plogis(logit_psi_ctrl_before), 3), "\n")
cat("  ψ Control  After: ", round(plogis(logit_psi_ctrl_before + b_period_after), 3), "\n")
cat("  ψ Stone pile Before:", round(plogis(logit_psi_ctrl_before + b_sp), 3), "\n")
cat("  ψ Stone pile After: ", round(plogis(logit_psi_ctrl_before + b_sp + b_period_after + b_sp_after), 3), "\n")
cat("  pA (box trap): ", round(plogis(logit_pA), 3), "\n")
cat("  pB (fyke net): ", round(plogis(logit_pB), 3), "\n")
cat("  Site SD (logit): ", round(sigma_site, 3), "\n")
cat("  BACI log-odds:  ", round(b_sp_after, 3),
    "≈", round(plogis(logit_psi_ctrl_before + b_sp + b_period_after + b_sp_after) -
               plogis(logit_psi_ctrl_before + b_period_after), 3),
    "probability units\n")
```

```{r}
#| label: power-simulation
power_cache <- file.path(der_dir, "power_analysis.rds")

if (!file.exists(power_cache)) {

  # Simulate one dataset and test if the BACI term's 95% CI excludes zero
  one_sim <- function(n_pairs, n_after_rounds,
                      logit_psi_ctrl_before, b_period_after, b_sp, b_sp_after,
                      logit_pA, logit_pB, sigma_site) {

    pA <- plogis(logit_pA)
    pB <- plogis(logit_pB)
    n_before_rounds <- 1L   # matches study design: 1 Before round

    site_re <- rnorm(n_pairs * 2, 0, sigma_site)
    rows    <- vector("list", n_pairs * 2 * (n_before_rounds + n_after_rounds))
    idx     <- 0L

    for (pair in seq_len(n_pairs)) {
      for (st in c("Control", "Stone_pile")) {
        site_idx <- (pair - 1L) * 2L + if (st == "Control") 1L else 2L
        re_i     <- site_re[site_idx]

        for (period in c("Before", "After")) {
          n_rounds <- if (period == "Before") n_before_rounds else n_after_rounds
          for (r in seq_len(n_rounds)) {
            logit_psi <- logit_psi_ctrl_before +
              (if (period == "After")      b_period_after else 0) +
              (if (st == "Stone_pile")     b_sp           else 0) +
              (if (st == "Stone_pile" && period == "After") b_sp_after else 0) +
              re_i
            psi <- plogis(logit_psi)
            z   <- rbinom(1L, 1L, psi)
            yA  <- rbinom(1L, 1L, z * pA)
            yB  <- rbinom(1L, 1L, z * pB)
            idx <- idx + 1L
            rows[[idx]] <- data.frame(
              pair_id   = pair,
              site_id   = site_idx,
              site_type = st,
              period    = period,
              detected  = as.integer(yA | yB)
            )
          }
        }
      }
    }

    sim_data <- dplyr::bind_rows(rows) |>
      dplyr::mutate(
        period    = factor(period,    levels = c("Before", "After")),
        site_type = factor(site_type, levels = c("Control", "Stone_pile")),
        site_id   = factor(site_id),
        pair_id   = factor(pair_id)
      )

    # Fast logistic GLMM surrogate — skip if data is perfectly separated
    m <- tryCatch(
      lme4::glmer(detected ~ period * site_type + (1 | pair_id / site_id),
                  data = sim_data, family = binomial,
                  control = lme4::glmerControl(optimizer = "bobyqa")),
      error   = function(e) NULL,
      warning = function(w) NULL
    )
    if (is.null(m)) return(NA)

    coef_name <- grep("periodAfter:site_typeStone_pile",
                      names(lme4::fixef(m)), value = TRUE)
    if (length(coef_name) == 0) return(NA)
    est <- lme4::fixef(m)[coef_name]
    se  <- sqrt(diag(vcov(m)))[coef_name]
    # Wald 95% CI lower bound > 0 → BACI effect detected
    (est - 1.96 * se) > 0
  }

  run_power <- function(n_pairs, n_after_rounds, n_sims = 300) {
    results <- vapply(seq_len(n_sims), function(i)
      one_sim(n_pairs, n_after_rounds,
              logit_psi_ctrl_before, b_period_after, b_sp, b_sp_after,
              logit_pA, logit_pB, sigma_site),
      logical(1))
    mean(results, na.rm = TRUE)
  }

  set.seed(2026)
  power_grid <- tidyr::crossing(
    n_pairs        = c(3, 5, 7, 10, 15, 20, 30),
    n_after_rounds = c(2, 4, 6, 8, 10, 12)
  ) |>
    dplyr::mutate(
      power = purrr::map2_dbl(n_pairs, n_after_rounds, run_power,
                               .progress = TRUE)
    )

  saveRDS(power_grid, power_cache)
  cat("Power analysis complete and cached.\n")

} else {
  power_grid <- readRDS(power_cache)
  cat("Loaded cached power analysis.\n")
}

print(power_grid |> tidyr::pivot_wider(names_from = n_after_rounds,
                                        values_from = power,
                                        names_prefix = "rounds_"))
```

```{r}
#| label: fig-power-plots-setup
#| include: false
p_lines <- power_grid |>
  dplyr::mutate(n_after_rounds = factor(n_after_rounds,
                  labels = paste0(c(2,4,6,8,10,12), " After rounds"))) |>
  ggplot(aes(x = n_pairs, y = power, colour = n_after_rounds,
             group = n_after_rounds)) +
  geom_hline(yintercept = 0.8, linetype = "dashed", colour = "grey40") +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  annotate("text", x = max(power_grid$n_pairs) * 0.98, y = 0.82,
           label = "80 % power", hjust = 1, colour = "grey40", size = 3.5) +
  scale_colour_manual(
    values = c("#313695","#4575b4","#74add1","#f46d43","#d73027","#a50026"),
    name   = NULL
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_x_continuous(breaks = c(3, 5, 7, 10, 15, 20, 30)) +
  labs(title = "Occupancy (BACI)",
       x = "Number of site pairs (Stone pile + Control)",
       y = "Statistical power") +
  base_theme +
  theme(panel.grid.minor = element_blank(), legend.position = "right")
```

```{r}
#| label: tbl-power-summary
#| tbl-cap: "Minimum site pairs required to reach 80% power for the occupancy BACI effect, by number of After-period monitoring rounds."
# What does it take to hit 80% power at the current effect size?
power_grid |>
  dplyr::filter(power >= 0.8) |>
  dplyr::group_by(n_after_rounds) |>
  dplyr::slice_min(n_pairs, n = 1) |>
  dplyr::ungroup() |>
  dplyr::transmute(
    `After rounds` = n_after_rounds,
    `Site pairs needed` = n_pairs,
    `Power achieved` = scales::percent(power, accuracy = 1)
  ) |>
  knitr::kable()
```

# CPUE-based Power Analysis (M2aak — Kōura abundance BACI)

The occupancy analysis sits near a ceiling (ψ ≈ 0.81–0.95) so the detectable
BACI signal is compressed. CPUE captures the abundance dimension that occupancy
misses. M2aak (Kōura CPUE, Stone_pile vs Control, full BACI) shows a positive
but non-significant BACI of 0.311 log-units (×1.36 on the ratio scale, p = 0.800).
This power analysis asks: how many site pairs and monitoring rounds would be
needed to detect an effect of that size at 80% power?

**Model structure simulated:**
`log(CPUE) ~ period * site_type + monitoring_int + (1|pair_id) + (1|site_id)`

Fixed effects come directly from M2aak. Random effect SDs and the log-scale
residual SD are set to ecologically plausible defaults — update them from
`VarCorr(M2aak)` and `sigma(M2aak)` if you want exact values.

```{r}
#| label: cpue-params
library(lme4)

# Read fixed effects directly from the model summary CSV (written by analysis.qmd)
m2_params <- readr::read_csv(
  file.path("outputs", "tbl-models-summary-full.csv"),
  show_col_types = FALSE
)

get_fe <- function(mid, trm, col = "estimate") {
  m2_params |> dplyr::filter(model_id == mid, term == trm) |> dplyr::pull({{ col }})
}

cpue_b0      <- get_fe("M2aak", "(Intercept)")
cpue_b_after <- get_fe("M2aak", "PeriodAfter")
cpue_b_sp    <- get_fe("M2aak", "siteStone_pile")
cpue_b_time  <- get_fe("M2aak", "monitoring_int")
cpue_b_baci  <- get_fe("M2aak", "PeriodAfter:siteStone_pile")
cpue_baci_se <- get_fe("M2aak", "PeriodAfter:siteStone_pile", "std.error")
cpue_baci_lo <- get_fe("M2aak", "PeriodAfter:siteStone_pile", "conf.low")
cpue_baci_hi <- get_fe("M2aak", "PeriodAfter:siteStone_pile", "conf.high")
cpue_baci_p  <- get_fe("M2aak", "PeriodAfter:siteStone_pile", "p.value")

# Variance components — written by analysis.qmd via VarCorr(M2aak)
vc_path <- file.path(der_dir, "M2aak_variance_components.csv")
if (file.exists(vc_path)) {
  vc <- readr::read_csv(vc_path, show_col_types = FALSE)
  sigma_pair_cpue  <- vc$sigma_pair
  sigma_site_cpue  <- vc$sigma_site
  sigma_resid_cpue <- vc$sigma_resid
} else {
  sigma_pair_cpue  <- 1.2999
  sigma_site_cpue  <- 0.8087
  sigma_resid_cpue <- 0.9611
  warning("M2aak_variance_components.csv not found — using values from 2026-08-10 session.")
}

cat("M2aak parameters (from tbl-models-summary-full.csv):\n")
cat(sprintf("  BACI estimate:   %.3f (SE = %.3f, p = %.3f)\n",
            cpue_b_baci, cpue_baci_se, cpue_baci_p))
cat(sprintf("  95%% CI on BACI: [%.2f, %.2f]\n", cpue_baci_lo, cpue_baci_hi))
cat(sprintf("  Ratio scale:     exp(%.3f) = %.2f×\n", cpue_b_baci, exp(cpue_b_baci)))
cat(sprintf("  PeriodAfter:     %.3f (strong time effect; both sites gain)\n", cpue_b_after))
cat(sprintf("  monitoring_int:  %.3f (declining trend over rounds)\n", cpue_b_time))
```

```{r}
#| label: cpue-power-sim

one_sim_cpue <- function(n_pairs, n_after_rounds) {

  dat <- tidyr::crossing(
    pair_idx  = seq_len(n_pairs),
    site_type = c("Control", "Stone_pile"),
    round_no  = 0L:as.integer(n_after_rounds)
  ) |>
    dplyr::mutate(
      pair_id        = factor(paste0("p", pair_idx)),
      site_id        = factor(paste0("p", pair_idx, "_", site_type)),
      period         = factor(ifelse(round_no == 0L, "Before", "After"),
                              levels = c("Before", "After")),
      monitoring_int = as.numeric(round_no),
      is_sp          = as.integer(site_type == "Stone_pile"),
      is_after       = as.integer(period == "After")
    )

  # Sample random effects for this iteration
  re_pair <- stats::setNames(
    rnorm(n_pairs, 0, sigma_pair_cpue),
    paste0("p", seq_len(n_pairs))
  )
  re_site <- stats::setNames(
    rnorm(n_pairs * 2L, 0, sigma_site_cpue),
    paste0("p", rep(seq_len(n_pairs), each = 2L), "_",
           c("Control", "Stone_pile"))
  )

  dat <- dat |>
    dplyr::mutate(
      eta  = cpue_b0 +
             cpue_b_after * is_after +
             cpue_b_sp    * is_sp +
             cpue_b_time  * monitoring_int +
             cpue_b_baci  * (is_after * is_sp) +
             re_pair[as.character(pair_id)] +
             re_site[as.character(site_id)],
      cpue = exp(stats::rnorm(dplyr::n(), eta, sigma_resid_cpue))
    )

  fit <- tryCatch(
    suppressWarnings(
      lme4::lmer(log(cpue) ~ period * site_type + monitoring_int +
                   (1 | pair_id) + (1 | site_id),
                 data = dat, REML = FALSE)
    ),
    error = function(e) NULL
  )

  if (is.null(fit)) return(NA_real_)

  cf  <- lme4::fixef(fit)
  vcv <- as.matrix(vcov(fit))
  nm  <- "periodAfter:site_typeStone_pile"
  if (!nm %in% names(cf)) return(NA_real_)

  est <- cf[nm]
  se  <- sqrt(vcv[nm, nm])
  if (is.na(se) || se < 1e-9) return(NA_real_)

  # Two-sided 95% CI must exclude zero in positive direction
  as.numeric((est - 1.96 * se) > 0)
}

run_power_cpue <- function(n_pairs, n_after_rounds, n_sims = 300L) {
  results <- vapply(
    seq_len(n_sims),
    function(s) {
      set.seed(s * 100L + n_pairs * 10L + n_after_rounds)
      one_sim_cpue(n_pairs, n_after_rounds)
    },
    numeric(1L)
  )
  mean(results, na.rm = TRUE)
}

cpue_power_cache <- file.path(der_dir, "cpue_power_analysis.rds")

if (!file.exists(cpue_power_cache)) {
  cat("Running CPUE power analysis (200 sims × 42 scenarios)...\n")

  cpue_power_grid <- tidyr::crossing(
    n_pairs        = c(3L, 5L, 7L, 10L, 15L, 20L, 30L),
    n_after_rounds = c(2L, 4L, 6L, 8L, 10L, 12L)
  ) |>
    dplyr::mutate(
      power = purrr::map2_dbl(n_pairs, n_after_rounds, run_power_cpue,
                              .progress = TRUE)
    )

  saveRDS(cpue_power_grid, cpue_power_cache)
  cat("CPUE power analysis complete and cached.\n")

} else {
  cpue_power_grid <- readRDS(cpue_power_cache)
  cat("Loaded cached CPUE power analysis.\n")
}

print(cpue_power_grid |>
        tidyr::pivot_wider(names_from = n_after_rounds, values_from = power,
                           names_prefix = "rounds_"))
```

```{r}
#| label: fig-cpue-power-plots-setup
#| include: false
p_lines_cpue <- cpue_power_grid |>
  dplyr::mutate(n_after_rounds = factor(n_after_rounds,
                  labels = paste0(c(2, 4, 6, 8, 10, 12), " After rounds"))) |>
  ggplot(aes(x = n_pairs, y = power, colour = n_after_rounds,
             group = n_after_rounds)) +
  geom_hline(yintercept = 0.8, linetype = "dashed", colour = "grey40") +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  annotate("text", x = max(cpue_power_grid$n_pairs) * 0.98, y = 0.82,
           label = "80 % power", hjust = 1, colour = "grey40", size = 3.5) +
  scale_colour_manual(
    values = c("#313695", "#4575b4", "#74add1", "#f46d43", "#d73027", "#a50026"),
    name   = NULL
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_x_continuous(breaks = c(3, 5, 7, 10, 15, 20, 30)) +
  labs(title = "Kōura CPUE (BACI)",
       x = "Number of site pairs (Stone pile + Control)",
       y = "Statistical power") +
  base_theme +
  theme(panel.grid.minor = element_blank(), legend.position = "right")
```

```{r}
#| label: fig-power-curves-combined
#| include: false
#| cache: false
#| fig-cap: "Simulation-based power curves for the occupancy BACI (left) and kōura CPUE BACI (right). Each line represents a different number of After-period monitoring rounds. Dashed line marks 80% power."
#| fig-width: 12
#| fig-height: 5
p_power_combined <- p_lines + p_lines_cpue +
  patchwork::plot_layout(guides = "collect") &
  theme(legend.position = "right")

ggsave(file.path("outputs", "fig-power-curves-combined.png"),
       p_power_combined, width = 12, height = 5, dpi = 300)
p_power_combined
```

```{r}
#| label: tbl-cpue-power-summary
#| tbl-cap: "Minimum site pairs required to reach 80% power for the kōura CPUE BACI effect (M2aak), by number of After-period monitoring rounds."
at_80 <- cpue_power_grid |>
  dplyr::filter(power >= 0.8) |>
  dplyr::group_by(n_after_rounds) |>
  dplyr::slice_min(n_pairs, n = 1) |>
  dplyr::ungroup() |>
  dplyr::transmute(
    `After rounds`      = n_after_rounds,
    `Site pairs needed` = n_pairs,
    `Power achieved`    = scales::percent(power, accuracy = 1)
  )

cap <- paste0(
  "CPUE BACI power (M2aak: exp(BACI) = ", round(exp(cpue_b_baci), 2), "×; ",
  "σ_pair = ", sigma_pair_cpue, ", σ_site = ", sigma_site_cpue,
  ", σ_resid = ", sigma_resid_cpue, "). ",
  "Log-normal surrogate (lmer)."
)

if (nrow(at_80) == 0) {
  cat("No scenario in the grid reaches 80% power.\n")
  cat(sprintf("Maximum power achieved: %.0f%% (n_pairs = 30, rounds = 8).\n",
              100 * max(cpue_power_grid$power)))
  cat("The between-pair variance (σ_pair = 1.30) dominates, matching the\n")
  cat("occupancy result: more rounds do not compensate for too few pairs.\n")
  cpue_power_grid |>
    dplyr::filter(power == max(power)) |>
    knitr::kable()
} else {
  knitr::kable(at_80)
}
```

# CPUE effect size trajectory — when will detection become feasible?

The power analyses above assume the current BACI effect size stays fixed. But
M2bak (kōura CPUE, after-period trajectory) shows stone pile CPUE is declining
more slowly than controls (+0.243 log-units per round divergence). If this
trajectory continues, the cumulative CPUE BACI effect grows with each monitoring
round, eventually crossing the detection threshold for 5 pairs.

Note: this projection applies to **CPUE detection only**. Occupancy detection
is structurally limited by the near-ceiling ψ values and does not benefit
meaningfully from a growing effect — the signal is compressed regardless of
effect size.

At 4 quarterly rounds per year, 5–6 additional rounds corresponds to
approximately **1.5 years** of continued monitoring — assuming the trajectory holds.

**Caveat:** kōura colonisation of stone pile habitat is expected to saturate
as structures approach carrying capacity. The linear projection is therefore
an optimistic lower bound — the true detection horizon is likely later.

```{r}
#| label: fig-trajectory-projection
#| include: false
#| cache: false
#| fig-cap: "Projected kōura CPUE BACI effect size over additional monitoring rounds. The solid line shows the linear projection based on the estimated post-construction trajectory; shaded ribbon shows 95% uncertainty interval. Dashed horizontal line marks the detection threshold for 80% power with five site pairs."

# Parameters read from model summary CSV (written by analysis.qmd)
baci_now       <- cpue_b_baci   # M2aak BACI — already loaded in cpue-params
baci_se_now    <- cpue_baci_se  # M2aak BACI SE

traj_per_round <- get_fe("M2bak", "siteStone_pile:monitoring_int")
traj_se        <- get_fe("M2bak", "siteStone_pile:monitoring_int", "std.error")

# Detection threshold: SE at n=30 back-calculated from simulation results
ref_power_30 <- cpue_power_grid |>
  dplyr::filter(n_pairs == 30) |>
  dplyr::summarise(p = mean(power)) |>
  dplyr::pull(p)
se_at_30 <- baci_now / (qnorm(ref_power_30) + qnorm(0.975))
se_at_5  <- se_at_30 * sqrt(30 / 5)
threshold  <- (qnorm(0.975) + qnorm(0.80)) * se_at_5

cat(sprintf("Detection threshold (80%% power, 5 pairs): %.2f log-units (%.1f×)\n",
            threshold, exp(threshold)))
cat(sprintf("Current BACI: %.3f log-units (%.2f×)\n", baci_now, exp(baci_now)))
cat(sprintf("Gap to threshold: %.2f log-units\n", threshold - baci_now))
cat(sprintf("Rounds needed (linear): %.1f additional quarterly rounds (~%.1f years)\n",
            (threshold - baci_now) / traj_per_round,
            (threshold - baci_now) / traj_per_round / 4))

# Build projection data
future_rounds <- 0:10
proj <- tibble::tibble(
  additional_rounds = future_rounds,
  baci_proj  = baci_now + traj_per_round * future_rounds,
  # propagate uncertainty: current BACI SE + trajectory SE × rounds
  baci_se    = sqrt(baci_se_now^2 + (traj_se * future_rounds)^2),
  baci_lwr   = baci_proj - 1.96 * baci_se,
  baci_upr   = baci_proj + 1.96 * baci_se)

rounds_to_80 <- (threshold - baci_now) / traj_per_round

p_traj <- ggplot(proj, aes(x = additional_rounds, y = baci_proj)) +
  geom_ribbon(aes(ymin = baci_lwr, ymax = baci_upr),
              fill = "black", alpha = 0.15) +
  geom_line(colour = "black", linewidth = 1.1) +
  geom_pointrange(
    data = tibble::tibble(
      additional_rounds = 0,
      baci_proj = baci_now,
      baci_lwr  = baci_now - 1.96 * baci_se_now,
      baci_upr  = baci_now + 1.96 * baci_se_now),
    aes(ymin = baci_lwr, ymax = baci_upr),
    colour = "black", size = 0.8, linewidth = 1) +
  geom_hline(yintercept = threshold, linetype = "dashed",
             colour = "grey30", linewidth = 0.8) +
  annotate("text", x = 9.5, y = threshold - 0.3,
           label = "80% power threshold", hjust = 0.8,
           colour = "grey30", size = 3.5) +
  geom_vline(xintercept = rounds_to_80, linetype = "dotted",
             colour = "grey50", linewidth = 0.7) +
  annotate("rect", xmin = rounds_to_80, xmax = 10,
           ymin = -Inf, ymax = Inf,
           fill = "grey80", alpha = 0.10) +
  scale_x_continuous(breaks = seq(0, 10, 2),
                     name = "Additional quarterly monitoring rounds beyond current study") +
  scale_y_continuous(
    name    = "Projected BACI effect (log-units)",
    sec.axis = sec_axis(~ exp(.), name = "Ratio scale (×)",
                        breaks = c(1, 2, 4, 6, 8),
                        labels = c("1×", "2×", "4×", "6×", "8×"))) +
  base_theme +
  theme(panel.grid.minor = element_blank())

ggsave(file.path("outputs", "fig-trajectory-projection.png"),
       p_traj, width = 5, height = 4, dpi = 300)

p_traj
```

```{r}
#| label: power-scalars-csv
#| include: false

# Analytically compute n_pairs needed to detect baci_now at 80% power.
# SE scales as se_at_n = se_at_30 * sqrt(30 / n), so:
#   baci_now = (z_alpha + z_beta) * se_at_n  →  n = 30 * (se_at_30 * z_sum / baci_now)^2
z_sum         <- qnorm(0.975) + qnorm(0.80)
n_pairs_needed <- ceiling(30 * (se_at_30 * z_sum / baci_now)^2)

max_cpue_power <- max(cpue_power_grid$power)
max_occ_power  <- max(power_grid$power)

power_scalars <- tibble::tibble(
  key   = c("baci_effect", "baci_se",
            "threshold_80", "threshold_ratio",
            "traj_per_round", "traj_se",
            "rounds_to_80", "years_to_80",
            "n_pairs_needed",
            "max_cpue_power", "max_occ_power"),
  value = c(round(baci_now, 3),       round(baci_se_now, 3),
            round(threshold, 2),      round(threshold / baci_now, 1),
            round(traj_per_round, 3), round(traj_se, 3),
            round(rounds_to_80, 1),   round(rounds_to_80 / 4, 1),
            n_pairs_needed,
            round(max_cpue_power, 3), round(max_occ_power, 3))
)

readr::write_csv(power_scalars, file.path("outputs", "tbl-power-scalars.csv"))
power_scalars
```


# Session info
```{r}
#| label: session-info
#| include: true
#| code-fold: true
sessionInfo()
```