---
title: "Kōura two-method occupancy analysis — Lake Rotoiti stone-pile BACI"
format: html
execute:
  message: false
  warning: false
---

# Setup
```{r}
#| label: setup
library(brms)
library(cmdstanr)
library(tidyverse)
library(tidybayes)
library(posterior)
library(bayesplot)


options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

occ_dir <- file.path("outputs", "occupancy")
dir.create(occ_dir, showWarnings = FALSE, recursive = TRUE)

location_levels <- c("Control", "Stone_pile", "Reference1", "Reference2")
period_levels   <- c("Before", "After")
season_levels   <- c("Summer", "Autumn", "Winter", "Spring")
block_levels    <- c("Impact", "Reference")

round_lookup <- tibble(
  round  = 0:6,
  period = c("Before", rep("After", 6)),
  season = c("Summer", "Autumn", "Winter", "Spring", "Summer", "Autumn", "Winter"),
  year   = c(1, 1, 1, 1, 2, 2, 2)
) |>
  mutate(
    period = factor(period, levels = period_levels),
    season = factor(season, levels = season_levels)
  )

print(round_lookup)
```

# Custom two-method occupancy family

Two detection methods deployed at every site every round:

- **Method A** — Box traps (~32 % detection probability)
- **Method B** — Fyke nets (~63 % detection probability)

The custom brms/Stan family marginalises the latent occupancy state *z* over both
methods jointly, and tolerates partial missingness (one trap lost to storms) via
four `vint()` helper columns per occasion: `yA_obs`, `yB_obs` (trap retrieved?),
`yA`, `yB` (kōura detected?).

```{r}
#| label: occupancy-family
stan_funs_occ <- "
  real occ2_lpmf(int y, real psi, real pA, real pB,
                  int yA_obs, int yB_obs, int yA, int yB) {
    real ll_given_occ = 0;
    int any_detect = 0;

    if (yA_obs == 1) {
      ll_given_occ += yA * log(pA) + (1 - yA) * log1m(pA);
      if (yA == 1) any_detect = 1;
    }
    if (yB_obs == 1) {
      ll_given_occ += yB * log(pB) + (1 - yB) * log1m(pB);
      if (yB == 1) any_detect = 1;
    }

    if (any_detect == 1) {
      return log(psi) + ll_given_occ;
    } else {
      return log_sum_exp(log(psi) + ll_given_occ, log1m(psi));
    }
  }
"

stanvars_occ <- brms::stanvar(scode = stan_funs_occ, block = "functions")

# brms custom family — dpars must start with "mu" (brms >= 2.21)
# mu maps to psi (true occupancy); pA and pB are detection submodels
# vars must use brms internal vint names matching vint() order in formula
occ2 <- custom_family(
  name  = "occ2",
  dpars = c("mu", "pA", "pB"),
  links = c("logit", "logit", "logit"),
  lb    = c(0, 0, 0),
  ub    = c(1, 1, 1),
  type  = "int",
  vars  = c("vint1[n]", "vint2[n]", "vint3[n]", "vint4[n]")
)

log_lik_occ2 <- function(i, prep) {
  psi    <- brms::get_dpar(prep, "mu", i = i)
  pA     <- brms::get_dpar(prep, "pA", i = i)
  pB     <- brms::get_dpar(prep, "pB", i = i)
  yA_obs <- prep$data$vint1[i]; yB_obs <- prep$data$vint2[i]
  yA     <- prep$data$vint3[i]; yB     <- prep$data$vint4[i]
  ll_given_occ <- 0; any_detect <- 0
  if (yA_obs == 1) {
    ll_given_occ <- ll_given_occ + yA * log(pA) + (1 - yA) * log1p(-pA)
    if (yA == 1) any_detect <- 1
  }
  if (yB_obs == 1) {
    ll_given_occ <- ll_given_occ + yB * log(pB) + (1 - yB) * log1p(-pB)
    if (yB == 1) any_detect <- 1
  }
  if (any_detect == 1) log(psi) + ll_given_occ
  else log(psi * exp(ll_given_occ) + (1 - psi))
}

posterior_predict_occ2 <- function(i, prep, ...) {
  psi    <- brms::get_dpar(prep, "mu", i = i)
  pA     <- brms::get_dpar(prep, "pA", i = i)
  pB     <- brms::get_dpar(prep, "pB", i = i)
  yA_obs <- prep$data$vint1[i]; yB_obs <- prep$data$vint2[i]
  z      <- rbinom(length(psi), 1, psi)
  yA_sim <- ifelse(yA_obs == 1, rbinom(length(psi), 1, z * pA), NA_integer_)
  yB_sim <- ifelse(yB_obs == 1, rbinom(length(psi), 1, z * pB), NA_integer_)
  rowSums(cbind(yA_sim, yB_sim), na.rm = TRUE)
}

# posterior_epred returns psi (true occupancy), not detection-weighted catch
posterior_epred_occ2 <- function(prep) {
  brms::get_dpar(prep, "mu")
}

cat("Custom family 'occ2' defined.\n")
```

# Import and prepare data
```{r}
#| label: import-data
model_data_path <- file.path("outputs", "occupancy", "model_data.rds")

if (!file.exists(model_data_path)) {

  combined <- readr::read_csv(file.path("data", "derived", "Combined_data.csv"),
                               show_col_types = FALSE)
  fish     <- readr::read_csv(file.path("data", "derived", "Fish_data.csv"),
                               show_col_types = FALSE)

  backbone <- combined |>
    dplyr::transmute(
      site_id            = as.integer(site_id),
      round              = as.integer(monitoring),
      monitoring_id      = monitoring_id,
      location_type      = dplyr::case_when(
        Site_type_detail == "Stone_pile"  ~ "Stone_pile",
        Site_type_detail == "Control"     ~ "Control",
        Site_type_detail == "Reference_1" ~ "Reference1",
        Site_type_detail == "Reference_2" ~ "Reference2",
        TRUE                              ~ NA_character_),
      block              = dplyr::case_when(
        Site_type_detail %in% c("Stone_pile", "Control")      ~ "Impact",
        Site_type_detail %in% c("Reference_1", "Reference_2") ~ "Reference",
        TRUE                                                    ~ NA_character_),
      temperature        = as.numeric(Temperature),
      monitoring_quality = as.character(monitoring_quality),
      pair_id            = pair_id
    ) |>
    dplyr::filter(!is.na(location_type))

  # One row per kōura individual → collapse to presence per site×round×method
  koura_by_method <- fish |>
    dplyr::filter(species == "Kōura") |>
    dplyr::group_by(site_id, monitoring_id, net_type) |>
    dplyr::summarise(detected = 1L, .groups = "drop") |>
    dplyr::mutate(site_id = as.integer(site_id))

  # Cross all valid site×round combinations with both methods; fill 0 for no-catch
  all_combos <- backbone |>
    dplyr::select(site_id, monitoring_id) |>
    tidyr::crossing(net_type = c("Box", "Fyke"))

  detection_wide <- all_combos |>
    dplyr::left_join(
      koura_by_method |> dplyr::select(site_id, monitoring_id, net_type, detected),
      by = c("site_id", "monitoring_id", "net_type")) |>
    dplyr::mutate(detected = tidyr::replace_na(detected, 0L)) |>
    tidyr::pivot_wider(names_from = net_type, values_from = detected) |>
    dplyr::rename(yA = Box, yB = Fyke)

  real_data <- backbone |>
    dplyr::left_join(detection_wide, by = c("site_id", "monitoring_id")) |>
    dplyr::mutate(
      # monitoring_quality == "complete" → both traps usable
      usable = monitoring_quality == "complete",
      yA_obs = as.integer(usable),
      yB_obs = as.integer(usable),
      yA     = dplyr::if_else(usable, yA, NA_integer_),
      yB     = dplyr::if_else(usable, yB, NA_integer_)
    ) |>
    dplyr::filter(!(yA_obs == 0 & yB_obs == 0)) |>   # drop zero-info occasions
    dplyr::left_join(round_lookup, by = "round") |>
    dplyr::mutate(
      location_type = factor(location_type, levels = location_levels),
      block         = factor(block, levels = block_levels),
      location_id   = paste(site_id, location_type, sep = "_"),
      temp_z        = as.numeric(scale(temperature))  # global standardisation
    ) |>
    dplyr::select(site_id, block, location_type, location_id, round,
                  period, season, year, temperature, temp_z,
                  yA_obs, yB_obs, yA, yB, pair_id, monitoring_quality)

  # Stan requires no NAs: replace NA detection with 0 (yA_obs/yB_obs flags
  # tell Stan to ignore the value when the trap was lost)
  model_data <- real_data |>
    dplyr::mutate(
      y_dummy = 0L,
      yA = dplyr::if_else(is.na(yA), 0L, yA),
      yB = dplyr::if_else(is.na(yB), 0L, yB)
    )

  # Validation
  stopifnot(
    all(model_data$yA %in% c(0L, 1L)),
    all(model_data$yB %in% c(0L, 1L)),
    all(model_data$round %in% 0:6),
    all(model_data$yA_obs == 1 | model_data$yB_obs == 1)
  )

  saveRDS(model_data, model_data_path)
  cat("Imported and saved:", nrow(model_data), "occasions\n")

} else {
  model_data <- readRDS(model_data_path)
  cat("Loaded cached model data:", nrow(model_data), "occasions\n")
}

cat("\nOccasions by location × period × season:\n")
print(dplyr::count(model_data, location_type, period, season))

cat("\nKōura detection rates by method:\n")
cat("  Box trap (yA):", round(mean(model_data$yA), 3), "\n")
cat("  Fyke net (yB):", round(mean(model_data$yB), 3), "\n")
```

# Priors

Weakly informative normal(0, 1.5) priors on log-odds scale for fixed effects
(allows ±3 units ≈ occupancy/detection probability ranging ~0.05–0.95).
Normal(0, 1) on random-effect SDs.

```{r}
#| label: priors
priors_occ <- c(
  prior(normal(0, 1.5), class = "b"),
  prior(normal(0, 1.5), class = "b", dpar = "pA"),
  prior(normal(0, 1.5), class = "b", dpar = "pB"),
  prior(normal(0, 1),   class = "sd"),
  prior(normal(0, 1),   class = "sd", dpar = "pA"),
  prior(normal(0, 1),   class = "sd", dpar = "pB")
)
```

# Model 1: Full model (all seasons)

Primary model using all 128 occasions. Tests `location_type × period × season`
interaction. Provides whole-study headline occupancy estimates and uses the full
data to estimate variance components (site random effects, detection–temperature
slope) most precisely.

```{r}
#| label: fit-full
full_path <- file.path("outputs", "occupancy", "fit_full.rds")

if (!file.exists(full_path)) {

  bf_full <- bf(
    y_dummy | vint(yA_obs, yB_obs, yA, yB) ~
      location_type * period * season + (1 | site_id),
    pA ~ 1 + temp_z + (1 | site_id),
    pB ~ 1 + temp_z + (1 | site_id),
    family = occ2
  )

  fit_full <- brm(
    bf_full,
    data     = model_data,
    prior    = priors_occ,
    stanvars = stanvars_occ,
    chains   = 4, iter = 4000, warmup = 1500,
    seed     = 2026,
    control  = list(adapt_delta = 0.97, max_treedepth = 13)
  )

  saveRDS(fit_full, full_path)
  cat("Fitted and saved full model.\n")

} else {
  fit_full <- readRDS(full_path)
  cat("Loaded cached full model.\n")
}

summary(fit_full)
cat("\nMax Rhat:   ", max(brms::rhat(fit_full), na.rm = TRUE), "\n")
cat("Min ESS:    ", min(posterior::summarise_draws(fit_full)$ess_bulk, na.rm = TRUE), "\n")
cat("Divergences:", sum(brms::nuts_params(fit_full, pars = "divergent__")$Value), "\n")
```

```{r}
#| label: traceplots-full
bayesplot::mcmc_trace(fit_full, regex_pars = "^b_") +
  labs(title = "Full model — fixed-effect trace plots")
```

# Model 2: Summer-only model

Confirmatory model restricted to Summer rounds (R0 Before, R4 After). Targets
the a-priori hypothesis that refuge value is greatest in summer when catfish
pressure and thermal stratification are highest. Treated as a robustness check
on the summer cells of the full model, not a replacement for it.

```{r}
#| label: fit-summer
summer_path <- file.path("outputs", "occupancy", "fit_summer.rds")
summer_data <- model_data |> dplyr::filter(season == "Summer")

cat("Summer subset:", nrow(summer_data), "occasions\n")
print(dplyr::count(summer_data, location_type, period))

if (!file.exists(summer_path)) {

  bf_summer <- bf(
    y_dummy | vint(yA_obs, yB_obs, yA, yB) ~
      location_type * period + (1 | site_id),
    pA ~ 1 + temp_z + (1 | site_id),
    pB ~ 1 + temp_z + (1 | site_id),
    family = occ2
  )

  fit_summer <- brm(
    bf_summer,
    data     = summer_data,
    prior    = priors_occ,
    stanvars = stanvars_occ,
    chains   = 4, iter = 4000, warmup = 1500,
    seed     = 2026,
    control  = list(adapt_delta = 0.97, max_treedepth = 13)
  )

  saveRDS(fit_summer, summer_path)
  cat("Fitted and saved summer model.\n")

} else {
  fit_summer <- readRDS(summer_path)
  cat("Loaded cached summer model.\n")
}

summary(fit_summer)
cat("\nMax Rhat:   ", max(brms::rhat(fit_summer), na.rm = TRUE), "\n")
cat("Min ESS:    ", min(posterior::summarise_draws(fit_summer)$ess_bulk, na.rm = TRUE), "\n")
cat("Divergences:", sum(brms::nuts_params(fit_summer, pars = "divergent__")$Value), "\n")
```

```{r}
#| label: traceplots-summer
bayesplot::mcmc_trace(fit_summer, regex_pars = "^b_") +
  labs(title = "Summer model — fixed-effect trace plots")
```

# Model 3: Warm-season model (Summer + Autumn)

Ecological rationale: thermal stratification in Lake Rotoiti peaks in summer and
persists through autumn, compressing kōura into the oxygenated nearshore zone
where stone piles sit. Catfish predation pressure is also highest in these two
seasons. This model tests the restoration signal in the combined warm season
(76 occasions) with more power than summer-only (40 occasions).

Season is included as an additive covariate — not a full three-way interaction —
because the data support estimating seasonal baseline differences in occupancy
but not the full `location_type × period × season` interaction in this subset.

```{r}
#| label: fit-warm
warm_path <- file.path("outputs", "occupancy", "fit_warm.rds")
warm_data <- model_data |>
  dplyr::filter(season %in% c("Summer", "Autumn")) |>
  dplyr::mutate(season = droplevels(season))

cat("Warm subset:", nrow(warm_data), "occasions\n")
print(dplyr::count(warm_data, location_type, period, season))

if (!file.exists(warm_path)) {

  bf_warm <- bf(
    y_dummy | vint(yA_obs, yB_obs, yA, yB) ~
      location_type * period + season + (1 | site_id),
    pA ~ 1 + temp_z + (1 | site_id),
    pB ~ 1 + temp_z + (1 | site_id),
    family = occ2
  )

  fit_warm <- brm(
    bf_warm,
    data     = warm_data,
    prior    = priors_occ,
    stanvars = stanvars_occ,
    chains   = 4, iter = 4000, warmup = 1500,
    seed     = 2026,
    control  = list(adapt_delta = 0.99, max_treedepth = 13)
  )

  saveRDS(fit_warm, warm_path)
  cat("Fitted and saved warm model.\n")

} else {
  fit_warm <- readRDS(warm_path)
  cat("Loaded cached warm model.\n")
}

summary(fit_warm)
cat("\nMax Rhat:   ", max(brms::rhat(fit_warm), na.rm = TRUE), "\n")
cat("Min ESS:    ", min(posterior::summarise_draws(fit_warm)$ess_bulk, na.rm = TRUE), "\n")
cat("Divergences:", sum(brms::nuts_params(fit_warm, pars = "divergent__")$Value), "\n")
```

```{r}
#| label: traceplots-warm
bayesplot::mcmc_trace(fit_warm, regex_pars = "^b_") +
  labs(title = "Warm model — fixed-effect trace plots")
```

# Extract posterior contrasts

Four contrasts of ecological interest, evaluated in the After period:

1. **Stone_pile − Control** — did rocks help, relative to the paired control?
2. **Reference1 − Reference2** — spatial noise floor (how similar are reference sites?)
3. **Control − Reference (mean)** — how far below reference quality is the control?
4. **Stone_pile − Reference (mean)** — how close to optimal did stone piles get?

Plus the **BACI double-difference**: `(Stone_pile After−Before) − (Control After−Before)` —
the most defensible causal estimate, as it removes any pre-existing gap between site types.

```{r}
#| label: contrast-functions
summarise_draws_vec <- function(x, label) {
  tibble(
    contrast     = label,
    mean         = mean(x),
    median       = median(x),
    lower95      = quantile(x, 0.025),
    upper95      = quantile(x, 0.975),
    p_direction  = mean(x > 0)
  )
}

get_psi_draws <- function(fit, model_data, include_season = TRUE) {
  grid_vars <- if (include_season) {
    model_data |> dplyr::distinct(location_type, period, season)
  } else {
    model_data |> dplyr::distinct(location_type, period)
  }
  newdata <- grid_vars |>
    dplyr::mutate(
      site_id  = model_data$site_id[1],
      temp_z   = 0, yA_obs = 1, yB_obs = 1, yA = 0, yB = 0, y_dummy = 0
    )
  draws <- posterior_epred(fit, newdata = newdata, re_formula = NA, dpar = "mu")
  colnames(draws) <- if (include_season) {
    paste(newdata$location_type, newdata$period, newdata$season, sep = "|")
  } else {
    paste(newdata$location_type, newdata$period, sep = "|")
  }
  list(draws = draws, grid = newdata)
}

build_contrasts <- function(draws, grid, period_val, season_val = NULL) {
  key <- function(loc) {
    if (is.null(season_val)) paste(loc, period_val, sep = "|")
    else paste(loc, period_val, season_val, sep = "|")
  }
  d_add      <- draws[, key("Stone_pile")]
  d_ctrl     <- draws[, key("Control")]
  d_ref1     <- draws[, key("Reference1")]
  d_ref2     <- draws[, key("Reference2")]
  d_ref_mean <- (d_ref1 + d_ref2) / 2
  bind_rows(
    summarise_draws_vec(d_add  - d_ctrl,      "Stone_pile - Control"),
    summarise_draws_vec(d_ref1 - d_ref2,      "Reference1 - Reference2"),
    summarise_draws_vec(d_ctrl - d_ref_mean,  "Control - Reference (mean)"),
    summarise_draws_vec(d_add  - d_ref_mean,  "Stone_pile - Reference (mean)")
  ) |>
    dplyr::mutate(period = period_val,
                  season = if (is.null(season_val)) NA_character_ else season_val)
}

build_baci <- function(draws, include_season, season_val = NULL) {
  key <- function(loc, period) {
    if (include_season) paste(loc, period, season_val, sep = "|")
    else paste(loc, period, sep = "|")
  }
  d <- (draws[, key("Stone_pile", "After")] - draws[, key("Stone_pile", "Before")]) -
       (draws[, key("Control",    "After")] - draws[, key("Control",    "Before")])
  summarise_draws_vec(d, "BACI: (Stone_pile After-Before) - (Control After-Before)") |>
    dplyr::mutate(period = "Before-vs-After",
                  season = if (include_season) season_val else NA_character_)
}

run_contrasts <- function(fit, model_data, model_label, include_season = TRUE) {
  psi_out <- get_psi_draws(fit, model_data, include_season)
  draws   <- psi_out$draws
  grid    <- psi_out$grid

  cell_summary <- tibble(
    location_type = grid$location_type,
    period        = grid$period,
    season        = if (include_season) grid$season else NA
  ) |>
    dplyr::mutate(
      psi_mean = apply(draws, 2, mean),
      psi_lwr  = apply(draws, 2, quantile, 0.025),
      psi_upr  = apply(draws, 2, quantile, 0.975)
    )

  if (include_season) {
    seasons_present      <- unique(grid$season)
    contrasts_after      <- purrr::map_dfr(seasons_present,
                              ~ build_contrasts(draws, grid, "After", .x))
    seasons_with_before  <- grid |>
      dplyr::filter(period == "Before") |> dplyr::pull(season) |> unique()
    baci <- purrr::map_dfr(seasons_with_before,
               ~ build_baci(draws, include_season = TRUE, season_val = .x))
  } else {
    contrasts_after <- build_contrasts(draws, grid, "After")
    baci            <- build_baci(draws, include_season = FALSE)
  }

  contrasts_all <- dplyr::bind_rows(contrasts_after, baci) |>
    dplyr::mutate(model = model_label)

  readr::write_csv(cell_summary,
    file.path("outputs", "occupancy", paste0(model_label, "_psi_estimates.csv")))
  readr::write_csv(contrasts_all,
    file.path("outputs", "occupancy", paste0(model_label, "_contrasts.csv")))

  # Shared display helpers
  site_labels  <- c(Control = "Control", Stone_pile = "Stone pile",
                    Reference1 = "Reference 1", Reference2 = "Reference 2")
  site_colours <- c(Control = "#E69F00", Stone_pile = "#009E73",
                    Reference1 = "#56B4E9", Reference2 = "#CC79A7")
  season_colours <- c(Summer = "#D55E00", Autumn = "#E69F00",
                      Winter = "#56B4E9", Spring = "#009E73")

  # Tidy contrast labels for display
  contrasts_plot <- contrasts_all |>
    dplyr::mutate(
      contrast = dplyr::recode(contrast,
        "Stone_pile - Control"                                         = "Stone pile vs Control",
        "Reference1 - Reference2"                                      = "Reference 1 vs Reference 2",
        "Control - Reference (mean)"                                   = "Control vs Reference",
        "Stone_pile - Reference (mean)"                                = "Stone pile vs Reference",
        "BACI: (Stone_pile After-Before) - (Control After-Before)"     = "BACI (Stone pile − Control)",
        "BACI interaction: (Stone_pile After-Before) - (Control After-Before)" = "BACI (Stone pile − Control)"
      ),
      period = dplyr::recode(period,
        "After"          = "After treatment",
        "Before-vs-After"= "BACI"
      )
    )

  # ---- Plot 1: occupancy probability ----------------------------------------
  # Only draw connecting lines for location × season groups that have
  # both Before and After estimates (avoids geom_line warning in Autumn panel)
  lines_data <- cell_summary |>
    dplyr::group_by(location_type, if (include_season) season else NULL) |>
    dplyr::filter(dplyr::n_distinct(period) > 1) |>
    dplyr::ungroup()

  p1 <- ggplot(cell_summary,
               aes(x = period, y = psi_mean,
                   colour = location_type, group = location_type)) +
    geom_line(data = lines_data,
              position = position_dodge(0.35), linewidth = 0.7, alpha = 0.6) +
    geom_pointrange(aes(ymin = psi_lwr, ymax = psi_upr),
                    position = position_dodge(0.35),
                    size = 0.5, linewidth = 0.8, fatten = 3) +
    {if (include_season) facet_wrap(~season, nrow = 1)} +
    scale_colour_manual(values = site_colours, labels = site_labels,
                        name = "Site type") +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25),
                       labels = scales::percent_format(accuracy = 1)) +
    scale_x_discrete(labels = c(Before = "Before", After = "After")) +
    labs(y = "Occupancy probability (ψ)", x = NULL) +
    theme_bw(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      strip.background = element_rect(fill = "grey92", colour = NA),
      strip.text       = element_text(face = "bold"),
      legend.position  = "right"
    )
  ggsave(file.path("outputs", "occupancy", paste0(model_label, "_psi_plot.png")),
         p1, width = 10, height = 5, dpi = 300)

  # ---- Plot 2: contrast forest plot -----------------------------------------
  # Separate panels for After and BACI; fix x range symmetrically around 0
  x_lim <- max(abs(c(contrasts_plot$lower95, contrasts_plot$upper95)),
               na.rm = TRUE) * 1.05

  p2 <- ggplot(contrasts_plot,
               aes(x = mean, y = forcats::fct_rev(factor(contrast)),
                   colour = if (include_season) season else NULL)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50", linewidth = 0.6) +
    geom_pointrange(aes(xmin = lower95, xmax = upper95),
                    position = position_dodge(0.6),
                    size = 0.4, linewidth = 0.8, fatten = 3) +
    facet_wrap(~period, scales = "free_y", nrow = 1) +
    {if (include_season)
        scale_colour_manual(values = season_colours, name = "Season")
      else
        scale_colour_manual(values = c("After treatment" = "grey30"), guide = "none")} +
    scale_x_continuous(limits = c(-x_lim, x_lim),
                       labels = scales::label_number(accuracy = 0.1)) +
    labs(x = "Posterior contrast (probability scale)", y = NULL) +
    theme_bw(base_size = 12) +
    theme(
      panel.grid.minor   = element_blank(),
      panel.grid.major.y = element_blank(),
      strip.background   = element_rect(fill = "grey92", colour = NA),
      strip.text         = element_text(face = "bold"),
      axis.text.y        = element_text(size = 10),
      legend.position    = "right"
    )
  ggsave(file.path("outputs", "occupancy", paste0(model_label, "_contrasts_plot.png")),
         p2, width = 10, height = 5, dpi = 300)

  cat("\n====", model_label, "====\n")
  print(contrasts_all |>
    dplyr::select(model, contrast, period, season, mean, lower95, upper95, p_direction))

  list(cell_summary = cell_summary, contrasts = contrasts_all,
       psi_plot = p1, contrast_plot = p2)
}
```

```{r}
#| label: contrasts-full
results_full <- run_contrasts(fit_full, model_data,
                               model_label    = "full_model",
                               include_season = TRUE)
results_full$psi_plot
results_full$contrast_plot
```

```{r}
#| label: contrasts-summer
results_summer <- run_contrasts(fit_summer, summer_data,
                                 model_label    = "summer_model",
                                 include_season = FALSE)
results_summer$psi_plot
results_summer$contrast_plot
```

```{r}
#| label: contrasts-warm
results_warm <- run_contrasts(fit_warm, warm_data,
                               model_label    = "warm_model",
                               include_season = TRUE)
results_warm$psi_plot
results_warm$contrast_plot
```

# Combined contrast summary
```{r}
#| label: combined-summary
all_contrasts <- dplyr::bind_rows(
  results_full$contrasts,
  results_summer$contrasts,
  results_warm$contrasts
)

readr::write_csv(all_contrasts,
  file.path("outputs", "occupancy", "all_models_contrasts.csv"))

all_contrasts |>
  dplyr::select(model, contrast, period, season, mean, lower95, upper95, p_direction) |>
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  knitr::kable(caption = "Posterior contrasts across all three occupancy models (95 % CrI).")
```

# Power analysis — sites and rounds needed for statistical detection

The current study shows a consistent positive direction for the stone pile effect
but wide credible intervals due to limited site replication. This simulation-based
power analysis asks: **how many site pairs, or how many additional monitoring rounds,
would be needed to reliably detect the observed effect size?**

**Method:** Parameters (occupancy probabilities, detection rates, site-level variance)
are drawn from the posterior means of the warm-season model. For each combination
of site pairs and After rounds, 300 datasets are simulated and a logistic GLMM
is fitted as a fast surrogate. Power = proportion of simulations where the 95 % CI
of the BACI term (Stone pile × Period interaction) excludes zero.

```{r}
#| label: power-extract-params
# Extract posterior means from the warm model
fe         <- brms::fixef(fit_warm)[, "Estimate"]
draws_warm <- posterior::as_draws_df(fit_warm)

logit_psi_ctrl_before <- fe["Intercept"]
b_period_after        <- fe["periodAfter"]
b_sp                  <- fe["location_typeStone_pile"]
b_sp_after            <- fe["location_typeStone_pile:periodAfter"]
logit_pA              <- fe["pA_Intercept"]
logit_pB              <- fe["pB_Intercept"]

# Site-level random effect SD (occupancy submodel)
sigma_site <- posterior::summarise_draws(fit_warm) |>
  dplyr::filter(variable == "sd_site_id__Intercept") |>
  dplyr::pull(mean)

# Report the key parameters being used
cat("Parameters extracted from warm model posterior means:\n")
cat("  ψ Control  Before:", round(plogis(logit_psi_ctrl_before), 3), "\n")
cat("  ψ Control  After: ", round(plogis(logit_psi_ctrl_before + b_period_after), 3), "\n")
cat("  ψ Stone pile Before:", round(plogis(logit_psi_ctrl_before + b_sp), 3), "\n")
cat("  ψ Stone pile After: ", round(plogis(logit_psi_ctrl_before + b_sp + b_period_after + b_sp_after), 3), "\n")
cat("  pA (box trap): ", round(plogis(logit_pA), 3), "\n")
cat("  pB (fyke net): ", round(plogis(logit_pB), 3), "\n")
cat("  Site SD (logit): ", round(sigma_site, 3), "\n")
cat("  BACI log-odds:  ", round(b_sp_after, 3),
    "≈", round(plogis(logit_psi_ctrl_before + b_sp + b_period_after + b_sp_after) -
               plogis(logit_psi_ctrl_before + b_period_after), 3),
    "probability units\n")
```

```{r}
#| label: power-simulation
power_cache <- file.path("outputs", "occupancy", "power_analysis.rds")

if (!file.exists(power_cache)) {

  # Simulate one dataset and test if the BACI term's 95% CI excludes zero
  one_sim <- function(n_pairs, n_after_rounds,
                      logit_psi_ctrl_before, b_period_after, b_sp, b_sp_after,
                      logit_pA, logit_pB, sigma_site) {

    pA <- plogis(logit_pA)
    pB <- plogis(logit_pB)
    n_before_rounds <- 1L   # matches study design: 1 Before round

    site_re <- rnorm(n_pairs * 2, 0, sigma_site)
    rows    <- vector("list", n_pairs * 2 * (n_before_rounds + n_after_rounds))
    idx     <- 0L

    for (pair in seq_len(n_pairs)) {
      for (st in c("Control", "Stone_pile")) {
        site_idx <- (pair - 1L) * 2L + if (st == "Control") 1L else 2L
        re_i     <- site_re[site_idx]

        for (period in c("Before", "After")) {
          n_rounds <- if (period == "Before") n_before_rounds else n_after_rounds
          for (r in seq_len(n_rounds)) {
            logit_psi <- logit_psi_ctrl_before +
              (if (period == "After")      b_period_after else 0) +
              (if (st == "Stone_pile")     b_sp           else 0) +
              (if (st == "Stone_pile" && period == "After") b_sp_after else 0) +
              re_i
            psi <- plogis(logit_psi)
            z   <- rbinom(1L, 1L, psi)
            yA  <- rbinom(1L, 1L, z * pA)
            yB  <- rbinom(1L, 1L, z * pB)
            idx <- idx + 1L
            rows[[idx]] <- data.frame(
              pair_id   = pair,
              site_id   = site_idx,
              site_type = st,
              period    = period,
              detected  = as.integer(yA | yB)
            )
          }
        }
      }
    }

    sim_data <- dplyr::bind_rows(rows) |>
      dplyr::mutate(
        period    = factor(period,    levels = c("Before", "After")),
        site_type = factor(site_type, levels = c("Control", "Stone_pile")),
        site_id   = factor(site_id),
        pair_id   = factor(pair_id)
      )

    # Fast logistic GLMM surrogate — skip if data is perfectly separated
    m <- tryCatch(
      lme4::glmer(detected ~ period * site_type + (1 | pair_id / site_id),
                  data = sim_data, family = binomial,
                  control = lme4::glmerControl(optimizer = "bobyqa")),
      error   = function(e) NULL,
      warning = function(w) NULL
    )
    if (is.null(m)) return(NA)

    coef_name <- grep("periodAfter:site_typeStone_pile",
                      names(lme4::fixef(m)), value = TRUE)
    if (length(coef_name) == 0) return(NA)
    est <- lme4::fixef(m)[coef_name]
    se  <- sqrt(diag(vcov(m)))[coef_name]
    # Wald 95% CI lower bound > 0 → BACI effect detected
    (est - 1.96 * se) > 0
  }

  run_power <- function(n_pairs, n_after_rounds, n_sims = 300) {
    results <- vapply(seq_len(n_sims), function(i)
      one_sim(n_pairs, n_after_rounds,
              logit_psi_ctrl_before, b_period_after, b_sp, b_sp_after,
              logit_pA, logit_pB, sigma_site),
      logical(1))
    mean(results, na.rm = TRUE)
  }

  set.seed(2026)
  power_grid <- tidyr::crossing(
    n_pairs        = c(3, 5, 7, 10, 15, 20, 30),
    n_after_rounds = c(2, 4, 6, 8, 10, 12)
  ) |>
    dplyr::mutate(
      power = purrr::map2_dbl(n_pairs, n_after_rounds, run_power,
                               .progress = TRUE)
    )

  saveRDS(power_grid, power_cache)
  cat("Power analysis complete and cached.\n")

} else {
  power_grid <- readRDS(power_cache)
  cat("Loaded cached power analysis.\n")
}

print(power_grid |> tidyr::pivot_wider(names_from = n_after_rounds,
                                        values_from = power,
                                        names_prefix = "rounds_"))
```

```{r}
#| label: power-plots
# ---- Heatmap ----------------------------------------------------------------
p_heat <- ggplot(power_grid,
                 aes(x = factor(n_after_rounds), y = factor(n_pairs),
                     fill = power)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = scales::percent(power, accuracy = 1)),
            size = 3.2, colour = ifelse(power_grid$power > 0.6, "white", "grey20")) +
  geom_contour(aes(x = as.numeric(factor(n_after_rounds)),
                   y = as.numeric(factor(n_pairs)),
                   z = power),
               breaks = 0.8, colour = "black", linewidth = 1, linetype = "dashed",
               inherit.aes = FALSE) +
  scale_fill_gradientn(
    colours = c("#d73027", "#fc8d59", "#fee090", "#e0f3f8", "#74add1", "#313695"),
    values  = scales::rescale(c(0, 0.2, 0.4, 0.6, 0.8, 1)),
    limits  = c(0, 1),
    labels  = scales::percent_format(accuracy = 1),
    name    = "Power"
  ) +
  scale_x_discrete(name = "Number of After-period monitoring rounds") +
  scale_y_discrete(name = "Number of site pairs\n(Stone pile + Control)") +
  theme_bw(base_size = 12) +
  theme(
    panel.grid    = element_blank(),
    legend.position = "right"
  )

ggsave(file.path("outputs", "occupancy", "power_heatmap.png"),
       p_heat, width = 8, height = 6, dpi = 300)

# ---- Lines by n_after_rounds ------------------------------------------------
p_lines <- power_grid |>
  dplyr::mutate(n_after_rounds = factor(n_after_rounds,
                  labels = paste0(c(2,4,6,8,10,12), " After rounds"))) |>
  ggplot(aes(x = n_pairs, y = power, colour = n_after_rounds,
             group = n_after_rounds)) +
  geom_hline(yintercept = 0.8, linetype = "dashed", colour = "grey40") +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  annotate("text", x = max(power_grid$n_pairs) * 0.98, y = 0.82,
           label = "80 % power", hjust = 1, colour = "grey40", size = 3.5) +
  scale_colour_manual(
    values = c("#313695","#4575b4","#74add1","#f46d43","#d73027","#a50026"),
    name   = NULL
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_x_continuous(breaks = c(3, 5, 7, 10, 15, 20, 30)) +
  labs(x = "Number of site pairs (Stone pile + Control)",
       y = "Statistical power") +
  theme_bw(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position  = "right"
  )

ggsave(file.path("outputs", "occupancy", "power_curves.png"),
       p_lines, width = 8, height = 5, dpi = 300)

p_heat
p_lines
```

```{r}
#| label: power-summary
# What does it take to hit 80% power at the current effect size?
power_grid |>
  dplyr::filter(power >= 0.8) |>
  dplyr::group_by(n_after_rounds) |>
  dplyr::slice_min(n_pairs, n = 1) |>
  dplyr::ungroup() |>
  dplyr::transmute(
    `After rounds` = n_after_rounds,
    `Site pairs needed` = n_pairs,
    `Power achieved` = scales::percent(power, accuracy = 1)
  ) |>
  knitr::kable(caption = paste0(
    "Minimum site pairs required to reach 80 % power at the observed effect size ",
    "(BACI ≈ ", round(b_sp_after, 2), " log-odds), by number of After-period rounds."))
```

---

# CPUE-based Power Analysis (M2aak — Kōura abundance BACI)

The occupancy analysis sits near a ceiling (ψ ≈ 0.81–0.95) so the detectable
BACI signal is compressed. CPUE captures the abundance dimension that occupancy
misses. M2aak (Kōura CPUE, Stone_pile vs Control, full BACI) shows a positive
but non-significant BACI of 0.311 log-units (×1.36 on the ratio scale, p = 0.800).
This power analysis asks: how many site pairs and monitoring rounds would be
needed to detect an effect of that size at 80% power?

**Model structure simulated:**
`log(CPUE) ~ period * site_type + monitoring_int + (1|pair_id) + (1|site_id)`

Fixed effects come directly from M2aak. Random effect SDs and the log-scale
residual SD are set to ecologically plausible defaults — update them from
`VarCorr(M2aak)` and `sigma(M2aak)` if you want exact values.

```{r}
#| label: cpue-params
library(lme4)

# Fixed effects from M2aak (tbl-models-summary-full.csv)
cpue_b0      <- -2.753   # Intercept: log CPUE, Control, Before, monitoring_int = 0
cpue_b_after <-  2.374   # PeriodAfter step-change (large: kōura increase after construction)
cpue_b_sp    <-  0.751   # Stone_pile vs Control baseline difference
cpue_b_time  <- -0.253   # monitoring_int slope (per round; both sites declining over time)
cpue_b_baci  <-  0.311   # BACI: extra Stone_pile gain in After period — the detection target

# Random effects — update from VarCorr(M2aak) once available
sigma_pair_cpue  <- 1.2999  # pair-level intercept SD (from VarCorr(M2aak))
sigma_site_cpue  <- 0.8087  # site-within-pair SD (from VarCorr(M2aak))
sigma_resid_cpue <- 0.9611  # Tweedie dispersion phi (from sigma(M2aak))

cat("M2aak parameters:\n")
cat(sprintf("  BACI estimate:   %.3f (SE = 1.232, p = 0.800)\n", cpue_b_baci))
cat(sprintf("  95%% CI on BACI: [%.2f, %.2f]\n", -2.104, 2.727))
cat(sprintf("  Ratio scale:     exp(%.3f) = %.2f×\n", cpue_b_baci, exp(cpue_b_baci)))
cat(sprintf("  PeriodAfter:     %.3f (strong time effect; both sites gain)\n", cpue_b_after))
cat(sprintf("  monitoring_int:  %.3f (declining trend over rounds)\n", cpue_b_time))
```

```{r}
#| label: cpue-power-sim

one_sim_cpue <- function(n_pairs, n_after_rounds) {

  dat <- tidyr::crossing(
    pair_idx  = seq_len(n_pairs),
    site_type = c("Control", "Stone_pile"),
    round_no  = 0L:as.integer(n_after_rounds)
  ) |>
    dplyr::mutate(
      pair_id        = factor(paste0("p", pair_idx)),
      site_id        = factor(paste0("p", pair_idx, "_", site_type)),
      period         = factor(ifelse(round_no == 0L, "Before", "After"),
                              levels = c("Before", "After")),
      monitoring_int = as.numeric(round_no),
      is_sp          = as.integer(site_type == "Stone_pile"),
      is_after       = as.integer(period == "After")
    )

  # Sample random effects for this iteration
  re_pair <- stats::setNames(
    rnorm(n_pairs, 0, sigma_pair_cpue),
    paste0("p", seq_len(n_pairs))
  )
  re_site <- stats::setNames(
    rnorm(n_pairs * 2L, 0, sigma_site_cpue),
    paste0("p", rep(seq_len(n_pairs), each = 2L), "_",
           c("Control", "Stone_pile"))
  )

  dat <- dat |>
    dplyr::mutate(
      eta  = cpue_b0 +
             cpue_b_after * is_after +
             cpue_b_sp    * is_sp +
             cpue_b_time  * monitoring_int +
             cpue_b_baci  * (is_after * is_sp) +
             re_pair[as.character(pair_id)] +
             re_site[as.character(site_id)],
      cpue = exp(stats::rnorm(dplyr::n(), eta, sigma_resid_cpue))
    )

  fit <- tryCatch(
    suppressWarnings(
      lme4::lmer(log(cpue) ~ period * site_type + monitoring_int +
                   (1 | pair_id) + (1 | site_id),
                 data = dat, REML = FALSE)
    ),
    error = function(e) NULL
  )

  if (is.null(fit)) return(NA_real_)

  cf  <- lme4::fixef(fit)
  vcv <- as.matrix(vcov(fit))
  nm  <- "periodAfter:site_typeStone_pile"
  if (!nm %in% names(cf)) return(NA_real_)

  est <- cf[nm]
  se  <- sqrt(vcv[nm, nm])
  if (is.na(se) || se < 1e-9) return(NA_real_)

  # Two-sided 95% CI must exclude zero in positive direction
  as.numeric((est - 1.96 * se) > 0)
}

run_power_cpue <- function(n_pairs, n_after_rounds, n_sims = 200L) {
  results <- vapply(
    seq_len(n_sims),
    function(s) {
      set.seed(s * 100L + n_pairs * 10L + n_after_rounds)
      one_sim_cpue(n_pairs, n_after_rounds)
    },
    numeric(1L)
  )
  mean(results, na.rm = TRUE)
}

cpue_power_cache <- file.path("outputs", "occupancy", "cpue_power_analysis.rds")

if (!file.exists(cpue_power_cache)) {
  cat("Running CPUE power analysis (200 sims × 42 scenarios)...\n")

  cpue_power_grid <- tidyr::crossing(
    n_pairs        = c(3L, 5L, 7L, 10L, 15L, 20L, 30L),
    n_after_rounds = c(2L, 4L, 6L, 8L, 10L, 12L)
  ) |>
    dplyr::mutate(
      power = purrr::map2_dbl(n_pairs, n_after_rounds, run_power_cpue,
                              .progress = TRUE)
    )

  saveRDS(cpue_power_grid, cpue_power_cache)
  cat("CPUE power analysis complete and cached.\n")

} else {
  cpue_power_grid <- readRDS(cpue_power_cache)
  cat("Loaded cached CPUE power analysis.\n")
}

print(cpue_power_grid |>
        tidyr::pivot_wider(names_from = n_after_rounds, values_from = power,
                           names_prefix = "rounds_"))
```

```{r}
#| label: cpue-power-plots

p_heat_cpue <- ggplot(cpue_power_grid,
                      aes(x = factor(n_after_rounds), y = factor(n_pairs),
                          fill = power)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = scales::percent(power, accuracy = 1)),
            size = 3.2, colour = "grey20") +
  scale_fill_gradientn(
    colours = c("#d73027", "#fc8d59", "#fee090", "#e0f3f8", "#74add1", "#313695"),
    values  = scales::rescale(c(0, 0.2, 0.4, 0.6, 0.8, 1)),
    limits  = c(0, 1),
    labels  = scales::percent_format(accuracy = 1),
    name    = "Power"
  ) +
  scale_x_discrete(name = "Number of After-period monitoring rounds") +
  scale_y_discrete(name = "Number of site pairs\n(Stone pile + Control)") +
  theme_bw(base_size = 12) +
  theme(panel.grid = element_blank(), legend.position = "right")

ggsave(file.path("outputs", "occupancy", "cpue_power_heatmap.png"),
       p_heat_cpue, width = 8, height = 6, dpi = 300)

p_lines_cpue <- cpue_power_grid |>
  dplyr::mutate(n_after_rounds = factor(n_after_rounds,
                  labels = paste0(c(2, 4, 6, 8, 10, 12), " After rounds"))) |>
  ggplot(aes(x = n_pairs, y = power, colour = n_after_rounds,
             group = n_after_rounds)) +
  geom_hline(yintercept = 0.8, linetype = "dashed", colour = "grey40") +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  annotate("text", x = max(cpue_power_grid$n_pairs) * 0.98, y = 0.82,
           label = "80 % power", hjust = 1, colour = "grey40", size = 3.5) +
  scale_colour_manual(
    values = c("#313695", "#4575b4", "#74add1", "#f46d43", "#d73027", "#a50026"),
    name   = NULL
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_x_continuous(breaks = c(3, 5, 7, 10, 15, 20, 30)) +
  labs(x = "Number of site pairs (Stone pile + Control)",
       y = "Statistical power") +
  theme_bw(base_size = 12) +
  theme(panel.grid.minor = element_blank(), legend.position = "right")

ggsave(file.path("outputs", "occupancy", "cpue_power_curves.png"),
       p_lines_cpue, width = 8, height = 5, dpi = 300)

p_heat_cpue
p_lines_cpue
```

```{r}
#| label: cpue-power-summary
at_80 <- cpue_power_grid |>
  dplyr::filter(power >= 0.8) |>
  dplyr::group_by(n_after_rounds) |>
  dplyr::slice_min(n_pairs, n = 1) |>
  dplyr::ungroup() |>
  dplyr::transmute(
    `After rounds`      = n_after_rounds,
    `Site pairs needed` = n_pairs,
    `Power achieved`    = scales::percent(power, accuracy = 1)
  )

cap <- paste0(
  "CPUE BACI power (M2aak: exp(BACI) = ", round(exp(cpue_b_baci), 2), "×; ",
  "σ_pair = ", sigma_pair_cpue, ", σ_site = ", sigma_site_cpue,
  ", σ_resid = ", sigma_resid_cpue, "). ",
  "Log-normal surrogate (lmer)."
)

if (nrow(at_80) == 0) {
  cat("No scenario in the grid reaches 80% power.\n")
  cat(sprintf("Maximum power achieved: %.0f%% (n_pairs = 30, rounds = 8).\n",
              100 * max(cpue_power_grid$power)))
  cat("The between-pair variance (σ_pair = 1.30) dominates, matching the\n")
  cat("occupancy result: more rounds do not compensate for too few pairs.\n")
  cpue_power_grid |>
    dplyr::filter(power == max(power)) |>
    knitr::kable(caption = cap)
} else {
  knitr::kable(at_80, caption = cap)
}
```
